---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)

run_readme_examples <- FALSE
if (requireNamespace("SCProkaR", quietly = TRUE)) {
  run_readme_examples <- TRUE
} else if (file.exists("DESCRIPTION") && requireNamespace("devtools", quietly = TRUE)) {
  devtools::load_all(".", quiet = TRUE)
  run_readme_examples <- TRUE
}
```

# SCProkaR

<!-- badges: start -->
[![pkgdown](https://img.shields.io/badge/docs-pkgdown-blue.svg)](https://monash-li-lab.github.io/scProka/)
<!-- badges: end -->

SCProkaR is a bacterial and microbial single-cell RNA-seq toolkit built around
`SingleCellExperiment`. It standardizes raw inputs, calculates bacterial QC
metrics such as rRNA fraction, integrates batches with lightweight backends,
benchmarks corrected embeddings including user-supplied embeddings, and
supports integrated clustering for downstream analysis. The package also
includes pseudobulk differential expression utilities and Time Aware
Trajectory Analysis (TATA) for time-informed trajectory inference.

## Installation

You can install the development version of SCProkaR like so:

``` r
remotes::install_github("Monash-Li-Lab/scProka")
```

## Package website

- Core workflow vignette:
  <https://monash-li-lab.github.io/scProka/articles/scprokar-workflow.html>
- Comprehensive PDF tutorial source:
  `vignettes/scprokar-pdf-tutorial.Rmd`
- TATA vignette:
  <https://monash-li-lab.github.io/scProka/articles/tata-workflow.html>
- Function reference:
  <https://monash-li-lab.github.io/scProka/reference/index.html>

## Example data and workflow

The package ships with `sce1`, a small three-treatment
`SingleCellExperiment` example object used throughout the main workflow
vignette.

```{r example-data, eval = run_readme_examples}
library(SCProkaR)
data("sce1", package = "SCProkaR")

sce1
table(sce1$treatments)
table(sce1$timepoints)
```

A minimal workflow is:

```{r example, eval = run_readme_examples}
sce <- sce1
sce$batch <- factor(sce$treatments)
sce <- RunBacQC(sce)

if (requireNamespace("batchelor", quietly = TRUE)) {
  sce <- IntegrateBacData(sce, batch_col = "batch", method = "mnn", dims = 1:10)
  integrated_reduction <- "integrated_mnn"
} else {
  fallback_latent <- matrix(
    rnorm(ncol(sce) * 10),
    ncol = 10,
    dimnames = list(colnames(sce), paste0("dim", seq_len(10)))
  )
  sce <- RegisterIntegrationEmbedding(sce, embedding = fallback_latent, method_name = "mock")
  integrated_reduction <- "integrated_mock"
}

sce <- RunIntegratedUMAP(sce, reduction = integrated_reduction)
sce <- RunIntegratedClustering(sce, reduction = integrated_reduction, cluster_col = "integrated_clusters")

tata <- run_tata(
  sce = sce,
  dimred = integrated_reduction,
  time_col = "timepoints",
  cluster_col = "tata_cluster",
  k = 20,
  do_pseudotime = TRUE,
  seed = 101
)

names(tata)
```

If your input is a Seurat `.rds` object, you can pass it directly to
`CreateBacObject()` and SCProkaR will extract the assay counts, metadata, and
stored reductions:

```{r seurat-example, eval = FALSE}
seu <- readRDS("sample1.rds")
sce <- CreateBacObject(
  seu,
  seurat_assay = "RNA",
  seurat_layer = "counts",
  sample_col = "orig.ident"
)
```

If your input is a 10x Genomics Cell Ranger output directory, you can also pass
the directory path directly to `CreateBacObject()`. With
`run_unintegrated = TRUE`, SCProkaR will run a standard unintegrated workflow
analogous to normalization, HVG selection, PCA, neighbour graph construction,
clustering, and UMAP in Seurat, but without any Seurat dependency:

```{r tenx-example, eval = FALSE}
sce <- CreateBacObject(
  "sample1/outs",
  run_unintegrated = TRUE,
  unintegrated_dims = 1:30,
  unintegrated_cluster_col = "unintegrated_clusters",
  unintegrated_umap_name = "umap.unintegrated",
  unintegrated_resolution = 2
)
```

For multiple samples, create and QC each sample separately, then merge with
`MergeBacObjects()` before integration:

```{r multi-sample-example, eval = FALSE}
seu1 <- readRDS("sample1.rds")
seu2 <- readRDS("sample2.rds")

seu1$sample_id <- "sample1"
seu1$batch <- "sample1"
seu2$sample_id <- "sample2"
seu2$batch <- "sample2"

sce1 <- CreateBacObject(seu1, seurat_assay = "RNA", sample_col = "sample_id", batch_col = "batch")
sce2 <- CreateBacObject(seu2, seurat_assay = "RNA", sample_col = "sample_id", batch_col = "batch")

sce1 <- RunBacQC(sce1)
sce2 <- RunBacQC(sce2)

sce1 <- FilterBacCells(sce1, min_counts = 200, min_features = 50, max_rrna_fraction = 0.2)
sce2 <- FilterBacCells(sce2, min_counts = 200, min_features = 50, max_rrna_fraction = 0.2)

sce <- MergeBacObjects(sce1, sce2, gene_mode = "intersect")
sce <- IntegrateBacData(sce, batch_col = "batch", method = "mnn")
sce <- RunIntegratedUMAP(sce, reduction = "integrated_mnn")
sce <- RunIntegratedClustering(sce, reduction = "integrated_mnn", cluster_col = "integrated_clusters")

PlotReduction(sce, reduction = "umap", colour_by = "batch")
PlotReduction(sce, reduction = "umap_integrated_mnn", colour_by = "batch")
PlotReduction(sce, reduction = "umap_integrated_mnn", colour_by = "integrated_clusters")
```

If the original cell barcodes overlap across samples, `MergeBacObjects()` will
automatically make them unique by prefixing them with the sample or batch label
when available.

## Optional Backends

Some workflows require optional packages:

- `batchelor` for MNN integration
- `harmony` for Harmony integration
- `SeuratObject` only when you pass a Seurat `.rds` directly into `CreateBacObject()`
- `igraph` for integrated graph clustering
- `edgeR` for pseudobulk differential expression
- `pkgdown` for building the package website

The public API is intentionally centered on one object type so downstream
methods can share assays, metadata, reduced dimensions, and analysis results.
Custom embeddings stored in `reducedDims(sce)` can also be benchmarked directly.
