---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)

run_readme_examples <- FALSE
if (requireNamespace("SCProkaR", quietly = TRUE)) {
  run_readme_examples <- TRUE
} else if (file.exists("DESCRIPTION") && requireNamespace("devtools", quietly = TRUE)) {
  devtools::load_all(".", quiet = TRUE)
  run_readme_examples <- TRUE
}
```

# SCProkaR

<!-- badges: start -->
<!-- badges: end -->

SCProkaR is a bacterial and microbial single-cell RNA-seq toolkit built around
`SingleCellExperiment`. It standardizes raw inputs, calculates bacterial QC
metrics such as rRNA fraction, integrates batches with lightweight backends,
benchmarks corrected embeddings including user-supplied embeddings, and
supports integrated clustering for downstream analysis.

## Installation

You can install the development version of SCProkaR like so:

``` r
remotes::install_github("jinxinz/SCProkaR")
```

## Example

The core workflow is:

```{r example, eval = run_readme_examples}
library(SCProkaR)

set.seed(1)
counts <- matrix(
  rpois(6 * 12, lambda = 5),
  nrow = 6,
  dimnames = list(
    c("rrsA", "rrlB", "rplC", "geneD", "geneE", "geneF"),
    paste0("cell", seq_len(12))
  )
)

cell_metadata <- data.frame(
  batch = rep(c("batch1", "batch2"), each = 6),
  cell_type = rep(c("A", "B", "C"), length.out = 12),
  row.names = colnames(counts),
  stringsAsFactors = FALSE
)

sce <- CreateBacObject(counts, cell_metadata = cell_metadata, batch_col = "batch")
sce <- RunBacQC(sce)

if (requireNamespace("batchelor", quietly = TRUE)) {
  sce <- IntegrateBacData(sce, batch_col = "batch", method = "mnn", dims = 1:2)
  integrated_reduction <- "integrated_mnn"
} else {
  fallback_latent <- matrix(
    rnorm(ncol(sce) * 2),
    ncol = 2,
    dimnames = list(colnames(sce), c("dim1", "dim2"))
  )
  sce <- RegisterIntegrationEmbedding(sce, embedding = fallback_latent, method_name = "mock")
  integrated_reduction <- "integrated_mock"
}

custom_latent <- matrix(
  rnorm(ncol(sce) * 2),
  ncol = 2,
  dimnames = list(colnames(sce), c("dim1", "dim2"))
)
sce <- RegisterIntegrationEmbedding(sce, embedding = custom_latent, method_name = "scanorama")
bench <- BenchmarkIntegration(sce, batch_col = "batch", label_col = "cell_type")

if (requireNamespace("igraph", quietly = TRUE)) {
  sce <- RunIntegratedClustering(sce, reduction = integrated_reduction)
}
```

If your input is a Seurat `.rds` object, you can pass it directly to
`CreateBacObject()` and SCProkaR will extract the assay counts, metadata, and
stored reductions:

```{r seurat-example, eval = FALSE}
seu <- readRDS("sample1.rds")
sce <- CreateBacObject(
  seu,
  seurat_assay = "RNA",
  seurat_layer = "counts",
  sample_col = "orig.ident"
)
```

For multiple samples, create and QC each sample separately, then merge with
`MergeBacObjects()` before integration:

```{r multi-sample-example, eval = FALSE}
seu1 <- readRDS("sample1.rds")
seu2 <- readRDS("sample2.rds")

seu1$sample_id <- "sample1"
seu1$batch <- "sample1"
seu2$sample_id <- "sample2"
seu2$batch <- "sample2"

sce1 <- CreateBacObject(seu1, seurat_assay = "RNA", sample_col = "sample_id", batch_col = "batch")
sce2 <- CreateBacObject(seu2, seurat_assay = "RNA", sample_col = "sample_id", batch_col = "batch")

sce1 <- RunBacQC(sce1)
sce2 <- RunBacQC(sce2)

sce1 <- FilterBacCells(sce1, min_counts = 200, min_features = 50, max_rrna_fraction = 0.2)
sce2 <- FilterBacCells(sce2, min_counts = 200, min_features = 50, max_rrna_fraction = 0.2)

sce <- MergeBacObjects(sce1, sce2, gene_mode = "intersect")
sce <- IntegrateBacData(sce, batch_col = "batch", method = "mnn")
sce <- RunIntegratedUMAP(sce, reduction = "integrated_mnn")
sce <- RunIntegratedClustering(sce, reduction = "integrated_mnn", cluster_col = "integrated_clusters")

PlotReduction(sce, reduction = "umap", colour_by = "batch")
PlotReduction(sce, reduction = "umap_integrated_mnn", colour_by = "batch")
PlotReduction(sce, reduction = "umap_integrated_mnn", colour_by = "integrated_clusters")
```

If the original cell barcodes overlap across samples, `MergeBacObjects()` will
automatically make them unique by prefixing them with the sample or batch label
when available.

## Optional Backends

Some workflows require optional packages:

- `batchelor` for MNN integration
- `harmony` for Harmony integration
- `SeuratObject` only when you pass a Seurat `.rds` directly into `CreateBacObject()`
- `igraph` for integrated graph clustering

The public API is intentionally centered on one object type so downstream
methods can share assays, metadata, reduced dimensions, and analysis results.
Custom embeddings stored in `reducedDims(sce)` can also be benchmarked directly.
