---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# SCProkaR

<!-- badges: start -->
<!-- badges: end -->

SCProkaR is a bacterial and microbial single-cell RNA-seq toolkit built around
`SingleCellExperiment`. It standardizes raw inputs, calculates bacterial QC
metrics such as rRNA fraction, integrates batches with lightweight backends,
benchmarks corrected embeddings including user-supplied embeddings,
reconstructs PAGA-inspired trajectories, and supports pseudobulk and
time-series differential expression.

## Installation

You can install the development version of SCProkaR like so:

``` r
remotes::install_github("jinxinz/SCProkaR")
```

## Example

The core workflow is:

```{r example}
library(SCProkaR)

sce <- CreateBacObject(counts)
sce <- RunBacQC(sce)
sce <- IntegrateBacData(sce, batch_col = "batch", method = "mnn")
sce <- RegisterIntegrationEmbedding(sce, embedding = custom_latent, method_name = "scanorama")
bench <- BenchmarkIntegration(sce, batch_col = "batch", label_col = "cell_type")
sce <- RunBacPAGA(sce, cluster_col = "cluster", reduction = "integrated_mnn")
pb <- AggregatePseudobulk(sce, sample_col = "sample_id", group_cols = c("condition"))
```

If your input is a Seurat `.rds` object, you can pass it directly to
`CreateBacObject()` and SCProkaR will extract the assay counts, metadata, and
stored reductions:

```{r seurat-example, eval = FALSE}
seu <- readRDS("sample1.rds")
sce <- CreateBacObject(
  seu,
  seurat_assay = "RNA",
  seurat_layer = "counts",
  sample_col = "orig.ident"
)
```

For multiple samples, create and QC each sample separately, then merge with
`MergeBacObjects()` before integration:

```{r multi-sample-example, eval = FALSE}
seu1 <- readRDS("sample1.rds")
seu2 <- readRDS("sample2.rds")

seu1$sample_id <- "sample1"
seu1$batch <- "sample1"
seu2$sample_id <- "sample2"
seu2$batch <- "sample2"

sce1 <- CreateBacObject(seu1, seurat_assay = "RNA", sample_col = "sample_id", batch_col = "batch")
sce2 <- CreateBacObject(seu2, seurat_assay = "RNA", sample_col = "sample_id", batch_col = "batch")

sce1 <- RunBacQC(sce1)
sce2 <- RunBacQC(sce2)

sce1 <- FilterBacCells(sce1, min_counts = 200, min_features = 50, max_rrna_fraction = 0.2)
sce2 <- FilterBacCells(sce2, min_counts = 200, min_features = 50, max_rrna_fraction = 0.2)

sce <- MergeBacObjects(sce1, sce2, gene_mode = "intersect")
sce <- IntegrateBacData(sce, batch_col = "batch", method = "mnn")
sce <- RunIntegratedUMAP(sce, reduction = "integrated_mnn")

PlotReduction(sce, reduction = "umap", colour_by = "batch")
PlotReduction(sce, reduction = "umap_integrated_mnn", colour_by = "batch")
PlotReduction(sce, reduction = "umap_integrated_mnn", colour_by = "seurat_clusters")
```

If the original cell barcodes overlap across samples, `MergeBacObjects()` will
automatically make them unique by prefixing them with the sample or batch label
when available.

## Optional Backends

Some workflows require optional packages:

- `batchelor` for MNN integration
- `harmony` for Harmony integration
- `SeuratObject` only when you pass a Seurat `.rds` directly into `CreateBacObject()`
- `slingshot` for lineage-aware pseudotime
- `edgeR` for pseudobulk DE
- `tradeSeq` for time-series DE

The public API is intentionally centered on one object type so downstream
methods can share assays, metadata, reduced dimensions, and analysis results.
Custom embeddings stored in `reducedDims(sce)` can also be benchmarked directly.
