---
title: "Pseudobulk Differential Expression Tutorial"
format:
  html:
    toc: true
    code-fold: true
execute:
  echo: true
  warning: false
  message: false
---

# Overview

This vignette shows how to use the pseudobulk and edgeR differential expression
functions in `scProkaR` on simulated single-cell data.

The tutorial covers four steps:

1. simulate a branching multi-drug dataset
2. create pseudo-replicate sample labels for pseudobulk aggregation
3. run pairwise edgeR differential expression
4. run spline-based time-series differential expression and plot fitted curves

Because the simulator returns single cells rather than biological replicates, we
create replicate-like pseudobulk groups within each `condition x timepoint`
stratum. This is useful for demonstrating the workflow, but for real analyses
you would usually pseudobulk by true sample, donor, or experimental unit.

# Load the package

```{r}
library(devtools)
devtools::load_all("..", quiet = TRUE, reset = TRUE)

library(ggplot2)
library(SingleCellExperiment)
```

# Simulate a multi-drug branching dataset

The multidrug simulator is a good fit for DE tutorials because it already
contains:

- real experimental `timepoint`
- `condition` labels
- gradual progression rather than isolated clusters
- branch-specific expression programs

```{r}
set.seed(101)

sce <- simulate_tata_multidrug_sce(
  n_cells = 5000,
  n_features = 100,
  n_pcs = 20,
  seed = 101
)

table(sce$condition, sce$timepoint)
```

```{r}
umap_df <- as.data.frame(SingleCellExperiment::reducedDim(sce, "UMAP"))
colnames(umap_df)[1:2] <- c("UMAP_1", "UMAP_2")
umap_df$condition <- sce$condition
umap_df$timepoint <- factor(sce$timepoint, levels = sort(unique(sce$timepoint)))

ggplot(umap_df, aes(UMAP_1, UMAP_2, colour = condition)) +
  geom_point(size = 0.45, alpha = 0.75) +
  theme_classic() +
  labs(title = "Simulated multidrug dataset", colour = "condition")
```

```{r}
ggplot(umap_df, aes(UMAP_1, UMAP_2, colour = timepoint)) +
  geom_point(size = 0.45, alpha = 0.75) +
  scale_colour_viridis_d(option = "plasma") +
  theme_classic() +
  labs(title = "Simulated dataset coloured by timepoint", colour = "timepoint")
```

# Create pseudo-replicate sample IDs

The DE workflows operate on pseudobulk samples. For this tutorial, we create
four pseudo-replicates inside each `condition x timepoint` group by randomly
partitioning cells within each stratum.

```{r}
assign_pseudoreplicates <- function(
    sce,
    group_cols = c("condition", "timepoint"),
    n_reps = 4,
    seed = 101) {
  set.seed(seed)

  meta <- as.data.frame(SummarizedExperiment::colData(sce))
  strata <- interaction(meta[, group_cols, drop = FALSE], drop = TRUE, lex.order = TRUE)
  rep_id <- integer(ncol(sce))

  for (level_name in levels(strata)) {
    idx <- which(strata == level_name)
    rep_id[idx] <- sample(rep(seq_len(n_reps), length.out = length(idx)))
  }

  sce$pb_replicate <- factor(rep_id, levels = seq_len(n_reps))
  sce$pb_sample_id <- paste(
    "pb",
    as.character(sce$condition),
    paste0("t", sce$timepoint),
    paste0("r", sce$pb_replicate),
    sep = "_"
  )

  sce
}

sce <- assign_pseudoreplicates(sce, n_reps = 4, seed = 101)

table(sce$condition, sce$timepoint, sce$pb_replicate)[, , 1]
```

# Aggregate pseudobulk samples

We now sum counts within each pseudobulk sample. The output pseudobulk object
retains the sample-level metadata, including `condition`, `timepoint`, and the
number of cells contributing to each pseudobulk.

```{r}
pb <- aggregate_pseudobulk(
  sce,
  sample_cols = c("pb_sample_id", "condition", "timepoint", "pb_replicate"),
  aggregation = "sum",
  min_cells = 20
)

pb <- normalize_pseudobulk(pb, method = "TMM")

dim(pb)
head(as.data.frame(SummarizedExperiment::colData(pb)))
```

```{r}
pb_pca <- prcomp(t(as.matrix(SummarizedExperiment::assay(pb, "logcounts"))), scale. = TRUE)
pb_df <- as.data.frame(pb_pca$x[, 1:2, drop = FALSE])
colnames(pb_df)[1:2] <- c("PC1", "PC2")
pb_df$condition <- pb$condition
pb_df$timepoint <- factor(pb$timepoint, levels = sort(unique(pb$timepoint)))

ggplot(pb_df, aes(PC1, PC2, colour = condition, shape = timepoint)) +
  geom_point(size = 3.2) +
  theme_classic() +
  labs(title = "Pseudobulk PCA", colour = "condition", shape = "timepoint")
```

# Pairwise DE: timepoint comparisons with treatment as a covariate

The first analysis treats `timepoint` as the primary group while adding
`condition` as an additive covariate. By default, we compare each timepoint
against the reference timepoint `0`.

```{r}
pairwise_time <- run_edger_pairwise_de(
  pb,
  group_col = "timepoint",
  covariates = "condition",
  contrast_type = "reference",
  reference_level = "0",
  normalization = "TMM"
)

names(pairwise_time$tables)
```

```{r}
pairwise_sig_counts <- aggregate(
  significant ~ contrast,
  data = pairwise_time$combined_table,
  FUN = sum
)

ggplot(pairwise_sig_counts, aes(x = contrast, y = significant, fill = contrast)) +
  geom_col(width = 0.72) +
  theme_classic() +
  theme(legend.position = "none") +
  labs(
    title = "Significant genes in timepoint-vs-baseline contrasts",
    x = "contrast",
    y = "number of DE genes"
  )
```

```{r}
head(
  pairwise_time$tables[["24_vs_0"]][order(pairwise_time$tables[["24_vs_0"]]$FDR), c("gene", "logFC", "FDR")],
  10
)
```

```{r}
plot_pairwise_de_volcano(
  pairwise_time,
  contrast = "24_vs_0",
  fdr_cutoff = 0.05,
  lfc_cutoff = 0.5,
  top_n_labels = 8
)
```

```{r}
plot_pairwise_de_ma(
  pairwise_time,
  contrast = "24_vs_0",
  fdr_cutoff = 0.05
)
```

# Pairwise DE: treatment-timepoint composite groups

Sometimes we want a more explicit comparison such as `drug_A at 24h` versus
`drug_A at 0h`. This is handled by combining treatment and timepoint into one
composite group and then testing manual contrasts.

```{r}
pairwise_composite <- run_edger_pairwise_de(
  pb,
  group_col = "condition",
  combine_group_cols = c("condition", "timepoint"),
  contrast_type = "manual",
  manual_pairs = data.frame(
    group1 = c("drug_A__24", "drug_C__48", "drug_B__96"),
    group2 = c("drug_A__0", "drug_C__0", "drug_B__0"),
    stringsAsFactors = FALSE
  ),
  normalization = "TMM"
)

names(pairwise_composite$tables)
```

```{r}
pairwise_composite_sig <- aggregate(
  significant ~ contrast,
  data = pairwise_composite$combined_table,
  FUN = sum
)

ggplot(pairwise_composite_sig, aes(x = contrast, y = significant, fill = contrast)) +
  geom_col(width = 0.72) +
  theme_classic() +
  theme(legend.position = "none") +
  labs(
    title = "Significant genes in treatment-timepoint contrasts",
    x = "contrast",
    y = "number of DE genes"
  )
```

```{r}
plot_pairwise_de_volcano(
  pairwise_composite,
  contrast = "drug_A__24_vs_drug_A__0",
  fdr_cutoff = 0.05,
  lfc_cutoff = 0.5,
  top_n_labels = 8
)
```

# Spline-based time-series DE

The second DE workflow models time as a continuous variable using a natural
spline. Here we fit a separate spline model within each condition and ask which
genes vary significantly over time.

```{r}
spline_de <- run_edger_spline_de(
  pb,
  time_col = "timepoint",
  condition_col = "condition",
  df = 3,
  normalization = "TMM",
  return_curve_for = "significant"
)

names(spline_de$by_condition)
```

```{r}
spline_sig_counts <- aggregate(
  significant ~ condition,
  data = spline_de$combined_table,
  FUN = sum
)

ggplot(spline_sig_counts, aes(x = condition, y = significant, fill = condition)) +
  geom_col(width = 0.72) +
  theme_classic() +
  theme(legend.position = "none") +
  labs(
    title = "Genes with significant temporal change within each condition",
    x = "condition",
    y = "number of spline DE genes"
  )
```

```{r}
top_by_condition <- lapply(
  split(spline_de$combined_table, spline_de$combined_table$condition),
  function(df) {
    keep_cols <- intersect(c("gene", "F", "PValue", "FDR"), colnames(df))
    head(df[order(df$FDR), keep_cols, drop = FALSE], 6)
  }
)

top_by_condition
```

# Plot fitted spline curves together

The fitted curves below overlay conditions for the top spline-associated genes.
This makes it easy to see whether genes rise, fall, or diverge differently
across treatments.

```{r}
plot_time_series_deg_curves(spline_de, top_n = 9)
```

# One-step wrapper

If you prefer a single entry point, `run_scproka_de()` performs pseudobulk
aggregation and then dispatches to the requested DE workflow.

```{r}
wrapped_pairwise <- run_scproka_de(
  sce,
  mode = "pairwise",
  sample_cols = c("pb_sample_id", "condition", "timepoint", "pb_replicate"),
  aggregation = "sum",
  min_cells = 20,
  normalization = "TMM",
  group_col = "timepoint",
  covariates = "condition",
  contrast_type = "reference",
  reference_level = "0"
)

names(wrapped_pairwise)
```

```{r}
wrapped_spline <- run_scproka_de(
  sce,
  mode = "time_series",
  sample_cols = c("pb_sample_id", "condition", "timepoint", "pb_replicate"),
  aggregation = "sum",
  min_cells = 20,
  normalization = "TMM",
  time_col = "timepoint",
  condition_col = "condition",
  df = 3
)

names(wrapped_spline)
```

# Summary

This DE tutorial introduced three ways to work with the new module:

- `aggregate_pseudobulk()` and `normalize_pseudobulk()` for manual control
- `run_edger_pairwise_de()` for group-to-group edgeR contrasts
- `run_edger_spline_de()` plus `plot_time_series_deg_curves()` for time-series
  analyses
- `run_scproka_de()` when you want a single wrapper function
