---
title: "SCProkaR Core Workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{SCProkaR Core Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 5
)

if (file.exists("../DESCRIPTION")) {
  if (requireNamespace("devtools", quietly = TRUE)) {
    devtools::load_all("..")
  } else {
    library(SCProkaR)
  }
} else {
  library(SCProkaR)
}

run_workflow <- TRUE
run_de_demo <- TRUE
```

## Overview

This vignette shows three practical `SCProkaR` workflows:

- a three-batch preprocessing and integration workflow starting from Seurat
  `.rds` objects
- a pseudobulk differential-expression workflow on the same example dataset
  after integration
- a time-aware trajectory analysis workflow on the same example dataset

`SCProkaR` keeps the main analysis object as a
`SingleCellExperiment`, while still accepting Seurat input for convenient
conversion.

## Load packages

```{r}
library(SingleCellExperiment)
library(Seurat)
```

## Load Seurat `.rds` objects

The package includes a small example object, `sce1`, which we load directly
from `data/` in the same style as other Bioconductor workflow packages.

```{r}
data("sce1", package = "SCProkaR")
sce <- sce1

seu <- CreateSeuratObject(counts = assay(sce, "counts"), assay = "RNA", meta.data = as.data.frame(colData(sce)))

seu1 <- seu[, seu$treatments == "PMB0.5"]
seu2 <- seu[, seu$treatments == "PMB2"]
seu3 <- seu[, seu$treatments == "control"]
```

For clarity, add sample origin columns if they do not already exist:

```{r eval = run_workflow}

RunDimRed <- function(sce){
  sce <- NormalizeData(sce)
  sce <- FindVariableFeatures(sce, selection.method = "vst", nfeatures = 2000)
  all.genes <- rownames(sce)
  sce <- ScaleData(sce, features = all.genes)
  sce <- RunPCA(sce, features = VariableFeatures(object = sce))
  sce <- FindNeighbors(sce, dims = 1:10)
  sce <- RunUMAP(sce, dims = 1:10)
  return(sce)
}

seu1 <- RunDimRed(seu1)
seu1$batch <- "batch1"

seu2 <- RunDimRed(seu2)
seu2$batch <- "batch2"

seu3 <- RunDimRed(seu3)
seu3$batch <- "batch3"
```

## Create SCProkaR objects

`CreateBacObject()` can extract counts, metadata, and existing reductions such
as `pca` and `umap` directly from each Seurat object.

```{r eval = run_workflow}
sce1 <- CreateBacObject(
  seu1,
  seurat_assay = "RNA",
  seurat_layer = "counts",
  sample_col = "sample",
  batch_col = "batch"
)

sce2 <- CreateBacObject(
  seu2,
  seurat_assay = "RNA",
  seurat_layer = "counts",
  sample_col = "sample",
  batch_col = "batch"
)

sce3 <- CreateBacObject(
  seu3,
  seurat_assay = "RNA",
  seurat_layer = "counts",
  sample_col = "sample",
  batch_col = "batch"
)
```

## Quality control

Compute QC metrics on each sample separately:

```{r eval = run_workflow}
sce1 <- RunBacQC(sce1)
sce2 <- RunBacQC(sce2)
sce3 <- RunBacQC(sce3)
```

The QC results are stored in `colData(sce)`:

```{r eval = run_workflow}
head(as.data.frame(colData(sce1))[, c(
  "total_counts",
  "detected_features",
  "rrna_fraction",
  "pct_rrna"
)])
```

Filter cells using your chosen thresholds:

```{r eval = run_workflow}
sce1 <- FilterBacCells(
  sce1,
  min_counts = 50,
  min_features = 50,
  max_rrna_fraction = 0.20
)

sce2 <- FilterBacCells(
  sce2,
  min_counts = 50,
  min_features = 50,
  max_rrna_fraction = 0.20
)

sce3 <- FilterBacCells(
  sce3,
  min_counts = 50,
  min_features = 50,
  max_rrna_fraction = 0.20
)
```

## Merge samples

Use `MergeBacObjects()` instead of `cbind()` to avoid duplicate-barcode issues
and to align genes safely across samples.

```{r eval = run_workflow}
sce12 <- MergeBacObjects(sce1, sce2, gene_mode = "intersect")
sce <- MergeBacObjects(sce12, sce3, gene_mode = "intersect")
```

If cell barcodes overlap between samples, `MergeBacObjects()` automatically
makes them unique and stores the original barcode in
`colData(sce)$original_cell_id`.

If both input objects contain compatible reductions such as `pca` and `umap`,
`MergeBacObjects()` carries them forward into the merged object.

## Visualize unintegrated structure

If your Seurat objects already contained a UMAP, it is typically available after
conversion:

```{r eval = run_workflow}
PlotReduction(sce, reduction = "umap", colour_by = "batch")
PlotReduction(sce, reduction = "umap", colour_by = "clusters")
```

Interpretation:

- colouring by `batch` shows whether samples are separated before integration
- colouring by `seurat_clusters` or another biological label shows whether
  expected structure is present

`PlotReduction(..., facet_by = "batch")` is the `SCProkaR` equivalent of
Seurat's `split.by` visualization.

## Run integration

### Option 1: MNN

```{r eval = run_workflow}
sce <- IntegrateBacData(
  sce,
  batch_col = "batch",
  method = "mnn",
  dims = 1:30
)
```

### Option 2: Harmony

```{r eval = run_workflow}
sce <- IntegrateBacData(
  sce,
  batch_col = "batch",
  method = "harmony",
  dims = 1:30
)
```

The output embedding is stored as `integrated_mnn` or `integrated_harmony`.

## Visualize the integrated latent space

```{r eval = run_workflow}
PlotReduction(sce, reduction = "integrated_mnn", colour_by = "batch")
PlotReduction(sce, reduction = "integrated_harmony", colour_by = "batch")
```

This gives a quick direct view of the integrated latent space before running a
new UMAP.

## Run UMAP on the integrated embedding

`RunIntegratedUMAP()` computes a UMAP from any stored embedding:

```{r eval = run_workflow}
sce <- RunIntegratedUMAP(sce, reduction = "integrated_mnn")
sce <- RunIntegratedUMAP(sce, reduction = "integrated_harmony")
```

This stores a new reduced dimension named `umap_integrated_mnn`.

```{r eval = run_workflow}
PlotReduction(sce, reduction = "umap_integrated_mnn", colour_by = "batch")
PlotReduction(sce, reduction = "umap_integrated_mnn", colour_by = "clusters")
PlotReduction(sce, reduction = "umap_integrated_harmony", colour_by = "batch")
PlotReduction(sce, reduction = "umap_integrated_harmony", colour_by = "clusters")
```

## Cluster cells after integration

Future downstream analyses should use clusters identified from the integrated
embedding rather than the pre-integration batch structure.

`RunIntegratedClustering()` builds a kNN graph on the corrected embedding and
stores the resulting labels in `colData(sce)`.

```{r eval = run_workflow}
sce <- RunIntegratedClustering(
  sce,
  reduction = "integrated_mnn",
  cluster_col = "integrated_clusters",
  k = 20,
  resolution = 1
)
```

Here, `k` controls the neighbour graph size, while `resolution` controls how
finely that graph is partitioned into clusters. This mirrors the Seurat logic
of `FindNeighbors(k.param = ...)` followed by `FindClusters(resolution = ...)`.

Visualize the integrated clusters on the corrected UMAP:

```{r eval = run_workflow}
PlotReduction(sce, reduction = "umap_integrated_mnn", colour_by = "integrated_clusters")
PlotReduction(
  sce,
  reduction = "umap_integrated_mnn",
  colour_by = "integrated_clusters",
  facet_by = "batch"
)
```

## Seurat-style integration summary panels

The Seurat integration tutorial emphasizes three visual checks after alignment:

- compare unintegrated and integrated embeddings coloured by batch
- compare unintegrated and integrated embeddings coloured by cluster or label
- split the integrated embedding by batch/condition while keeping cluster
  colours fixed

`SCProkaR` provides `PlotIntegrationOverview()` for this pattern.

### MNN example

```{r eval = run_workflow}
mnn_plots <- PlotIntegrationOverview(
  sce,
  unintegrated_reduction = "umap",
  integrated_reduction = "umap_integrated_mnn",
  batch_col = "batch",
  label_col = "integrated_clusters",
  split_by = "batch"
)

mnn_plots$unintegrated_by_batch
mnn_plots$integrated_by_batch
mnn_plots$unintegrated_by_label
mnn_plots$integrated_by_label
mnn_plots$integrated_split
```

### Harmony example

```{r eval = run_workflow}
harmony_plots <- PlotIntegrationOverview(
  sce,
  unintegrated_reduction = "umap",
  integrated_reduction = "umap_integrated_harmony",
  batch_col = "batch",
  label_col = "integrated_clusters",
  split_by = "batch"
)

harmony_plots$unintegrated_by_batch
harmony_plots$integrated_by_batch
harmony_plots$unintegrated_by_label
harmony_plots$integrated_by_label
harmony_plots$integrated_split
```

Interpretation:

- `unintegrated_by_batch` should show stronger sample separation than
  `integrated_by_batch`
- `integrated_by_label` should preserve biologically meaningful group
  separation
- `integrated_split` is the closest equivalent to Seurat's
  `DimPlot(..., split.by = "stim")`

After successful integration:

- batches should mix better than in the original UMAP
- biological clusters should remain interpretable

## Benchmark integration quality

If you have a biological label such as `seurat_clusters`, you can benchmark the
integration with the package's scIB-inspired metric panel:

```{r eval = run_workflow}
bench <- BenchmarkIntegration(
  sce,
  batch_col = "batch",
  label_col = "clusters",
  methods = NULL
)

bench$ranking
bench$plots$overall
bench$plots$key_metrics
bench$plots$batch_removal
bench$plots$bio_conservation
bench$plots$tradeoff
bench$plots$heatmap
```

The key comparison figures are:

- `key_metrics`: per-metric score comparison across methods
- `batch_removal`: average of the batch-removal metrics
- `bio_conservation`: average of the biological-conservation metrics
- `tradeoff`: batch-removal versus bio-conservation, which is often the most
  useful summary of integration quality

## Pseudobulk differential expression

`SCProkaR` also includes pseudobulk differential-expression utilities built
around edgeR. Here we continue with the same example dataset used above rather
than switching to a separate simulated object.

The merged `sce` object currently contains one sample per treatment-timepoint
combination, with:

- `treatments` describing the perturbation group
- `timepoints` describing the experimental sampling time
- `sample` identifying each original sample

For this reason, we create pseudo-replicates inside each sample so the edgeR
workflow has replicate-like pseudobulk columns to operate on. This is useful
for demonstrating the mechanics of the workflow, but in a real study you should
aggregate by the true biological replicate, donor, or independent sample.

### Visualize the dataset before pseudobulk aggregation

```{r eval = run_de_demo}
sce$timepoints_factor <- factor(
  sce$timepoints,
  levels = sort(unique(sce$timepoints))
)

PlotReduction(sce, reduction = "umap_integrated_mnn", colour_by = "treatments")
PlotReduction(sce, reduction = "umap_integrated_mnn", colour_by = "timepoints_factor")
```

### Create pseudo-replicate sample labels

We split cells inside each original sample into pseudo-replicates. This keeps
the DE setup aligned with the real dataset already used in the vignette.

```{r eval = run_de_demo}
assign_pseudoreplicates <- function(
    sce,
    group_cols = c("sample", "treatments", "timepoints"),
    n_reps = 3,
    seed = 101
) {
  set.seed(seed)

  meta <- as.data.frame(SummarizedExperiment::colData(sce))
  strata <- interaction(meta[, group_cols, drop = FALSE], drop = TRUE, lex.order = TRUE)
  rep_id <- integer(ncol(sce))

  for (level_name in levels(strata)) {
    idx <- which(strata == level_name)
    rep_id[idx] <- sample(rep(seq_len(n_reps), length.out = length(idx)))
  }

  sce$pb_replicate <- factor(rep_id, levels = seq_len(n_reps))
  sce$pb_sample_id <- paste(
    as.character(sce$sample),
    paste0("r", sce$pb_replicate),
    sep = "_"
  )

  sce
}

sce_de <- assign_pseudoreplicates(sce, n_reps = 3, seed = 101)

with(
  as.data.frame(SummarizedExperiment::colData(sce_de)),
  table(sample, pb_replicate)
)
```

### Aggregate pseudobulk samples

`aggregate_pseudobulk()` sums counts across cells that share the same sample
definition and keeps the sample-level metadata alongside the aggregated counts.
Here we aggregate the pseudo-replicated object, which preserves all three
treatments while creating replicate-like pseudobulk columns inside each
sample-timepoint stratum.

```{r eval = run_de_demo}
pb <- aggregate_pseudobulk(
  sce_de,
  sample_cols = c("pb_sample_id", "sample", "pb_replicate", "clusters", "treatments", "timepoints"),
  aggregation = "sum",
  min_cells = 20
)

pb <- normalize_pseudobulk(pb, method = "TMM")
pb$timepoints_factor <- factor(pb$timepoints, levels = sort(unique(pb$timepoints)))

pb_pca <- stats::prcomp(
  t(as.matrix(SummarizedExperiment::assay(pb, "logcounts"))),
  scale. = TRUE
)
SingleCellExperiment::reducedDim(pb, "pb_pca") <- pb_pca$x[, 1:10, drop = FALSE]
```

```{r eval = run_de_demo}
PlotReduction(pb, reduction = "pb_pca", colour_by = "treatments", point_size = 5)
PlotReduction(pb, reduction = "pb_pca", colour_by = "timepoints_factor", point_size = 5)
```

### Pairwise differential expression

The first DE mode compares one group against another. A common default is to
compare each timepoint against baseline while adjusting for treatment as a
covariate. In this dataset, the `0` timepoint is only present in the control
group, so these contrasts should be interpreted as overall progression away
from the untreated baseline rather than a perfectly balanced treatment-adjusted
design. With all three treatments present in the model, the treatment term
absorbs broad perturbation-specific shifts while the timepoint contrast captures
the shared movement away from baseline.

```{r eval = run_de_demo}
pairwise_time <- run_edger_pairwise_de(
  pb,
  group_col = "timepoints",
  covariates = "treatments",
  contrast_type = "reference",
  reference_level = "0",
  normalization = "TMM"
)

names(pairwise_time$tables)
```

```{r eval = run_de_demo}
pairwise_sig_counts <- aggregate(
  significant ~ contrast,
  data = pairwise_time$combined_table,
  FUN = sum
)

ggplot2::ggplot(pairwise_sig_counts, ggplot2::aes(x = contrast, y = significant, fill = contrast)) +
  ggplot2::geom_col(width = 0.72) +
  ggplot2::theme_classic() +
  ggplot2::theme(legend.position = "none") +
  ggplot2::labs(
    title = "Significant genes in timepoint-versus-baseline contrasts",
    x = "contrast",
    y = "number of DE genes"
  )
```

Inspect one contrast in more detail:

```{r eval = run_de_demo}
head(
  pairwise_time$tables[["1_vs_0"]][
    order(pairwise_time$tables[["1_vs_0"]]$FDR),
    c("gene", "logFC", "FDR")
  ],
  10
)

plot_pairwise_de_volcano(
  pairwise_time,
  contrast = "1_vs_0",
  fdr_cutoff = 0.05,
  lfc_cutoff = 0.5,
  top_n_labels = 8
)

plot_pairwise_de_ma(
  pairwise_time,
  contrast = "1_vs_0",
  fdr_cutoff = 0.05
)
```

`SCProkaR` also supports composite treatment-timepoint groups when you want an
explicit within-treatment comparison such as `PMB0.5 at day 7` versus
`PMB0.5 at day 1`, while keeping all treatment groups in the same pseudobulk
object.

```{r eval = run_de_demo}
pairwise_composite <- run_edger_pairwise_de(
  pb,
  group_col = "treatments",
  combine_group_cols = c("treatments", "timepoints"),
  contrast_type = "manual",
  manual_pairs = data.frame(
    group1 = c("PMB0.5__7", "PMB2__7", "control__7", "PMB0.5__4"),
    group2 = c("PMB0.5__1", "PMB2__1", "control__1", "PMB0.5__1"),
    stringsAsFactors = FALSE
  ),
  normalization = "TMM"
)

plot_pairwise_de_volcano(
  pairwise_composite,
  contrast = "PMB0.5__7_vs_PMB0.5__1",
  fdr_cutoff = 0.05,
  lfc_cutoff = 0.5,
  top_n_labels = 8
)
```

### Spline-based time-series differential expression

The second DE mode treats time as continuous and uses a spline basis to detect
genes that change over time within each treatment group. Because the example
object now contains all three treatments, the result returns a separate spline
fit for each perturbation trajectory.

```{r eval = run_de_demo}
spline_de <- run_edger_spline_de(
  pb,
  time_col = "timepoints",
  condition_col = "treatments",
  df = 3,
  normalization = "TMM",
  return_curve_for = "significant"
)

names(spline_de$by_condition)
```

```{r eval = run_de_demo}
spline_sig_counts <- aggregate(
  significant ~ treatments,
  data = spline_de$combined_table,
  FUN = sum
)

ggplot2::ggplot(spline_sig_counts, ggplot2::aes(x = treatments, y = significant, fill = treatments)) +
  ggplot2::geom_col(width = 0.72) +
  ggplot2::theme_classic() +
  ggplot2::theme(legend.position = "none") +
  ggplot2::labs(
    title = "Genes with significant temporal change within each treatment",
    x = "treatment",
    y = "number of spline DE genes"
  )
```

```{r eval = run_de_demo}
top_by_condition <- lapply(
  split(spline_de$combined_table, spline_de$combined_table$treatments),
  function(df) {
    keep_cols <- intersect(c("gene", "F", "PValue", "FDR"), colnames(df))
    utils::head(df[order(df$FDR), keep_cols, drop = FALSE], 100)
  }
)

top_by_condition
```

Plot fitted curves for genes with strong time-dependent changes:

```{r eval = run_de_demo}
curve_genes <- unique(unlist(lapply(top_by_condition, function(df) df$gene[seq_len(min(6, nrow(df)))])))

plot_time_series_deg_curves(
  spline_de,
  genes = curve_genes,
  ncol = 4
)
```

For users who prefer a one-step interface, `run_scproka_de()` wraps the
pseudobulk and edgeR steps into one function call.

## Time-aware trajectory analysis with TATA

`SCProkaR` also includes Time Aware Trajectory Analysis (TATA), a
PAGA-inspired trajectory abstraction method that uses the observed
experimental time labels when scoring and directing cluster-level transitions.

For the practical workflow here, we continue with the same integrated object
and run TATA on the batch-corrected latent space. This gives a compact
cluster-level trajectory graph and a pseudotime estimate that can be
interpreted on the integrated UMAP.

### Run TATA on the integrated embedding

We use the MNN-integrated embedding for graph construction and the
`timepoints` column as the real experimental time variable.

```{r eval = run_workflow}
tata <- run_tata(
  sce = sce,
  dimred = "integrated_mnn",
  time_col = "timepoints",
  cluster_col = "tata_cluster",
  k = 20,
  cluster_method = "louvain",
  alpha = 1,
  beta = 1,
  direction_threshold = 0.10,
  prune_threshold = 0.001,
  time_weight_mode = "directional_confidence",
  do_pseudotime = TRUE,
  seed = 101
)

sce_tata <- tata$sce
sce_tata$tata_cluster <- factor(sce_tata$tata_cluster)
```

The result stores the updated `SingleCellExperiment`, the cell-cell graph, the
abstract cluster graph, the cluster edge table, and the cell-level pseudotime.

```{r eval = run_workflow}
names(tata)

head(
  tata$cluster_edge_table[
    order(-tata$cluster_edge_table$tata_weight),
    c(
      "cluster_a",
      "cluster_b",
      "topology_weight",
      "flow_score",
      "time_weight",
      "tata_weight",
      "direction_label",
      "kept"
    )
  ],
  10
)
```

### Visualize TATA clusters and pseudotime

```{r eval = run_workflow}
PlotReduction(sce_tata, reduction = "umap_integrated_mnn", colour_by = "tata_cluster")
PlotReduction(sce_tata, reduction = "umap_integrated_mnn", colour_by = "tata_pseudotime_scaled")
```

### Overlay the abstract TATA graph on the embedding

`plot_tata_cluster_graph()` places the cluster-level graph directly on top of
the embedding so the abstract transitions can be interpreted in the same
coordinate system as the cells.

```{r eval = run_workflow}
plot_tata_cluster_graph(
  tata,
  dimred = "umap_integrated_mnn",
  layout_mode = "embedding",
  show_cells = TRUE,
  colour_by = "timepoints",
  node_colour_by = "cluster",
  main = "TATA abstract graph aligned to the integrated UMAP"
)
```

### Plot the inferred trajectories

For a cleaner trajectory-style summary, `plot_tata_trajectory_embedding()`
draws fitted paths on top of the embedding. The `combined` mode is especially
useful in a workflow vignette because it compresses the graph into a smaller
set of representative trajectories.

```{r eval = run_workflow}
plot_tata_trajectory_embedding(
  tata,
  dimred = "umap_integrated_mnn",
  colour_by = "timepoints",
  mode = "combined",
  max_paths = 3,
  smooth_paths = TRUE
)
```

This practical example is meant to show how TATA fits into the main
`SCProkaR` workflow on a real dataset. For the full step-by-step TATA method
walkthrough, simulated multi-branch examples, and benchmarking against other
trajectory tools, see the separate TATA vignette.


## Notes

- `CreateBacObject()` accepts Seurat input directly, but the downstream analysis
  object remains a `SingleCellExperiment`.
- `MergeBacObjects()` is the recommended merge path for multiple samples in
  `SCProkaR`.
- `PlotReduction()` is a convenient visualization helper for both original and
  integrated reductions.
- `PlotIntegrationOverview()` provides the standard Seurat-style integration
  comparison panels.
- `RunIntegratedClustering()` provides the post-integration cluster labels for
  future downstream analyses.
- `RunIntegratedUMAP()` works on integrated embeddings as well as any custom
  latent space registered with `RegisterIntegrationEmbedding()`.
- Pseudobulk differential expression is covered in the same package workflow
  via `aggregate_pseudobulk()`, `run_edger_pairwise_de()`, and
  `run_edger_spline_de()`.
- TATA can be run directly on the integrated object with `run_tata()` and
  visualized with `plot_tata_cluster_graph()` and
  `plot_tata_trajectory_embedding()`.
- The separate TATA vignette focuses on the step-wise method explanation,
  simulated data, and benchmarking.
