---
title: "SCProkaR Comprehensive Analysis Tutorial"
subtitle: "10x input, QC, integration benchmarking, visualization, clustering, markers, and functional analysis"
author: "SCProkaR developers"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    fig_caption: true
geometry: margin=0.75in
fontsize: 10pt
vignette: >
  %\VignetteIndexEntry{SCProkaR Comprehensive PDF Tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  dpi = 150,
  message = TRUE,
  warning = TRUE
)

if (file.exists("../DESCRIPTION")) {
  if (requireNamespace("devtools", quietly = TRUE)) {
    devtools::load_all("..")
  } else {
    library(SCProkaR)
  }
} else {
  library(SCProkaR)
}

run_analysis <- FALSE
```

# Purpose

This tutorial is a full practical workflow for bacterial isolate comparisons
with `SCProkaR`. It focuses on visualization because integration decisions are
much easier to defend when the plots and benchmark metrics are reviewed
together.

The workflow has two levels:

1. isolate-pair analysis: benchmark `mnn` and `harmony` within matched sample
   pairs such as control versus treatment for the same isolate
2. all-sample analysis: merge all ten samples, benchmark both integration
   methods globally, visualize per isolate/sample, then use the chosen
   integrated embedding for clustering and downstream marker analysis

By default this PDF is safe to knit without local data because
`run_analysis <- FALSE` in the setup chunk. To run it on your data, set
`run_analysis <- TRUE` and update the sample paths below.

# Package Requirements

```{r libraries, eval=run_analysis}
library(SCProkaR)
library(SingleCellExperiment)
library(SummarizedExperiment)
library(ggplot2)
```

Optional packages used in selected sections:

```{r optional-packages, eval=run_analysis}
requireNamespace("batchelor", quietly = TRUE)
requireNamespace("harmony", quietly = TRUE)
requireNamespace("uwot", quietly = TRUE)
requireNamespace("scran", quietly = TRUE)
requireNamespace("patchwork", quietly = TRUE)
requireNamespace("clusterProfiler", quietly = TRUE)
```

The unintegrated and integrated workflows do not require Seurat. Seurat input is
supported through `CreateBacObject()` when `SeuratObject` is installed, but the
core object used by the package is always a `SingleCellExperiment`.

# Input Definition

This example uses ten Cell Ranger / 10x-style directories. The recommended
input is the filtered matrix directory for each sample.

```{r sample-info, eval=run_analysis}
sample_info <- data.frame(
  sample_id = c(
    "S39_C", "S39_T", "SPAO1_C", "SPAO1_T", "S25_C",
    "S25_T", "S44_C", "S44_T", "S46_C", "S46_T"
  ),
  isolate = c(
    "S39", "S39", "SPAO1", "SPAO1", "S25",
    "S25", "S44", "S44", "S46", "S46"
  ),
  condition = c(
    "C", "T", "C", "T", "C",
    "T", "C", "T", "C", "T"
  ),
  path = c(
    "A4131matrix/newFiltered/",
    "A4007matrix/newFiltered/",
    "A3806matrix/newFiltered/",
    "A3805matrix/newFiltered/",
    "A3808matrix/newFiltered/",
    "A3807matrix/newFiltered/",
    "A4008matrix/newFiltered/",
    "A4009matrix/newFiltered/",
    "A4010matrix/newFiltered/",
    "A4011matrix/newFiltered/"
  ),
  stringsAsFactors = FALSE
)

sample_info
```

# Create SCProkaR Objects

`CreateBacObject()` can read a 10x matrix directory directly. For 10x-derived
objects, `Symbol` is preferred as the feature-name column when available. If
your data uses a different column, change `feature_name_col`.

This tutorial sets `run_unintegrated = TRUE` so each sample gets a first-pass
PCA, clustering, and UMAP for QC visualization. For production analysis, the
final clustering is still performed after integration.

```{r create-objects, eval=run_analysis}
sce_list <- setNames(
  lapply(seq_len(nrow(sample_info)), function(i) {
    CreateBacObject(
      sample_info$path[i],
      sample_id_value = sample_info$sample_id[i],
      batch_value = sample_info$sample_id[i],
      condition_value = sample_info$condition[i],
      feature_name_col = "Symbol",
      run_unintegrated = TRUE,
      unintegrated_dims = 1:30,
      unintegrated_cluster_col = "unintegrated_clusters",
      unintegrated_umap_name = "umap.unintegrated",
      unintegrated_k = 15,
      unintegrated_resolution = 1
    )
  }),
  sample_info$sample_id
)

sce_list <- lapply(names(sce_list), function(nm) {
  sce <- sce_list[[nm]]
  sce$isolate <- sample_info$isolate[match(nm, sample_info$sample_id)]
  sce$condition <- sample_info$condition[match(nm, sample_info$sample_id)]
  sce$sample <- sce$sample_id
  sce
})
names(sce_list) <- sample_info$sample_id
```

# Per-Sample QC

```{r run-qc, eval=run_analysis}
sce_list <- lapply(sce_list, RunBacQC)

qc_df <- do.call(
  rbind,
  lapply(names(sce_list), function(nm) {
    sce <- sce_list[[nm]]
    counts_mat <- SummarizedExperiment::assay(sce, "counts")
    data.frame(
      cell = colnames(sce),
      sample_id = nm,
      isolate = unique(as.character(sce$isolate)),
      condition = unique(as.character(sce$condition)),
      rrna_fraction = sce$rrna_fraction,
      ribo_fraction = sce$ribo_fraction,
      nCount_RNA = Matrix::colSums(counts_mat),
      nFeature_RNA = Matrix::colSums(counts_mat > 0),
      stringsAsFactors = FALSE
    )
  })
)

qc_df$sample_id <- factor(qc_df$sample_id, levels = sample_info$sample_id)
```

## QC Visualizations Before Filtering

```{r qc-before-plots, eval=run_analysis, fig.width=10, fig.height=6}
ggplot(qc_df, aes(x = sample_id, y = rrna_fraction, fill = sample_id)) +
  geom_violin(scale = "width", trim = TRUE) +
  geom_boxplot(width = 0.12, outlier.size = 0.2, fill = "white") +
  theme_classic(base_size = 10) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "none") +
  labs(x = "Sample", y = "rRNA fraction", title = "rRNA fraction before filtering")

ggplot(qc_df, aes(x = sample_id, y = nFeature_RNA, fill = sample_id)) +
  geom_violin(scale = "width", trim = TRUE) +
  geom_boxplot(width = 0.12, outlier.size = 0.2, fill = "white") +
  theme_classic(base_size = 10) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "none") +
  labs(x = "Sample", y = "Detected features", title = "Detected features before filtering")
```

# Filter Cells

Adjust these thresholds for your dataset. The example values are conservative
starting points.

```{r filter-cells, eval=run_analysis}
sce_filtered_list <- setNames(
  lapply(sce_list, function(sce) {
    FilterBacCells(
      sce,
      min_counts = 50,
      min_features = 50,
      max_rrna_fraction = 0.10
    )
  }),
  names(sce_list)
)

qc_df_after <- do.call(
  rbind,
  lapply(names(sce_filtered_list), function(nm) {
    sce <- sce_filtered_list[[nm]]
    counts_mat <- SummarizedExperiment::assay(sce, "counts")
    data.frame(
      cell = colnames(sce),
      sample_id = nm,
      isolate = unique(as.character(sce$isolate)),
      condition = unique(as.character(sce$condition)),
      rrna_fraction = sce$rrna_fraction,
      ribo_fraction = sce$ribo_fraction,
      nCount_RNA = Matrix::colSums(counts_mat),
      nFeature_RNA = Matrix::colSums(counts_mat > 0),
      stringsAsFactors = FALSE
    )
  })
)

qc_df$stage <- "Before"
qc_df_after$stage <- "After"
qc_df_all <- rbind(qc_df, qc_df_after)
qc_df_all$sample_id <- factor(qc_df_all$sample_id, levels = sample_info$sample_id)
qc_df_all$stage <- factor(qc_df_all$stage, levels = c("Before", "After"))
```

## QC Visualizations Before and After Filtering

```{r qc-before-after-plots, eval=run_analysis, fig.width=11, fig.height=6}
ggplot(qc_df_all, aes(x = sample_id, y = rrna_fraction, fill = stage)) +
  geom_boxplot(outlier.size = 0.15) +
  theme_classic(base_size = 10) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(x = "Sample", y = "rRNA fraction", title = "rRNA fraction before and after filtering")

ggplot(qc_df_all, aes(x = sample_id, y = nFeature_RNA, fill = stage)) +
  geom_boxplot(outlier.size = 0.15) +
  theme_classic(base_size = 10) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(x = "Sample", y = "Detected features", title = "Detected features before and after filtering")
```

# Build Isolate-Pair Objects

Each isolate pair compares matched control and treatment samples.

```{r pair-objects, eval=run_analysis}
merged_list <- list(
  sce_1_2 = MergeBacObjects(
    objects = sce_filtered_list[c("S39_C", "S39_T")],
    gene_mode = "intersect"
  ),
  sce_3_4 = MergeBacObjects(
    objects = sce_filtered_list[c("SPAO1_C", "SPAO1_T")],
    gene_mode = "intersect"
  ),
  sce_5_6 = MergeBacObjects(
    objects = sce_filtered_list[c("S25_C", "S25_T")],
    gene_mode = "intersect"
  ),
  sce_7_8 = MergeBacObjects(
    objects = sce_filtered_list[c("S44_C", "S44_T")],
    gene_mode = "intersect"
  ),
  sce_9_10 = MergeBacObjects(
    objects = sce_filtered_list[c("S46_C", "S46_T")],
    gene_mode = "intersect"
  )
)

pair_list <- merged_list
```

## Sanity Check

```{r pair-sanity, eval=run_analysis}
lapply(pair_list, function(sce) {
  list(
    dims = dim(sce),
    coldata = colnames(SummarizedExperiment::colData(sce)),
    reductions = reducedDimNames(sce),
    sample_table = table(sce$sample_id)
  )
})
```

# Pair-Level Integration Benchmark

This helper runs `mnn` and `harmony`, computes UMAPs, clusters both integrated
spaces, runs `BenchmarkIntegration()`, and prepares the core plots for review.

```{r pair-benchmark-function, eval=run_analysis}
run_pair_benchmark <- function(
    sce,
    pair_name,
    batch_col = "batch",
    sample_col = "sample_id",
    label_col = "unintegrated_clusters",
    dims = 1:30,
    k = 15,
    resolution = 1
) {
  message("Running pair: ", pair_name)

  cd <- as.data.frame(SummarizedExperiment::colData(sce))
  if (!batch_col %in% colnames(cd)) stop("Missing batch column: ", batch_col)
  if (!sample_col %in% colnames(cd)) stop("Missing sample column: ", sample_col)

  label_col_use <- if (label_col %in% colnames(cd)) label_col else NULL

  sce <- IntegrateBacData(sce, batch_col = batch_col, method = "mnn", dims = dims)
  sce <- IntegrateBacData(sce, batch_col = batch_col, method = "harmony", dims = dims)
  sce <- RunIntegratedUMAP(sce, reduction = "integrated_mnn")
  sce <- RunIntegratedUMAP(sce, reduction = "integrated_harmony")

  sce <- RunIntegratedClustering(
    sce,
    reduction = "integrated_mnn",
    cluster_col = "integrated_clusters_mnn",
    dims = dims,
    k = k,
    resolution = resolution
  )
  sce <- RunIntegratedClustering(
    sce,
    reduction = "integrated_harmony",
    cluster_col = "integrated_clusters_harmony",
    dims = dims,
    k = k,
    resolution = resolution
  )

  bench <- BenchmarkIntegration(
    sce,
    batch_col = batch_col,
    label_col = label_col_use,
    methods = c("integrated_mnn", "integrated_harmony"),
    return_plots = TRUE
  )
  sce <- bench$sce

  unintegrated_reduction <- if ("umap.unintegrated" %in% reducedDimNames(sce)) {
    "umap.unintegrated"
  } else if ("PCA" %in% reducedDimNames(sce)) {
    "PCA"
  } else {
    NULL
  }

  plots <- list(
    unintegrated_by_sample = if (!is.null(unintegrated_reduction)) {
      PlotReduction(sce, reduction = unintegrated_reduction, colour_by = sample_col)
    } else NULL,
    mnn_by_sample = PlotReduction(sce, reduction = "umap_integrated_mnn", colour_by = sample_col),
    harmony_by_sample = PlotReduction(sce, reduction = "umap_integrated_harmony", colour_by = sample_col),
    unintegrated_by_cluster = if (!is.null(unintegrated_reduction) && !is.null(label_col_use)) {
      PlotReduction(sce, reduction = unintegrated_reduction, colour_by = label_col_use)
    } else NULL,
    mnn_by_cluster = PlotReduction(sce, reduction = "umap_integrated_mnn", colour_by = "integrated_clusters_mnn"),
    harmony_by_cluster = PlotReduction(sce, reduction = "umap_integrated_harmony", colour_by = "integrated_clusters_harmony"),
    mnn_facet = PlotReduction(
      sce,
      reduction = "umap_integrated_mnn",
      colour_by = "integrated_clusters_mnn",
      facet_by = sample_col
    ),
    harmony_facet = PlotReduction(
      sce,
      reduction = "umap_integrated_harmony",
      colour_by = "integrated_clusters_harmony",
      facet_by = sample_col
    )
  )

  ranking <- bench$ranking
  ranking$pair <- pair_name
  scores <- bench$scores
  scores$pair <- pair_name

  list(sce = sce, benchmark = bench, ranking = ranking, scores = scores, plots = plots)
}
```

```{r run-pair-benchmarks, eval=run_analysis}
pair_results <- lapply(names(pair_list), function(nm) {
  run_pair_benchmark(
    sce = pair_list[[nm]],
    pair_name = nm,
    batch_col = "batch",
    sample_col = "sample_id",
    label_col = "unintegrated_clusters",
    dims = 1:30,
    k = 15,
    resolution = 1
  )
})
names(pair_results) <- names(pair_list)

ranking_all <- do.call(rbind, lapply(pair_results, function(x) x$ranking))
scores_all <- do.call(rbind, lapply(pair_results, function(x) x$scores))

ranking_all
```

## Pair-Level Benchmark Figures

```{r pair-benchmark-example-plots, eval=run_analysis, fig.width=10, fig.height=7}
pair_results[["sce_1_2"]]$benchmark$plots$key_metrics
pair_results[["sce_1_2"]]$benchmark$plots$tradeoff
pair_results[["sce_1_2"]]$plots$unintegrated_by_sample
pair_results[["sce_1_2"]]$plots$mnn_by_sample
pair_results[["sce_1_2"]]$plots$harmony_by_sample
```

## Save Pair-Level Outputs

```{r save-pair-outputs, eval=run_analysis}
pair_outdir <- "integration_benchmark_by_isolate"
dir.create(pair_outdir, showWarnings = FALSE, recursive = TRUE)

write.csv(ranking_all, file.path(pair_outdir, "pair_benchmark_ranking.csv"), row.names = FALSE)
write.csv(scores_all, file.path(pair_outdir, "pair_benchmark_scores.csv"), row.names = FALSE)

for (nm in names(pair_results)) {
  res <- pair_results[[nm]]

  if (!is.null(res$plots$unintegrated_by_sample)) {
    ggsave(file.path(pair_outdir, paste0(nm, "_unintegrated_by_sample.pdf")),
      res$plots$unintegrated_by_sample, width = 6, height = 5
    )
  }
  ggsave(file.path(pair_outdir, paste0(nm, "_mnn_by_sample.pdf")),
    res$plots$mnn_by_sample, width = 6, height = 5
  )
  ggsave(file.path(pair_outdir, paste0(nm, "_harmony_by_sample.pdf")),
    res$plots$harmony_by_sample, width = 6, height = 5
  )
  ggsave(file.path(pair_outdir, paste0(nm, "_benchmark_key_metrics.pdf")),
    res$benchmark$plots$key_metrics, width = 10, height = 7
  )
  ggsave(file.path(pair_outdir, paste0(nm, "_benchmark_tradeoff.pdf")),
    res$benchmark$plots$tradeoff, width = 6, height = 5
  )
}
```

# All-Sample Integration Benchmark

Pair-level benchmarking shows method behavior within each isolate. The next
step asks whether the same choice behaves well when all ten samples are merged.

```{r merge-all-samples, eval=run_analysis}
sce_all <- MergeBacObjects(objects = sce_filtered_list, gene_mode = "intersect")
sce_all
table(sce_all$sample_id)
```

```{r integrate-all-samples, eval=run_analysis}
sce_all <- IntegrateBacData(sce_all, batch_col = "batch", method = "mnn", dims = 1:30)
sce_all <- IntegrateBacData(sce_all, batch_col = "batch", method = "harmony", dims = 1:30)
sce_all <- RunIntegratedUMAP(sce_all, reduction = "integrated_mnn")
sce_all <- RunIntegratedUMAP(sce_all, reduction = "integrated_harmony")

sce_all <- RunIntegratedClustering(
  sce_all,
  reduction = "integrated_mnn",
  cluster_col = "integrated_clusters_mnn",
  dims = 1:30,
  k = 15,
  resolution = 1
)
sce_all <- RunIntegratedClustering(
  sce_all,
  reduction = "integrated_harmony",
  cluster_col = "integrated_clusters_harmony",
  dims = 1:30,
  k = 15,
  resolution = 1
)
```

```{r benchmark-all-samples, eval=run_analysis}
bench_all <- BenchmarkIntegration(
  sce_all,
  batch_col = "batch",
  label_col = "unintegrated_clusters",
  methods = c("integrated_mnn", "integrated_harmony"),
  return_plots = TRUE
)

bench_all$ranking
```

## All-Sample Benchmark Figures

```{r all-benchmark-plots, eval=run_analysis, fig.width=10, fig.height=7}
bench_all$plots$key_metrics
bench_all$plots$batch_removal
bench_all$plots$bio_conservation
bench_all$plots$tradeoff
bench_all$plots$heatmap
```

# All-Sample Visualization

These plots compare the unintegrated view against MNN and Harmony after
integration. They are saved both as combined all-cell views and as faceted
per-sample views.

```{r all-visualization-plots, eval=run_analysis, fig.width=9, fig.height=6}
p_unint_sample <- PlotReduction(sce_all, reduction = "umap.unintegrated", colour_by = "sample_id")
p_mnn_sample <- PlotReduction(sce_all, reduction = "umap_integrated_mnn", colour_by = "sample_id")
p_harmony_sample <- PlotReduction(sce_all, reduction = "umap_integrated_harmony", colour_by = "sample_id")

p_unint_cluster <- PlotReduction(sce_all, reduction = "umap.unintegrated", colour_by = "unintegrated_clusters")
p_mnn_cluster <- PlotReduction(sce_all, reduction = "umap_integrated_mnn", colour_by = "integrated_clusters_mnn")
p_harmony_cluster <- PlotReduction(sce_all, reduction = "umap_integrated_harmony", colour_by = "integrated_clusters_harmony")

p_unint_sample
p_mnn_sample
p_harmony_sample
p_unint_cluster
p_mnn_cluster
p_harmony_cluster
```

## Per-Sample Faceted Views

```{r all-faceted-plots, eval=run_analysis, fig.width=13, fig.height=8}
p_unint_facet <- PlotReduction(
  sce_all,
  reduction = "umap.unintegrated",
  colour_by = "unintegrated_clusters",
  facet_by = "sample_id"
)
p_mnn_facet <- PlotReduction(
  sce_all,
  reduction = "umap_integrated_mnn",
  colour_by = "integrated_clusters_mnn",
  facet_by = "sample_id"
)
p_harmony_facet <- PlotReduction(
  sce_all,
  reduction = "umap_integrated_harmony",
  colour_by = "integrated_clusters_harmony",
  facet_by = "sample_id"
)

p_unint_facet
p_mnn_facet
p_harmony_facet
```

## Save All-Sample Outputs

```{r save-all-sample-outputs, eval=run_analysis}
all_outdir <- "integration_benchmark_all_samples"
dir.create(all_outdir, showWarnings = FALSE, recursive = TRUE)

write.csv(bench_all$scores, file.path(all_outdir, "benchmark_scores_all_samples.csv"), row.names = FALSE)
write.csv(bench_all$ranking, file.path(all_outdir, "benchmark_ranking_all_samples.csv"), row.names = FALSE)

ggsave(file.path(all_outdir, "benchmark_key_metrics.pdf"), bench_all$plots$key_metrics, width = 10, height = 7)
ggsave(file.path(all_outdir, "benchmark_tradeoff.pdf"), bench_all$plots$tradeoff, width = 6, height = 5)
ggsave(file.path(all_outdir, "all_unintegrated_by_sample.pdf"), p_unint_sample, width = 7, height = 5)
ggsave(file.path(all_outdir, "all_mnn_by_sample.pdf"), p_mnn_sample, width = 7, height = 5)
ggsave(file.path(all_outdir, "all_harmony_by_sample.pdf"), p_harmony_sample, width = 7, height = 5)
ggsave(file.path(all_outdir, "all_unintegrated_by_cluster.pdf"), p_unint_cluster, width = 7, height = 5)
ggsave(file.path(all_outdir, "all_mnn_by_cluster.pdf"), p_mnn_cluster, width = 7, height = 5)
ggsave(file.path(all_outdir, "all_harmony_by_cluster.pdf"), p_harmony_cluster, width = 7, height = 5)
ggsave(file.path(all_outdir, "all_unintegrated_facet_by_sample.pdf"), p_unint_facet, width = 14, height = 8)
ggsave(file.path(all_outdir, "all_mnn_facet_by_sample.pdf"), p_mnn_facet, width = 14, height = 8)
ggsave(file.path(all_outdir, "all_harmony_facet_by_sample.pdf"), p_harmony_facet, width = 14, height = 8)
```

# Choose Final Integration Method

Pick the method that is supported by the pair-level benchmark, all-sample
benchmark, and visual inspection. The code below picks the top-ranked global
method automatically.

```{r choose-method, eval=run_analysis}
chosen_method <- bench_all$ranking$method[[which.min(bench_all$ranking$rank)]]

chosen_cfg <- if (chosen_method == "integrated_mnn") {
  list(
    reduction = "integrated_mnn",
    umap = "umap_integrated_mnn",
    cluster_col = "integrated_clusters_mnn",
    short = "mnn"
  )
} else {
  list(
    reduction = "integrated_harmony",
    umap = "umap_integrated_harmony",
    cluster_col = "integrated_clusters_harmony",
    short = "harmony"
  )
}

chosen_cfg
```

```{r final-cluster-plots, eval=run_analysis, fig.width=13, fig.height=8}
p_final_clusters <- PlotReduction(
  sce_all,
  reduction = chosen_cfg$umap,
  colour_by = chosen_cfg$cluster_col
)
p_final_clusters_facet <- PlotReduction(
  sce_all,
  reduction = chosen_cfg$umap,
  colour_by = chosen_cfg$cluster_col,
  facet_by = "sample_id"
)

p_final_clusters
p_final_clusters_facet
```

# Marker Genes for Integrated Clusters

The marker analysis uses the chosen integrated cluster labels and `logcounts`.
This is useful for cluster interpretation. For formal differential expression,
use the pseudobulk workflow documented separately.

```{r marker-function, eval=run_analysis}
find_cluster_markers <- function(sce, cluster_col, top_n = 20) {
  if (!requireNamespace("scran", quietly = TRUE)) {
    stop("Install 'scran' for marker-gene analysis.")
  }

  clusters <- as.factor(SummarizedExperiment::colData(sce)[[cluster_col]])
  marker_list <- scran::findMarkers(
    sce,
    groups = clusters,
    assay.type = "logcounts",
    direction = "up"
  )

  marker_tables <- lapply(names(marker_list), function(cl) {
    df <- as.data.frame(marker_list[[cl]])
    df$gene <- rownames(df)
    df$cluster <- cl

    rank_col <- intersect(
      c("Top", "FDR", "summary.logFC", "mean.logFC.cohen", "mean.AUC"),
      colnames(df)
    )
    if (length(rank_col) > 0) {
      rank_col <- rank_col[[1]]
      df <- if (rank_col %in% c("Top", "FDR")) {
        df[order(df[[rank_col]], decreasing = FALSE), , drop = FALSE]
      } else {
        df[order(df[[rank_col]], decreasing = TRUE), , drop = FALSE]
      }
    }
    utils::head(df, top_n)
  })

  all_cols <- unique(unlist(lapply(marker_tables, colnames), use.names = FALSE))
  marker_tables <- lapply(marker_tables, function(df) {
    missing_cols <- setdiff(all_cols, colnames(df))
    for (nm in missing_cols) df[[nm]] <- NA
    df[, all_cols, drop = FALSE]
  })

  out <- do.call(rbind, marker_tables)
  rownames(out) <- NULL
  out
}
```

```{r run-markers, eval=run_analysis}
top_markers_all <- find_cluster_markers(
  sce_all,
  cluster_col = chosen_cfg$cluster_col,
  top_n = 20
)

utils::head(top_markers_all, 20)
write.csv(
  top_markers_all,
  file.path(all_outdir, paste0("top_markers_", chosen_cfg$short, "_clusters.csv")),
  row.names = FALSE
)
```

# Functional Enrichment

Functional enrichment requires a gene-to-term annotation table. For bacterial
datasets this is usually isolate-specific or database-specific, so SCProkaR
does not assume a built-in annotation source.

Expected format:

```{r term2gene-example, eval=FALSE}
term2gene <- data.frame(
  term = c("DNA replication", "DNA replication", "Ribosome"),
  gene = c("dnaA", "dnaN", "rplB"),
  stringsAsFactors = FALSE
)
```

```{r enrichment-function, eval=run_analysis}
run_functional_enrichment <- function(marker_df, term2gene, top_n = 100) {
  if (!requireNamespace("clusterProfiler", quietly = TRUE)) {
    stop("Install 'clusterProfiler' for functional enrichment.")
  }

  clusters <- unique(marker_df$cluster)
  enrich_list <- lapply(clusters, function(cl) {
    genes <- marker_df$gene[marker_df$cluster == cl]
    genes <- unique(utils::head(genes, top_n))

    enr <- clusterProfiler::enricher(
      gene = genes,
      TERM2GENE = term2gene
    )
    if (is.null(enr)) return(NULL)

    out <- as.data.frame(enr)
    out$cluster <- cl
    out
  })

  enrich_list <- enrich_list[!vapply(enrich_list, is.null, logical(1))]
  if (!length(enrich_list)) return(NULL)
  do.call(rbind, enrich_list)
}
```

Run the enrichment once you have a real annotation table:

```{r run-enrichment, eval=FALSE}
functional_all <- run_functional_enrichment(
  marker_df = top_markers_all,
  term2gene = term2gene,
  top_n = 100
)

write.csv(
  functional_all,
  file.path(all_outdir, paste0("functional_enrichment_", chosen_cfg$short, ".csv")),
  row.names = FALSE
)
```

# Save Final Objects

```{r save-final-objects, eval=run_analysis}
saveRDS(sce_all, file = file.path(all_outdir, "sce_all_integrated.rds"))
saveRDS(bench_all, file = file.path(all_outdir, "bench_all.rds"))
saveRDS(pair_results, file = file.path(pair_outdir, "pair_results.rds"))
```

# Interpretation Checklist

Use the plots and tables together:

- QC plots: confirm filtering removes low-quality cells without removing whole
  sample groups unexpectedly
- unintegrated by sample: shows the pre-correction sample structure
- MNN/Harmony by sample: should show improved batch mixing
- MNN/Harmony by cluster: should preserve coherent biological structure
- faceted sample plots: confirm no sample disappears from major states
- benchmark ranking: useful summary, but do not treat the mean score as the only
  decision criterion
- marker genes: use for cluster interpretation
- functional enrichment: use only with an annotation table appropriate for the
  isolate or gene namespace

# Render This Tutorial

From the package root:

```{r render-command, eval=FALSE}
rmarkdown::render(
  "vignettes/scprokar-pdf-tutorial.Rmd",
  output_format = "pdf_document"
)
```

If LaTeX is not available, install TinyTeX once:

```{r tinytex-command, eval=FALSE}
install.packages("tinytex")
tinytex::install_tinytex()
```
