---
title: "SCProkaR Core Workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{SCProkaR Core Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 5
)

if (file.exists("../DESCRIPTION")) {
  if (requireNamespace("devtools", quietly = TRUE)) {
    devtools::load_all("..")
  } else {
    library(SCProkaR)
  }
} else {
  library(SCProkaR)
}

run_workflow <- TRUE
run_de_demo <- TRUE
```

## Overview

This vignette shows two practical `SCProkaR` workflows:

- a two-sample preprocessing and integration workflow starting from Seurat
  `.rds` objects
- a pseudobulk differential-expression workflow on a simulated multi-condition
  time course

`SCProkaR` keeps the main analysis object as a
`SingleCellExperiment`, while still accepting Seurat input for convenient
conversion.

## Load packages

```{r}
library(SummarizedExperiment)
library(SingleCellExperiment)
```

## Load Seurat `.rds` objects

Replace the file paths below with your own sample files.

```{r eval = run_workflow}
seu1 <- readRDS("../inst/extdata/sample1.rds")
seu2 <- readRDS("../inst/extdata/sample2.rds")
```

For clarity, add sample origin columns if they do not already exist:

```{r eval = run_workflow}
seu1$sample_id <- "sample1"
seu1$batch <- "sample1"

seu2$sample_id <- "sample2"
seu2$batch <- "sample2"
```

## Create SCProkaR objects

`CreateBacObject()` can extract counts, metadata, and existing reductions such
as `pca` and `umap` directly from each Seurat object.

```{r eval = run_workflow}
sce1 <- CreateBacObject(
  seu1,
  seurat_assay = "RNA",
  seurat_layer = "counts",
  sample_col = "sample_id",
  batch_col = "batch"
)

sce2 <- CreateBacObject(
  seu2,
  seurat_assay = "RNA",
  seurat_layer = "counts",
  sample_col = "sample_id",
  batch_col = "batch"
)
```

## Quality control

Compute QC metrics on each sample separately:

```{r eval = run_workflow}
sce1 <- RunBacQC(sce1)
sce2 <- RunBacQC(sce2)
```

The QC results are stored in `colData(sce)`:

```{r eval = run_workflow}
head(as.data.frame(colData(sce1))[, c(
  "total_counts",
  "detected_features",
  "rrna_fraction",
  "pct_rrna"
)])
```

Filter cells using your chosen thresholds:

```{r eval = run_workflow}
sce1 <- FilterBacCells(
  sce1,
  min_counts = 50,
  min_features = 50,
  max_rrna_fraction = 0.20
)

sce2 <- FilterBacCells(
  sce2,
  min_counts = 50,
  min_features = 50,
  max_rrna_fraction = 0.20
)
```

## Merge samples

Use `MergeBacObjects()` instead of `cbind()` to avoid duplicate-barcode issues
and to align genes safely across samples.

```{r eval = run_workflow}
sce <- MergeBacObjects(sce1, sce2, gene_mode = "intersect")
```

If cell barcodes overlap between samples, `MergeBacObjects()` automatically
makes them unique and stores the original barcode in
`colData(sce)$original_cell_id`.

If both input objects contain compatible reductions such as `pca` and `umap`,
`MergeBacObjects()` carries them forward into the merged object.

## Visualize unintegrated structure

If your Seurat objects already contained a UMAP, it is typically available after
conversion:

```{r eval = run_workflow}
PlotReduction(sce, reduction = "umap", colour_by = "batch")
PlotReduction(sce, reduction = "umap", colour_by = "seurat_clusters")
```

Interpretation:

- colouring by `batch` shows whether samples are separated before integration
- colouring by `seurat_clusters` or another biological label shows whether
  expected structure is present

`PlotReduction(..., facet_by = "batch")` is the `SCProkaR` equivalent of
Seurat's `split.by` visualization.

## Run integration

### Option 1: MNN

```{r eval = run_workflow}
sce <- IntegrateBacData(
  sce,
  batch_col = "batch",
  method = "mnn",
  dims = 1:30
)
```

### Option 2: Harmony

```{r eval = run_workflow}
sce <- IntegrateBacData(
  sce,
  batch_col = "batch",
  method = "harmony",
  dims = 1:30
)
```

The output embedding is stored as `integrated_mnn` or `integrated_harmony`.

## Visualize the integrated latent space

```{r eval = run_workflow}
PlotReduction(sce, reduction = "integrated_mnn", colour_by = "batch")
PlotReduction(sce, reduction = "integrated_mnn", colour_by = "seurat_clusters")
PlotReduction(sce, reduction = "integrated_harmony", colour_by = "batch")
PlotReduction(sce, reduction = "integrated_harmony", colour_by = "seurat_clusters")
```

This gives a quick direct view of the integrated latent space before running a
new UMAP.

## Run UMAP on the integrated embedding

`RunIntegratedUMAP()` computes a UMAP from any stored embedding:

```{r eval = run_workflow}
sce <- RunIntegratedUMAP(sce, reduction = "integrated_mnn")
sce <- RunIntegratedUMAP(sce, reduction = "integrated_harmony")
```

This stores a new reduced dimension named `umap_integrated_mnn`.

```{r eval = run_workflow}
PlotReduction(sce, reduction = "umap_integrated_mnn", colour_by = "batch")
PlotReduction(sce, reduction = "umap_integrated_mnn", colour_by = "seurat_clusters")
PlotReduction(sce, reduction = "umap_integrated_harmony", colour_by = "batch")
PlotReduction(sce, reduction = "umap_integrated_harmony", colour_by = "seurat_clusters")
```

## Cluster cells after integration

Future downstream analyses should use clusters identified from the integrated
embedding rather than the pre-integration batch structure.

`RunIntegratedClustering()` builds a kNN graph on the corrected embedding and
stores the resulting labels in `colData(sce)`.

```{r eval = run_workflow}
sce <- RunIntegratedClustering(
  sce,
  reduction = "integrated_mnn",
  cluster_col = "integrated_clusters",
  k = 20,
  resolution = 1
)
```

Here, `k` controls the neighbour graph size, while `resolution` controls how
finely that graph is partitioned into clusters. This mirrors the Seurat logic
of `FindNeighbors(k.param = ...)` followed by `FindClusters(resolution = ...)`.

Visualize the integrated clusters on the corrected UMAP:

```{r eval = run_workflow}
PlotReduction(sce, reduction = "umap_integrated_mnn", colour_by = "integrated_clusters")
PlotReduction(
  sce,
  reduction = "umap_integrated_mnn",
  colour_by = "integrated_clusters",
  facet_by = "batch"
)
```

## Seurat-style integration summary panels

The Seurat integration tutorial emphasizes three visual checks after alignment:

- compare unintegrated and integrated embeddings coloured by batch
- compare unintegrated and integrated embeddings coloured by cluster or label
- split the integrated embedding by batch/condition while keeping cluster
  colours fixed

`SCProkaR` provides `PlotIntegrationOverview()` for this pattern.

### MNN example

```{r eval = run_workflow}
mnn_plots <- PlotIntegrationOverview(
  sce,
  unintegrated_reduction = "umap",
  integrated_reduction = "umap_integrated_mnn",
  batch_col = "batch",
  label_col = "integrated_clusters",
  split_by = "batch"
)

mnn_plots$unintegrated_by_batch
mnn_plots$integrated_by_batch
mnn_plots$unintegrated_by_label
mnn_plots$integrated_by_label
mnn_plots$integrated_split
```

### Harmony example

```{r eval = run_workflow}
harmony_plots <- PlotIntegrationOverview(
  sce,
  unintegrated_reduction = "umap",
  integrated_reduction = "umap_integrated_harmony",
  batch_col = "batch",
  label_col = "integrated_clusters",
  split_by = "batch"
)

harmony_plots$unintegrated_by_batch
harmony_plots$integrated_by_batch
harmony_plots$unintegrated_by_label
harmony_plots$integrated_by_label
harmony_plots$integrated_split
```

Interpretation:

- `unintegrated_by_batch` should show stronger sample separation than
  `integrated_by_batch`
- `integrated_by_label` should preserve biologically meaningful group
  separation
- `integrated_split` is the closest equivalent to Seurat's
  `DimPlot(..., split.by = "stim")`

After successful integration:

- batches should mix better than in the original UMAP
- biological clusters should remain interpretable

## Benchmark integration quality

If you have a biological label such as `seurat_clusters`, you can benchmark the
integration with the package's scIB-inspired metric panel:

```{r eval = run_workflow}
bench <- BenchmarkIntegration(
  sce,
  batch_col = "batch",
  label_col = "seurat_clusters",
  methods = NULL
)

bench$ranking
bench$plots$overall
bench$plots$key_metrics
bench$plots$batch_removal
bench$plots$bio_conservation
bench$plots$tradeoff
bench$plots$heatmap
```

The key comparison figures are:

- `key_metrics`: per-metric score comparison across methods
- `batch_removal`: average of the batch-removal metrics
- `bio_conservation`: average of the biological-conservation metrics
- `tradeoff`: batch-removal versus bio-conservation, which is often the most
  useful summary of integration quality

## Pseudobulk differential expression

`SCProkaR` also includes pseudobulk differential-expression utilities built
around edgeR. For this section we switch to a simulated multi-drug time-course
 because it contains treatment labels, repeated timepoints, and known temporal
 structure out of the box.

### Simulate a multi-condition time course

```{r eval = run_de_demo}
set.seed(101)

sce_de <- simulate_tata_multidrug_sce(
  n_cells = 1800,
  n_features = 80,
  n_pcs = 20,
  seed = 101
)

sce_de$timepoint_factor <- factor(
  sce_de$timepoint,
  levels = sort(unique(sce_de$timepoint))
)
```

Visualize the simulated embedding before aggregation:

```{r eval = run_de_demo}
PlotReduction(sce_de, reduction = "UMAP", colour_by = "condition")
PlotReduction(sce_de, reduction = "UMAP", colour_by = "timepoint_factor")
```

### Create pseudo-replicate sample labels

Because the simulator returns cells rather than biological replicates, we split
cells inside each `condition x timepoint` stratum into pseudo-replicates. In a
real experiment you would usually aggregate by the true biological sample,
donor, or experimental unit.

```{r eval = run_de_demo}
assign_pseudoreplicates <- function(
    sce,
    group_cols = c("condition", "timepoint"),
    n_reps = 3,
    seed = 101
) {
  set.seed(seed)

  meta <- as.data.frame(SummarizedExperiment::colData(sce))
  strata <- interaction(meta[, group_cols, drop = FALSE], drop = TRUE, lex.order = TRUE)
  rep_id <- integer(ncol(sce))

  for (level_name in levels(strata)) {
    idx <- which(strata == level_name)
    rep_id[idx] <- sample(rep(seq_len(n_reps), length.out = length(idx)))
  }

  sce$pb_replicate <- factor(rep_id, levels = seq_len(n_reps))
  sce$pb_sample_id <- paste(
    "pb",
    as.character(sce$condition),
    paste0("t", sce$timepoint),
    paste0("r", sce$pb_replicate),
    sep = "_"
  )

  sce
}

sce_de <- assign_pseudoreplicates(sce_de, n_reps = 3, seed = 101)

with(
  as.data.frame(SummarizedExperiment::colData(sce_de)),
  table(condition, timepoint, pb_replicate)[, , 1]
)
```

### Aggregate pseudobulk samples

`aggregate_pseudobulk()` sums counts across cells that share the same sample
definition and keeps the sample-level metadata alongside the aggregated counts.

```{r eval = run_de_demo}
pb <- aggregate_pseudobulk(
  sce_de,
  sample_cols = c("pb_sample_id", "condition", "timepoint", "pb_replicate"),
  aggregation = "sum",
  min_cells = 15
)

pb <- normalize_pseudobulk(pb, method = "TMM")
pb$timepoint_factor <- factor(pb$timepoint, levels = sort(unique(pb$timepoint)))

pb_pca <- stats::prcomp(
  t(as.matrix(SummarizedExperiment::assay(pb, "logcounts"))),
  scale. = TRUE
)
SingleCellExperiment::reducedDim(pb, "pb_pca") <- pb_pca$x[, 1:2, drop = FALSE]
```

```{r eval = run_de_demo}
PlotReduction(pb, reduction = "pb_pca", colour_by = "condition")
PlotReduction(pb, reduction = "pb_pca", colour_by = "timepoint_factor")
```

### Pairwise differential expression

The first DE mode compares one group against another. A common default is to
compare each timepoint against baseline while adjusting for treatment as a
covariate.

```{r eval = run_de_demo}
pairwise_time <- run_edger_pairwise_de(
  pb,
  group_col = "timepoint",
  covariates = "condition",
  contrast_type = "reference",
  reference_level = "0",
  normalization = "TMM"
)

names(pairwise_time$tables)
```

```{r eval = run_de_demo}
pairwise_sig_counts <- aggregate(
  significant ~ contrast,
  data = pairwise_time$combined_table,
  FUN = sum
)

ggplot2::ggplot(pairwise_sig_counts, ggplot2::aes(x = contrast, y = significant, fill = contrast)) +
  ggplot2::geom_col(width = 0.72) +
  ggplot2::theme_classic() +
  ggplot2::theme(legend.position = "none") +
  ggplot2::labs(
    title = "Significant genes in timepoint-versus-baseline contrasts",
    x = "contrast",
    y = "number of DE genes"
  )
```

Inspect one contrast in more detail:

```{r eval = run_de_demo}
head(
  pairwise_time$tables[["24_vs_0"]][
    order(pairwise_time$tables[["24_vs_0"]]$FDR),
    c("gene", "logFC", "FDR")
  ],
  10
)

plot_pairwise_de_volcano(
  pairwise_time,
  contrast = "24_vs_0",
  fdr_cutoff = 0.05,
  lfc_cutoff = 0.5,
  top_n_labels = 8
)

plot_pairwise_de_ma(
  pairwise_time,
  contrast = "24_vs_0",
  fdr_cutoff = 0.05
)
```

`SCProkaR` also supports composite treatment-timepoint groups when you want an
explicit contrast such as `drug_A at 24h` versus `drug_A at 0h`.

```{r eval = run_de_demo}
pairwise_composite <- run_edger_pairwise_de(
  pb,
  group_col = "condition",
  combine_group_cols = c("condition", "timepoint"),
  contrast_type = "manual",
  manual_pairs = data.frame(
    group1 = c("drug_A__24", "drug_C__48", "drug_B__96"),
    group2 = c("drug_A__0", "drug_C__0", "drug_B__0"),
    stringsAsFactors = FALSE
  ),
  normalization = "TMM"
)

plot_pairwise_de_volcano(
  pairwise_composite,
  contrast = "drug_A__24_vs_drug_A__0",
  fdr_cutoff = 0.05,
  lfc_cutoff = 0.5,
  top_n_labels = 8
)
```

### Spline-based time-series differential expression

The second DE mode treats time as continuous and uses a spline basis to detect
genes that change over time within each condition.

```{r eval = run_de_demo}
spline_de <- run_edger_spline_de(
  pb,
  time_col = "timepoint",
  condition_col = "condition",
  df = 3,
  normalization = "TMM",
  return_curve_for = "significant"
)

names(spline_de$by_condition)
```

```{r eval = run_de_demo}
spline_sig_counts <- aggregate(
  significant ~ condition,
  data = spline_de$combined_table,
  FUN = sum
)

ggplot2::ggplot(spline_sig_counts, ggplot2::aes(x = condition, y = significant, fill = condition)) +
  ggplot2::geom_col(width = 0.72) +
  ggplot2::theme_classic() +
  ggplot2::theme(legend.position = "none") +
  ggplot2::labs(
    title = "Genes with significant temporal change within each condition",
    x = "condition",
    y = "number of spline DE genes"
  )
```

```{r eval = run_de_demo}
top_by_condition <- lapply(
  split(spline_de$combined_table, spline_de$combined_table$condition),
  function(df) {
    keep_cols <- intersect(c("gene", "F", "PValue", "FDR"), colnames(df))
    utils::head(df[order(df$FDR), keep_cols, drop = FALSE], 6)
  }
)

top_by_condition
```

Plot fitted curves for genes with strong time-dependent changes:

```{r eval = run_de_demo}
curve_genes <- unique(unlist(lapply(top_by_condition, function(df) df$gene[seq_len(min(2, nrow(df)))])))

plot_time_series_deg_curves(
  spline_de,
  genes = curve_genes,
  ncol = 2
)
```

For users who prefer a one-step interface, `run_scproka_de()` wraps the
pseudobulk and edgeR steps into one function call.

## Save the processed object

```{r eval = run_workflow}
saveRDS(sce, file = "scprokar_integrated_sce.rds")
```

## Notes

- `CreateBacObject()` accepts Seurat input directly, but the downstream analysis
  object remains a `SingleCellExperiment`.
- `MergeBacObjects()` is the recommended merge path for multiple samples in
  `SCProkaR`.
- `PlotReduction()` is a convenient visualization helper for both original and
  integrated reductions.
- `PlotIntegrationOverview()` provides the standard Seurat-style integration
  comparison panels.
- `RunIntegratedClustering()` provides the post-integration cluster labels for
  future downstream analyses.
- `RunIntegratedUMAP()` works on integrated embeddings as well as any custom
  latent space registered with `RegisterIntegrationEmbedding()`.
- Pseudobulk differential expression is covered in the same package workflow
  via `aggregate_pseudobulk()`, `run_edger_pairwise_de()`, and
  `run_edger_spline_de()`.
- Time-aware trajectory analysis is documented separately in the TATA vignette.
