---
title: "scProkaR core workflow for microbial single-cell RNA-seq"
author:
- name: Ning Liu
  affiliation: University of Adelaide
- name: Jinxin Zhao
  affiliation: Monash University
date: "`r Sys.Date()`"
package: scProkaR
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
    toc_depth: 2
abstract: |
  This vignette walks through the end-to-end scProkaR workflow on a bacterial
  single-cell RNA-seq time course of cells exposed to polymyxin B at two
  concentrations. It covers object construction, prokaryote-specific quality
  control, batch integration and benchmarking, clustering and visualisation,
  and pseudobulk differential expression with edgeR. Every analysis object is
  a SingleCellExperiment, so the workflow composes with the rest of
  Bioconductor.
vignette: >
  %\VignetteIndexEntry{1. scProkaR core workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    crop = NULL,
    fig.width = 6,
    fig.height = 4.5,
    dpi = 96,
    fig.retina = 1
)
```

# Introduction

`scProkaR` is a workflow package for **microbial** single-cell
RNA sequencing. Bacterial scRNA-seq differs from mammalian scRNA-seq in ways
that break the usual assumptions of general-purpose single-cell tooling:
libraries are one to two orders of magnitude sparser (a few hundred UMIs per
cell is normal rather than pathological), bacteria have no mitochondrial
genome so the standard `pct_mito` quality-control axis does not exist, the
dominant technical confounder is instead the fraction of reads captured on
ribosomal RNA, and gene identifiers are a mixture of locus tags and gene
symbols rather than stable Ensembl IDs. Most published bacterial scRNA-seq
experiments are also perturbation *time courses* -- a drug or stress applied at
several doses and sampled at several timepoints -- so the analytical question is
usually "how does the population reorganise over time under treatment", not
"which cell types are present".

scProkaR addresses those specifics while staying inside the Bioconductor
single-cell ecosystem rather than beside it. Concretely, the package provides:

* `CreateBacObject()`, a single entry point that accepts a count matrix, a
  10x Genomics CellRanger output directory, a `SingleCellExperiment`, or a
  Seurat object, and always returns a `SingleCellExperiment`;
* `RunBacQC()` and `FilterBacCells()`, which compute and apply
  prokaryote-appropriate QC metrics including rRNA and ribosomal-protein
  fractions;
* `IntegrateBacData()` and `RegisterIntegrationEmbedding()`, which produce or
  adopt a batch-corrected embedding;
* `BenchmarkIntegration()`, which scores competing embeddings with
  scIB-*inspired* metrics so the choice of correction is an evidence-based
  one;
* `aggregate_pseudobulk()`, `normalize_pseudobulk()`,
  `run_edger_pairwise_de()` and `run_edger_spline_de()`, a pseudobulk
  differential-expression path built on `r Biocpkg("edgeR")`;
* Time Aware Trajectory Analysis (TATA), covered in a separate vignette.

**Relationship to existing Bioconductor infrastructure.** Every object that
scProkaR creates, modifies, or returns is a
`r Biocpkg("SingleCellExperiment")`. Counts live in the `counts` assay,
log-normalised values in `logcounts`, cell annotation in `colData()`,
embeddings in `reducedDims()`, and provenance for each scProkaR step in
`metadata(sce)$scProkaR`. Nothing is hidden in a bespoke class. That means you
can hand a scProkaR object straight to `r Biocpkg("scater")` for plotting and
per-cell QC, to `r Biocpkg("scran")` for variance modelling, marker detection
or graph-based clustering, or to `r Biocpkg("batchelor")` for any correction
method the package does not wrap; and equally you can hand any
`SingleCellExperiment` produced elsewhere to scProkaR. scProkaR itself uses
`r Biocpkg("scran")` for feature selection when it is installed, delegates MNN
correction to `r Biocpkg("batchelor")`, and delegates the differential-testing
statistics to `r Biocpkg("edgeR")`. The package is a bacteria-aware layer over
that infrastructure, not a replacement for it.

# Installation

```{r install, eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("scProkaR")
```

```{r load-packages}
library(scProkaR)
library(SingleCellExperiment)
library(ggplot2)
```

# The example data

The package ships one example object, `sce1`, a droplet-based bacterial
scRNA-seq experiment in which cells were challenged with polymyxin B (PMB) at
two concentrations and sampled over time. See `inst/scripts/make-sce1.R` for the
provenance of this object.

```{r load-data}
data("sce1", package = "scProkaR")
sce1
```

The design is a dose-by-time grid:

```{r describe-design}
table(sce1$treatments, sce1$timepoints)
```

There are three treatment arms -- an untreated `control` plus two polymyxin B
concentrations, `PMB0.5` and `PMB2` -- sampled at 0, 1, 4 and 7 hours. The
untreated baseline was only collected at hour 0, so the `0` timepoint exists
in the control arm alone; the treated arms start at hour 1. Each of the ten
resulting samples contributes exactly 500 cells, for 5000 cells in total, and
`colData()` records the sample of origin, the treatment, the timepoint, and a
pre-computed biological label `clusters` that we will use later as ground
truth when benchmarking integration.

```{r describe-coldata}
head(as.data.frame(colData(sce1)), 4)
```

The 3722 features are a mix of gene symbols and locus tags, which is typical of
a bacterial annotation: well-characterised genes carry a symbol, everything
else keeps its locus tag.

```{r describe-features}
head(rownames(sce1), 8)
```

Counts are sparse even by single-cell standards, which is the single most
important practical fact about this data type:

```{r describe-depth}
summary(Matrix::colSums(assay(sce1, "counts")))
```

## Subsetting for this vignette

To keep the vignette fast to build we take a stratified random subsample of
250 cells from each of the ten samples (2500 of 5000 cells). The design stays
completely balanced, so every step below behaves exactly as it would on the
full object; only the running time and the statistical power change. On your
own data, do not subsample -- run the workflow on all cells.

```{r subset-cells}
set.seed(101)
keep_cells <- unlist(lapply(
    split(seq_len(ncol(sce1)), sce1$sample),
    function(idx) sample(idx, 250)
))
sce_small <- sce1[, sort(keep_cells)]
dim(sce_small)
```

# Creating a scProkaR object

`CreateBacObject()` is the entry point. Here we drive it from a raw count
matrix plus a per-cell metadata `data.frame`, which is the most general case;
the `sample_col`, `batch_col`, `condition_col` and `time_col` arguments simply
tell scProkaR which existing metadata columns carry those roles.

Setting `run_unintegrated = TRUE` also runs the standard uncorrected
first-pass: log-normalisation, feature selection, PCA, graph-based clustering,
and (when `r CRANpkg("uwot")` is installed) a UMAP stored as
`umap.unintegrated`. That uncorrected view is what we will compare the
integrated embeddings against.

```{r create-object}
counts_matrix <- assay(sce_small, "counts")
cell_meta <- as.data.frame(colData(sce_small))

set.seed(101)
sce <- CreateBacObject(
    counts_matrix,
    cell_metadata = cell_meta,
    sample_col = "sample",
    batch_col = "treatments",
    condition_col = "treatments",
    time_col = "timepoints",
    organism = "bacteria",
    run_unintegrated = TRUE,
    unintegrated_dims = seq_len(20),
    unintegrated_k = 20,
    unintegrated_resolution = 1
)
sce
```

The result is an ordinary `SingleCellExperiment`, with the analysis history
recorded in its metadata:

```{r inspect-object}
reducedDimNames(sce)
names(S4Vectors::metadata(sce)$scProkaR)
```

## Other entry points

Two further input types are accepted by the same function.

**10x Genomics directories.** Pass the path to a CellRanger output folder
(the one containing `matrix.mtx.gz`, `barcodes.tsv.gz` and `features.tsv.gz`,
whether or not it is nested inside `filtered_feature_bc_matrix/`) and
scProkaR reads it directly, using `r Biocpkg("DropletUtils")` when that
package is available and falling back to `r CRANpkg("Matrix")` otherwise:

```{r tenx-entry-point}
tenx_call <- quote(CreateBacObject(
    "path/to/filtered_feature_bc_matrix",
    sample_id_value = "PMB2_7",
    sample_col = "sample"
))
tenx_call
```

**Seurat objects.** If your data already lives in a Seurat object,
`CreateBacObject()` converts it, pulling the requested assay and layer,
carrying over `meta.data` and any existing reductions. This path needs
`r CRANpkg("SeuratObject")`, which is only a suggested dependency, so it is
guarded:

```{r seurat-entry-point}
if (requireNamespace("SeuratObject", quietly = TRUE)) {
    demo_cells <- seq_len(200)
    seu <- SeuratObject::CreateSeuratObject(
        counts = counts_matrix[, demo_cells],
        meta.data = cell_meta[demo_cells, , drop = FALSE]
    )
    sce_from_seurat <- CreateBacObject(
        seu,
        seurat_assay = "RNA",
        seurat_layer = "counts",
        sample_col = "sample",
        batch_col = "treatments"
    )
    print(dim(sce_from_seurat))
    print(head(colnames(colData(sce_from_seurat))))
} else {
    message("SeuratObject is not installed; skipping the Seurat entry point.")
}
```

The `Layer 'data' is empty` warning above is expected and harmless: the
conversion also looks for a normalised layer to carry over as `logcounts`, and
this demonstration object only has raw counts. scProkaR normalises for itself
when it needs to.

Whichever door you come in through, the object you get back is a
`SingleCellExperiment` and the rest of the workflow is identical. When you
build one object per sample, combine them with `MergeBacObjects()` rather than
`cbind()`: it aligns genes across objects, de-duplicates cell barcodes that
collide between samples (keeping the original in
`colData()$original_cell_id`), and carries compatible reductions forward.

# Quality control

`RunBacQC()` computes per-cell QC metrics and writes them into `colData()`.
Alongside the familiar `total_counts` and `detected_features`, it computes two
prokaryote-specific fractions:

* **`rrna_fraction`** -- the proportion of a cell's counts falling on ribosomal
  *RNA* genes. This is the bacterial analogue of the mitochondrial fraction in
  mammalian data. rRNA dominates the bacterial transcriptome, so most
  protocols try to deplete it; a cell with an unusually high residual rRNA
  fraction usually reflects failed depletion or a damaged, lysing cell whose
  mRNA has degraded away faster than its far more abundant rRNA.
* **`ribo_fraction`** -- the proportion of counts on ribosomal *protein* genes
  (`rpl*`, `rps*`). This is a biological signal rather than a defect: it
  tracks translational capacity and therefore growth rate, and it typically
  collapses under antibiotic stress.

Genes are classified by regular expression on the row names by default
(`rrna_pattern = "^(rrs|rrl|rrf)"`), which matches the standard bacterial
rRNA nomenclature. If your annotation carries an explicit gene-class column in
`rowData()`, pass its name as `gene_class_col` and it will be used instead.

```{r run-qc}
sce <- RunBacQC(sce)
head(as.data.frame(colData(sce))[, c(
    "total_counts",
    "detected_features",
    "rrna_fraction",
    "ribo_fraction"
)], 4)
```

An honest note about this particular dataset: `sce1` was built from an
annotation in which the rRNA loci had already been removed, so **no gene in
`sce1` matches the rRNA pattern and `rrna_fraction` is exactly zero for every
cell**.

```{r check-rrna}
sum(grepl("^(rrs|rrl|rrf)", rownames(sce), ignore.case = TRUE))
summary(sce$rrna_fraction)
```

The ribosomal-protein fraction, by contrast, is informative here and does vary
between arms:

```{r summarise-ribo}
round(tapply(sce$ribo_fraction, sce$treatments, median), 4)
```

Note that the ordering is not monotonic in dose: the high-dose `PMB2` arm has
the highest median ribosomal-protein fraction, above both the untreated
control and the low-dose arm. That is a reminder to treat `ribo_fraction` as a
covariate to inspect rather than a QC threshold to enforce -- it mixes
translational state, growth phase and stress response, and these medians pool
across all timepoints. Do not filter cells on it.

`FilterBacCells()` applies thresholds to those metrics. Because
`rrna_fraction` is identically zero above, the `max_rrna_fraction` cutoff is a
no-op on this dataset; we keep it in the call so the code is copy-pasteable
for data where rRNA *is* annotated, but the cells actually removed here are
removed on depth and complexity alone.

```{r filter-cells}
sce <- FilterBacCells(
    sce,
    min_counts = 50,
    min_features = 30,
    max_rrna_fraction = 0.2
)
dim(sce)
table(sce$treatments)
```

Choose thresholds by looking at your own distributions rather than copying
numbers. `min_counts = 50` is a reasonable floor for this protocol and would
be absurdly permissive for mammalian data.

# Integration across treatment groups

In this experiment the three treatment arms were processed as separate
batches, so `treatments` doubles as the batch variable. That makes it a
deliberately hard and honest test case: any correction strong enough to erase
the batch effect will also erase the drug effect. This is exactly the tension
that the benchmarking section below is designed to expose, and it is why you
should never accept an integration on faith.

## MNN correction

`IntegrateBacData(method = "mnn")` delegates to the mutual-nearest-neighbours
correction in `r Biocpkg("batchelor")`. That package is suggested rather than
required, so the call is guarded:

```{r integrate-mnn}
if (requireNamespace("batchelor", quietly = TRUE)) {
    set.seed(101)
    sce <- IntegrateBacData(
        sce,
        batch_col = "treatments",
        method = "mnn",
        dims = seq_len(20)
    )
    print(reducedDimNames(sce))
} else {
    message("batchelor is not installed; skipping MNN integration.")
}
```

## Harmony correction

`method = "harmony"` uses `r CRANpkg("harmony")`, likewise suggested and
likewise guarded:

```{r integrate-harmony}
if (requireNamespace("harmony", quietly = TRUE)) {
    set.seed(101)
    sce <- IntegrateBacData(
        sce,
        batch_col = "treatments",
        method = "harmony",
        dims = seq_len(20)
    )
    print(reducedDimNames(sce))
} else {
    message("harmony is not installed; skipping Harmony integration.")
}
```

Each backend writes its embedding to `reducedDims()` under
`integrated_<method>` and records the call in `metadata(sce)$scProkaR`.

## Registering your own embedding

`RegisterIntegrationEmbedding()` is the always-available path: it takes any
cell-by-dimension matrix you computed yourself -- from a method scProkaR does
not wrap, from a variational autoencoder, from a collaborator -- and registers
it so that every downstream scProkaR function treats it like a native
integration result. It depends on nothing beyond the package's own imports.

Here we register a deliberately simple correction: per-batch centring of the
PCA embedding, which removes the mean shift between treatment arms and nothing
else. It is a useful baseline precisely because it is crude -- if a
sophisticated method cannot beat mean-centring on your benchmark, that is
worth knowing.

```{r register-embedding}
pca_emb <- reducedDim(sce, "PCA")[, seq_len(20), drop = FALSE]
batch_vec <- as.character(sce$treatments)
batch_centres <- t(vapply(
    split(seq_len(nrow(pca_emb)), batch_vec),
    function(idx) colMeans(pca_emb[idx, , drop = FALSE]),
    numeric(ncol(pca_emb))
))
centred_emb <- pca_emb - batch_centres[batch_vec, , drop = FALSE]

sce <- RegisterIntegrationEmbedding(
    sce,
    embedding = centred_emb,
    method_name = "centred"
)
reducedDimNames(sce)
```

For an object that already contains external latent spaces, use
`RegisterExistingIntegrationEmbeddings(sce, reductions = "X_scVI")`, or pass
several names such as `c("seurat_cca", "X_scVI")`. The default copies them to
`integrated_cca` and `integrated_scvi`; `copy = FALSE` registers the original
reductions directly. Omitting `reductions` detects likely integration names.
This registers existing coordinates; it does not run the external methods.

Both registration functions protect existing embeddings from replacement
unless `overwrite = TRUE` is supplied. Registering the same values again is
allowed. `BenchmarkIntegration()` includes registered methods and other
`integrated_*` reductions by default, and accepts names without regard to case
(for example, `methods = "SCVI"`).

We now pick the reduction to carry forward. MNN is preferred when
`r Biocpkg("batchelor")` was available; otherwise the registered embedding
takes over, so the remainder of the vignette produces meaningful output on a
minimal installation.

```{r choose-reduction}
integrated_names <- grep(
    "^integrated_",
    reducedDimNames(sce),
    value = TRUE
)
main_reduction <- if ("integrated_mnn" %in% integrated_names) {
    "integrated_mnn"
} else {
    "integrated_centred"
}
main_reduction
```

# UMAP and clustering on the integrated space

`RunIntegratedUMAP()` runs a UMAP on any stored embedding and saves it as
`umap_<reduction>`. It requires `r CRANpkg("uwot")`, so we guard it and fall
back to plotting the first two dimensions of the integrated embedding itself
when `uwot` is unavailable.

```{r run-umap}
if (requireNamespace("uwot", quietly = TRUE)) {
    set.seed(101)
    sce <- RunIntegratedUMAP(
        sce,
        reduction = main_reduction,
        n_neighbors = 20
    )
    vis_reduction <- paste0("umap_", main_reduction)
} else {
    message("uwot is not installed; plotting the embedding directly.")
    vis_reduction <- main_reduction
}
vis_reduction
```

The uncorrected counterpart for side-by-side comparison is chosen the same
way:

```{r choose-unintegrated}
unintegrated_reduction <- if (
    "umap.unintegrated" %in% reducedDimNames(sce)
) {
    "umap.unintegrated"
} else {
    "PCA"
}
unintegrated_reduction
```

`RunIntegratedClustering()` builds a shared-nearest-neighbour graph on the
corrected embedding and partitions it. `k` sets the size of the neighbourhood
graph and `resolution` how finely that graph is cut; downstream analyses
should always use clusters derived from the *integrated* space rather than the
pre-integration ones.

```{r run-clustering}
set.seed(101)
sce <- RunIntegratedClustering(
    sce,
    reduction = main_reduction,
    cluster_col = "integrated_clusters",
    k = 20,
    algorithm = "louvain",
    resolution = 1
)
table(sce$integrated_clusters)
```

Cross-tabulating against the supplied biological labels shows how much of the
known structure the integrated clustering recovers. A clean block-diagonal
pattern means the two agree; substantial off-diagonal mass means the graph
partition is splitting or merging the reference groups.

```{r compare-clusters}
table(sce$integrated_clusters, sce$clusters)
```

# Visualisation

`PlotReduction()` draws any stored embedding, coloured by any `colData()`
column, and returns a plain `ggplot` object you can modify further.

```{r plot-batch}
PlotReduction(
    sce,
    reduction = vis_reduction,
    colour_by = "treatments",
    point_size = 0.5
)
```

```{r plot-labels}
PlotReduction(
    sce,
    reduction = vis_reduction,
    colour_by = "clusters",
    point_size = 0.5
)
```

Time can be overlaid the same way. Coercing the timepoint to a factor with an
explicit level order keeps the legend in chronological rather than
lexicographic order:

```{r plot-time}
sce$timepoint_factor <- factor(
    sce$timepoints,
    levels = c("0", "1", "4", "7")
)
PlotReduction(
    sce,
    reduction = vis_reduction,
    colour_by = "timepoint_factor",
    point_size = 0.5
)
```

`facet_by` splits one embedding into panels while holding the coordinates
fixed, which is the right way to ask whether a cluster is present in every
arm or driven by one:

```{r plot-facet}
PlotReduction(
    sce,
    reduction = vis_reduction,
    colour_by = "integrated_clusters",
    facet_by = "treatments",
    point_size = 0.5
)
```

`PlotIntegrationOverview()` assembles the standard before/after panel set in
one call and returns them as a named list.

```{r integration-overview}
overview <- PlotIntegrationOverview(
    sce,
    unintegrated_reduction = unintegrated_reduction,
    integrated_reduction = vis_reduction,
    batch_col = "treatments",
    label_col = "clusters",
    split_by = "treatments",
    point_size = 0.5
)
names(overview)
```

```{r overview-before-batch}
overview$unintegrated_by_batch
```

```{r overview-after-batch}
overview$integrated_by_batch
```

```{r overview-after-label}
overview$integrated_by_label
```

How to read these. `unintegrated_by_batch` should show more batch separation
than `integrated_by_batch` -- that is the correction doing its job.
`integrated_by_label` should still separate the biological labels -- that is
the correction *not* overdoing its job. `integrated_split` (the equivalent of
splitting a dimension plot by condition) shows whether each arm populates the
same regions. Remember the caveat from the previous section: here batch and
treatment are confounded by design, so perfect batch mixing would be a red
flag, not a success.

# Benchmarking the integration

Eyeballing embeddings is not enough, and it is easy to talk yourself into
liking whichever picture you drew last. `BenchmarkIntegration()` scores every
candidate embedding on two families of metrics and ranks them.

**These are scIB-inspired reimplementations, not the reference scIB
implementations.** The metric *names* follow Luecken et al. (2022) and the
intent is the same, but the code here is independent, is tuned to run quickly
on sparse bacterial data, and will not reproduce reference scIB numbers
exactly. Use the scores to compare methods *within* one run; do not quote them
as scIB values or compare them across studies.

```{r run-benchmark}
bench_methods <- c("PCA", integrated_names)
set.seed(101)
bench <- BenchmarkIntegration(
    sce,
    batch_col = "treatments",
    label_col = "clusters",
    methods = bench_methods
)
bench$ranking
```

The two families are:

* **Batch removal** -- `pcr_batch` (how much variance in the principal
  components is still explained by batch), `ilisi` (local batch mixing),
  `kbet_like` (whether each neighbourhood's batch composition matches the
  global one), and `graph_connectivity` (whether cells of the same label stay
  connected in the kNN graph).
* **Biological conservation** -- `clisi` (local label purity), `ari` and `nmi`
  (agreement between a graph clustering of the embedding and the reference
  labels), and `silhouette` (label separation in the embedding).

All metrics are rescaled so that higher is better, and `ranking` is the mean
across metrics.

```{r benchmark-scores}
head(bench$scores, 8)
```

```{r benchmark-overall}
bench$plots$overall
```

The trade-off plot is usually the most informative single figure: batch
removal on the x-axis, biological conservation on the y-axis. Methods towards
the top right are doing well on both; a method far to the right but low down
has mixed the batches by destroying biology.

```{r benchmark-tradeoff}
bench$plots$tradeoff
```

```{r benchmark-key-metrics}
bench$plots$key_metrics
```

The remaining panels -- `bench$plots$batch_removal`,
`bench$plots$bio_conservation` and `bench$plots$heatmap` -- break the same
scores down by category and by individual metric.

Do not read the overall ranking as a verdict. A single mean over metrics that
measure opposing goals is a summary, not a decision; pick the embedding whose
*position on the trade-off* matches what you need from the analysis. In a
confounded design like this one, an embedding that scores modestly on batch
removal but preserves the drug-response structure is the right choice.

# Pseudobulk differential expression

Per-cell differential testing is poorly calibrated when cells from the same
sample are treated as independent replicates. scProkaR therefore takes the
pseudobulk route: sum counts within each biological replicate, then test the
resulting sample-level matrix with `r Biocpkg("edgeR")`, which is a
well-understood negative-binomial framework for exactly this kind of data.

## Defining replicates

This dataset has one library per treatment-by-timepoint combination and no
true biological replication, so we split the cells of each sample into three
pseudo-replicates in order to demonstrate the mechanics. **This is a teaching
device, not a statistical recommendation.** Pseudo-replicates share the
technical and biological idiosyncrasies of their parent library, so they
understate the true variance and inflate significance. In a real study,
aggregate over genuine independent replicates -- separate cultures, separate
animals, separate donors.

```{r assign-replicates}
set.seed(101)
replicate_id <- integer(ncol(sce))
for (idx in split(seq_len(ncol(sce)), sce$sample)) {
    replicate_id[idx] <- sample(rep(seq_len(3), length.out = length(idx)))
}
sce$replicate <- factor(replicate_id)
sce$pb_id <- paste(sce$sample, sce$replicate, sep = "_r")
table(sce$sample, sce$replicate)
```

## Aggregating and normalising

`aggregate_pseudobulk()` sums counts over every unique combination of
`sample_cols` and carries the corresponding sample-level annotation into the
`colData()` of the result, which is itself a `SingleCellExperiment`.
`min_cells` drops aggregates built from too few cells to be trustworthy.

```{r aggregate-pseudobulk}
pb <- aggregate_pseudobulk(
    sce,
    sample_cols = c(
        "pb_id", "sample", "replicate", "treatments", "timepoints"
    ),
    aggregation = "sum",
    min_cells = 10
)
dim(pb)
head(as.data.frame(colData(pb))[, c(
    "pb_id", "treatments", "timepoints", "ncells"
)], 4)
```

`normalize_pseudobulk()` computes edgeR normalisation factors (TMM by
default), stores them and the library sizes in `colData()`, and adds a
`logcounts` assay of log-CPM values for visualisation. The statistical tests
below re-normalise internally from raw counts, so this step is for inspection
and plotting rather than a prerequisite.

```{r normalize-pseudobulk}
pb <- normalize_pseudobulk(pb, method = "TMM")
summary(pb$norm.factors)
```

A quick PCA of the pseudobulk profiles is the cheapest sanity check available:
replicates of a sample should sit together, and the dominant axis should
correspond to something you designed.

```{r pseudobulk-pca}
pb_pca <- stats::prcomp(
    t(as.matrix(assay(pb, "logcounts"))),
    scale. = TRUE
)
reducedDim(pb, "pb_pca") <- pb_pca$x[, seq_len(4), drop = FALSE]
pb$timepoint_factor <- factor(pb$timepoints, levels = c("0", "1", "4", "7"))
PlotReduction(pb, reduction = "pb_pca", colour_by = "treatments",
    point_size = 3)
```

```{r pseudobulk-pca-time}
PlotReduction(pb, reduction = "pb_pca", colour_by = "timepoint_factor",
    point_size = 3)
```

## Pairwise contrasts

`run_edger_pairwise_de()` fits a quasi-likelihood negative-binomial model and
extracts contrasts. Below, each timepoint is compared with the hour-0
baseline, with treatment included as a covariate so that arm-specific
offsets are absorbed by the model.

Interpret these contrasts carefully: hour 0 exists only in the control arm, so
`4_vs_0` describes overall movement away from the untreated baseline rather
than a perfectly balanced treatment-adjusted comparison.

```{r pairwise-de}
pairwise_time <- run_edger_pairwise_de(
    pb,
    group_col = "timepoints",
    covariates = "treatments",
    contrast_type = "reference",
    reference_level = "0",
    normalization = "TMM",
    fdr_cutoff = 0.05
)
names(pairwise_time$tables)
```

`$tables` holds one `data.frame` per contrast and `$combined_table` stacks
them with a `contrast` column, which makes summarising across contrasts easy.
Note that edgeR's `filterByExpr()` step removes genes with insufficient counts
across the design before testing, so the tables contain fewer than the 3722
input genes -- a larger effect here than it would be on the full dataset,
because we subsampled the cells.

```{r pairwise-summary}
table(
    pairwise_time$combined_table$contrast,
    pairwise_time$combined_table$significant
)
```

```{r pairwise-top-genes}
tab_4_vs_0 <- pairwise_time$tables[["4_vs_0"]]
head(tab_4_vs_0[order(tab_4_vs_0$FDR), c("gene", "logFC", "FDR")], 10)
```

A volcano plot shows effect size against evidence. Points beyond the vertical
guides and above the horizontal one pass both the fold-change and the FDR
threshold; the labelled genes are the strongest hits.

```{r pairwise-volcano}
plot_pairwise_de_volcano(
    pairwise_time,
    contrast = "4_vs_0",
    fdr_cutoff = 0.05,
    lfc_cutoff = 1,
    top_n_labels = 8
)
```

The companion MA plot puts log-fold-change against average abundance, which is
the better view for spotting fold changes that are driven purely by
low-expression noise:

```{r pairwise-ma}
plot_pairwise_de_ma(
    pairwise_time,
    contrast = "4_vs_0",
    fdr_cutoff = 0.05
)
```

When you want an explicit comparison between composite groups --
"PMB2 at 7 hours versus control at 7 hours" -- combine the grouping columns and
supply the pairs directly:

```{r pairwise-composite}
pairwise_dose <- run_edger_pairwise_de(
    pb,
    group_col = "treatments",
    combine_group_cols = c("treatments", "timepoints"),
    contrast_type = "manual",
    manual_pairs = data.frame(
        group1 = c("PMB2__7", "PMB0.5__7"),
        group2 = c("control__7", "control__7"),
        stringsAsFactors = FALSE
    ),
    normalization = "TMM"
)
names(pairwise_dose$tables)
```

```{r pairwise-composite-volcano}
plot_pairwise_de_volcano(
    pairwise_dose,
    contrast = "PMB2__7_vs_control__7",
    fdr_cutoff = 0.05,
    lfc_cutoff = 1,
    top_n_labels = 8
)
```

## Time-course testing with splines

Pairwise contrasts discretise time. `run_edger_spline_de()` instead treats
time as continuous, fits a natural cubic spline basis, and tests all spline
coefficients jointly -- so a gene is called significant if its expression
changes with time in *any* smooth way, including non-monotonic responses that
a pair of endpoints would miss. With `condition_col` set, a separate fit is
returned for each treatment arm.

Choose `df` against the number of distinct timepoints you actually have. The
treated arms here have only three (1, 4 and 7 hours), so `df = 2` is the
sensible ceiling; `df = 3` would leave almost no residual degrees of freedom.

```{r spline-de}
spline_de <- run_edger_spline_de(
    pb,
    time_col = "timepoints",
    condition_col = "treatments",
    df = 2,
    normalization = "TMM",
    return_curve_for = "significant",
    fdr_cutoff = 0.05
)
names(spline_de$by_condition)
```

```{r spline-summary}
table(
    spline_de$combined_table$treatments,
    spline_de$combined_table$significant
)
```

More genes changing over time under polymyxin B than in the untreated control
is what you would hope to see: the control arm captures baseline drift such as
growth-phase transitions, and the treated arms add the drug response on top.

```{r spline-top-genes}
spline_top <- spline_de$combined_table[
    order(spline_de$combined_table$FDR),
    c("gene", "treatments", "FDR")
]
head(spline_top, 10)
```

`plot_time_series_deg_curves()` draws the fitted trajectories for chosen genes
so the shape of each response is visible, not just its p-value:

```{r spline-curves}
curve_genes <- head(unique(spline_top$gene), 6)
plot_time_series_deg_curves(spline_de, genes = curve_genes, ncol = 3)
```

If you prefer a single call, `run_scproka_de()` wraps aggregation,
normalisation and testing into one function.

# Where to go next

The trajectory side of the package -- Time Aware Trajectory Analysis, a
time-informed trajectory abstraction for perturbation time courses -- is
covered in the companion vignette, `vignette("tata-workflow", package =
"scProkaR")`.

Because the object is a plain `SingleCellExperiment` throughout, you can also
continue in the wider ecosystem: `r Biocpkg("scater")` for further plotting,
`r Biocpkg("scran")` for marker detection, or any Bioconductor package that
accepts a `SingleCellExperiment`.

# References and further reading

* Amezquita RA, Lun ATL, Becht E, *et al.* (2020). Orchestrating single-cell
  analysis with Bioconductor. *Nature Methods* 17, 137-145.
  <https://bioconductor.org/books/release/OSCA/>
* Robinson MD, McCarthy DJ, Smyth GK (2010). edgeR: a Bioconductor package for
  differential expression analysis of digital gene expression data.
  *Bioinformatics* 26, 139-140.
  <https://bioconductor.org/packages/edgeR>
* Chen Y, Lun ATL, Smyth GK (2016). From reads to genes to pathways:
  differential expression analysis of RNA-Seq experiments using Rsubread and
  the edgeR quasi-likelihood pipeline. *F1000Research* 5, 1438.
* Haghverdi L, Lun ATL, Morgan MD, Marioni JC (2018). Batch effects in
  single-cell RNA-sequencing data are corrected by matching mutual nearest
  neighbors. *Nature Biotechnology* 36, 421-427.
  <https://bioconductor.org/packages/batchelor>
* Korsunsky I, Millard N, Fan J, *et al.* (2019). Fast, sensitive and accurate
  integration of single-cell data with Harmony. *Nature Methods* 16,
  1289-1296. <https://CRAN.R-project.org/package=harmony>
* Luecken MD, Buttner M, Chaichoompu K, *et al.* (2022). Benchmarking
  atlas-level data integration in single-cell genomics. *Nature Methods* 19,
  41-50. The metrics in `BenchmarkIntegration()` are inspired by this work but
  are independent reimplementations.
* McInnes L, Healy J, Melville J (2018). UMAP: Uniform Manifold Approximation
  and Projection for dimension reduction. *arXiv:1802.03426*.
  <https://CRAN.R-project.org/package=uwot>
* Lun ATL, McCarthy DJ, Marioni JC (2016). A step-by-step workflow for
  low-level analysis of single-cell RNA-seq data with Bioconductor.
  *F1000Research* 5, 2122. <https://bioconductor.org/packages/scran>
* Imdahl F, Vafadarnejad E, Homberger C, Saliba AE, Vogel J (2020).
  Single-cell RNA-sequencing reports growth-condition-specific global
  transcriptomes of individual bacteria. *Nature Microbiology* 5, 1202-1206.

# Session information

```{r session-info}
sessionInfo()
```
