---
title: "SCProkaR Two-Sample Workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{SCProkaR Two-Sample Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 5
)

if (file.exists("../DESCRIPTION")) {
  if (requireNamespace("devtools", quietly = TRUE)) {
    devtools::load_all("..")
  } else {
    library(SCProkaR)
  }
} else {
  library(SCProkaR)
}

run_workflow <- TRUE
```

## Overview

This vignette shows a practical `SCProkaR` workflow for two microbial
single-cell RNA-seq samples stored as Seurat `.rds` objects. The workflow:

- loads each sample directly with `CreateBacObject()`
- computes QC metrics including rRNA fraction
- filters low-quality cells
- merges samples safely with `MergeBacObjects()`
- performs batch integration
- visualizes unintegrated and integrated structure with Seurat-style panels
- optionally benchmarks the integration

`SCProkaR` keeps the main analysis object as a
`SingleCellExperiment`, while still accepting Seurat input for convenient
conversion.

## Load packages

```{r}
library(SummarizedExperiment)
library(SingleCellExperiment)
```

## Load Seurat `.rds` objects

Replace the file paths below with your own sample files.

```{r eval = run_workflow}
seu1 <- readRDS("../inst/extdata/sample1.rds")
seu2 <- readRDS("../inst/extdata/sample2.rds")
```

For clarity, add sample origin columns if they do not already exist:

```{r eval = run_workflow}
seu1$sample_id <- "sample1"
seu1$batch <- "sample1"

seu2$sample_id <- "sample2"
seu2$batch <- "sample2"
```

## Create SCProkaR objects

`CreateBacObject()` can extract counts, metadata, and existing reductions such
as `pca` and `umap` directly from each Seurat object.

```{r eval = run_workflow}
sce1 <- CreateBacObject(
  seu1,
  seurat_assay = "RNA",
  seurat_layer = "counts",
  sample_col = "sample_id",
  batch_col = "batch"
)

sce2 <- CreateBacObject(
  seu2,
  seurat_assay = "RNA",
  seurat_layer = "counts",
  sample_col = "sample_id",
  batch_col = "batch"
)
```

## Quality control

Compute QC metrics on each sample separately:

```{r eval = run_workflow}
sce1 <- RunBacQC(sce1)
sce2 <- RunBacQC(sce2)
```

The QC results are stored in `colData(sce)`:

```{r eval = run_workflow}
head(as.data.frame(colData(sce1))[, c(
  "total_counts",
  "detected_features",
  "rrna_fraction",
  "pct_rrna"
)])
```

Filter cells using your chosen thresholds:

```{r eval = run_workflow}
sce1 <- FilterBacCells(
  sce1,
  min_counts = 50,
  min_features = 50,
  max_rrna_fraction = 0.20
)

sce2 <- FilterBacCells(
  sce2,
  min_counts = 50,
  min_features = 50,
  max_rrna_fraction = 0.20
)
```

## Merge samples

Use `MergeBacObjects()` instead of `cbind()` to avoid duplicate-barcode issues
and to align genes safely across samples.

```{r eval = run_workflow}
sce <- MergeBacObjects(sce1, sce2, gene_mode = "intersect")
```

If cell barcodes overlap between samples, `MergeBacObjects()` automatically
makes them unique and stores the original barcode in
`colData(sce)$original_cell_id`.

If both input objects contain compatible reductions such as `pca` and `umap`,
`MergeBacObjects()` carries them forward into the merged object.

## Visualize unintegrated structure

If your Seurat objects already contained a UMAP, it is typically available after
conversion:

```{r eval = run_workflow}
PlotReduction(sce, reduction = "umap", colour_by = "batch")
PlotReduction(sce, reduction = "umap", colour_by = "seurat_clusters")
```

Interpretation:

- colouring by `batch` shows whether samples are separated before integration
- colouring by `seurat_clusters` or another biological label shows whether
  expected structure is present

`PlotReduction(..., facet_by = "batch")` is the `SCProkaR` equivalent of
Seurat's `split.by` visualization.

## Run integration

### Option 1: MNN

```{r eval = run_workflow}
sce <- IntegrateBacData(
  sce,
  batch_col = "batch",
  method = "mnn",
  dims = 1:30
)
```

### Option 2: Harmony

```{r eval = run_workflow}
sce <- IntegrateBacData(
  sce,
  batch_col = "batch",
  method = "harmony",
  dims = 1:30
)
```

The output embedding is stored as `integrated_mnn` or `integrated_harmony`.

## Visualize the integrated latent space

```{r eval = run_workflow}
PlotReduction(sce, reduction = "integrated_mnn", colour_by = "batch")
PlotReduction(sce, reduction = "integrated_mnn", colour_by = "seurat_clusters")
PlotReduction(sce, reduction = "integrated_harmony", colour_by = "batch")
PlotReduction(sce, reduction = "integrated_harmony", colour_by = "seurat_clusters")
```

This gives a quick direct view of the integrated latent space before running a
new UMAP.

## Run UMAP on the integrated embedding

`RunIntegratedUMAP()` computes a UMAP from any stored embedding:

```{r eval = run_workflow}
sce <- RunIntegratedUMAP(sce, reduction = "integrated_mnn")
sce <- RunIntegratedUMAP(sce, reduction = "integrated_harmony")
```

This stores a new reduced dimension named `umap_integrated_mnn`.

```{r eval = run_workflow}
PlotReduction(sce, reduction = "umap_integrated_mnn", colour_by = "batch")
PlotReduction(sce, reduction = "umap_integrated_mnn", colour_by = "seurat_clusters")
PlotReduction(sce, reduction = "umap_integrated_harmony", colour_by = "batch")
PlotReduction(sce, reduction = "umap_integrated_harmony", colour_by = "seurat_clusters")
```

## Cluster cells after integration

Future downstream analyses should use clusters identified from the integrated
embedding rather than the pre-integration batch structure.

`RunIntegratedClustering()` builds a kNN graph on the corrected embedding and
stores the resulting labels in `colData(sce)`.

```{r eval = run_workflow}
sce <- RunIntegratedClustering(
  sce,
  reduction = "integrated_mnn",
  cluster_col = "integrated_clusters",
  k = 20,
  resolution = 1
)
```

Here, `k` controls the neighbour graph size, while `resolution` controls how
finely that graph is partitioned into clusters. This mirrors the Seurat logic
of `FindNeighbors(k.param = ...)` followed by `FindClusters(resolution = ...)`.

Visualize the integrated clusters on the corrected UMAP:

```{r eval = run_workflow}
PlotReduction(sce, reduction = "umap_integrated_mnn", colour_by = "integrated_clusters")
PlotReduction(
  sce,
  reduction = "umap_integrated_mnn",
  colour_by = "integrated_clusters",
  facet_by = "batch"
)
```

## Seurat-style integration summary panels

The Seurat integration tutorial emphasizes three visual checks after alignment:

- compare unintegrated and integrated embeddings coloured by batch
- compare unintegrated and integrated embeddings coloured by cluster or label
- split the integrated embedding by batch/condition while keeping cluster
  colours fixed

`SCProkaR` provides `PlotIntegrationOverview()` for this pattern.

### MNN example

```{r eval = run_workflow}
mnn_plots <- PlotIntegrationOverview(
  sce,
  unintegrated_reduction = "umap",
  integrated_reduction = "umap_integrated_mnn",
  batch_col = "batch",
  label_col = "integrated_clusters",
  split_by = "batch"
)

mnn_plots$unintegrated_by_batch
mnn_plots$integrated_by_batch
mnn_plots$unintegrated_by_label
mnn_plots$integrated_by_label
mnn_plots$integrated_split
```

### Harmony example

```{r eval = run_workflow}
harmony_plots <- PlotIntegrationOverview(
  sce,
  unintegrated_reduction = "umap",
  integrated_reduction = "umap_integrated_harmony",
  batch_col = "batch",
  label_col = "integrated_clusters",
  split_by = "batch"
)

harmony_plots$unintegrated_by_batch
harmony_plots$integrated_by_batch
harmony_plots$unintegrated_by_label
harmony_plots$integrated_by_label
harmony_plots$integrated_split
```

Interpretation:

- `unintegrated_by_batch` should show stronger sample separation than
  `integrated_by_batch`
- `integrated_by_label` should preserve biologically meaningful group
  separation
- `integrated_split` is the closest equivalent to Seurat's
  `DimPlot(..., split.by = "stim")`

After successful integration:

- batches should mix better than in the original UMAP
- biological clusters should remain interpretable

## Benchmark integration quality

If you have a biological label such as `seurat_clusters`, you can benchmark the
integration with the package's scIB-inspired metric panel:

```{r eval = run_workflow}
bench <- BenchmarkIntegration(
  sce,
  batch_col = "batch",
  label_col = "seurat_clusters",
  methods = NULL
)

bench$ranking
bench$plots$overall
bench$plots$key_metrics
bench$plots$batch_removal
bench$plots$bio_conservation
bench$plots$tradeoff
bench$plots$heatmap
```

The key comparison figures are:

- `key_metrics`: per-metric score comparison across methods
- `batch_removal`: average of the batch-removal metrics
- `bio_conservation`: average of the biological-conservation metrics
- `tradeoff`: batch-removal versus bio-conservation, which is often the most
  useful summary of integration quality

## Save the processed object

```{r eval = run_workflow}
saveRDS(sce, file = "scprokar_integrated_sce.rds")
```

## Notes

- `CreateBacObject()` accepts Seurat input directly, but the downstream analysis
  object remains a `SingleCellExperiment`.
- `MergeBacObjects()` is the recommended merge path for multiple samples in
  `SCProkaR`.
- `PlotReduction()` is a convenient visualization helper for both original and
  integrated reductions.
- `PlotIntegrationOverview()` provides the standard Seurat-style integration
  comparison panels.
- `RunIntegratedClustering()` provides the post-integration cluster labels for
  future downstream analyses.
- `RunIntegratedUMAP()` works on integrated embeddings as well as any custom
  latent space registered with `RegisterIntegrationEmbedding()`.
