---
title: "SCProkaR TATA Workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{SCProkaR TATA Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 5
)

if (file.exists("../DESCRIPTION")) {
  if (requireNamespace("devtools", quietly = TRUE)) {
    devtools::load_all("..")
  } else {
    library(SCProkaR)
  }
} else {
  library(SCProkaR)
}

run_tata_demo <- TRUE
```

## Overview

This vignette demonstrates Time Aware Trajectory Analysis (TATA), a
PAGA-inspired trajectory abstraction that uses real experimental time labels to
score and direct cluster-level transitions.

The workflow below:

- simulates a center-rooted multi-drug time course
- visualizes the stored PCA and UMAP representations
- builds the cell-level kNN graph and graph-based clusters
- computes topology and time-consistency scores between clusters
- constructs the abstract TATA graph
- runs the one-step `run_tata()` wrapper and overlays the inferred trajectories

The simulation is intentionally awkward for topology-only methods because the
earliest cells are near the middle of the embedding rather than at a clear tip.

## Load packages

```{r}
library(SingleCellExperiment)
library(SummarizedExperiment)
```

## Simulate a multi-drug trajectory

```{r eval = run_tata_demo}
set.seed(101)

sce <- simulate_tata_multidrug_sce(
  n_cells = 2500,
  n_features = 100,
  n_pcs = 20,
  seed = 101
)

sce$timepoint_factor <- factor(
  sce$timepoint,
  levels = sort(unique(sce$timepoint))
)
```

Inspect the simulation design:

```{r eval = run_tata_demo}
with(as.data.frame(SummarizedExperiment::colData(sce)), table(condition, timepoint))
```

Visualize the stored embedding:

```{r eval = run_tata_demo}
PlotReduction(sce, reduction = "UMAP", colour_by = "timepoint_factor")
PlotReduction(sce, reduction = "UMAP", colour_by = "condition")
PlotReduction(sce, reduction = "UMAP", colour_by = "simulated_state")
```

## Build the cell-cell graph

TATA starts from a low-dimensional representation. Here we use the stored PCA
coordinates and connect each cell to its nearest neighbours.

```{r eval = run_tata_demo}
knn <- build_knn_graph(
  x = sce,
  dimred = "PCA",
  k = 30
)

c(
  n_cells = igraph::vcount(knn$graph),
  n_edges = igraph::ecount(knn$graph),
  k = knn$k
)
```

## Cluster cells into coarse states

The cell graph is clustered into coarse states that serve as nodes in the
abstract TATA graph.

```{r eval = run_tata_demo}
sce$tata_cluster_step <- factor(
  cluster_graph_states(
    cell_graph = knn$graph,
    method = "louvain",
    seed = 101
  )
)

table(sce$tata_cluster_step)
PlotReduction(sce, reduction = "UMAP", colour_by = "tata_cluster_step")
```

## Compute topology and time scores

TATA combines two ingredients for each cluster pair:

- a topology score based on how strongly the two clusters are connected in the
  cell-cell graph
- a time-consistency score based on whether those inter-cluster edges follow
  the observed experimental time order

```{r eval = run_tata_demo}
topology_table <- compute_topology_weights(
  adjacency = knn$adjacency,
  clusters = sce$tata_cluster_step
)

time_table <- compute_time_weights(
  adjacency = knn$adjacency,
  clusters = sce$tata_cluster_step,
  timepoint = sce$timepoint
)

head(
  topology_table[
    order(-topology_table$topology_weight),
    c("cluster_a", "cluster_b", "observed_edges", "expected_edges", "topology_weight")
  ],
  8
)

head(
  time_table[
    order(-time_table$time_consistency),
    c("cluster_a", "cluster_b", "n_time_edges", "flow_score", "time_weight", "time_consistency")
  ],
  8
)
```

```{r eval = run_tata_demo}
edge_metric_table <- merge(
  topology_table,
  time_table,
  by = c("cluster_a", "cluster_b"),
  all.x = TRUE,
  sort = FALSE
)

ggplot2::ggplot(
  edge_metric_table,
  ggplot2::aes(
    x = topology_weight,
    y = flow_score,
    size = observed_edges,
    colour = time_consistency
  )
) +
  ggplot2::geom_hline(yintercept = c(-0.10, 0.10), linetype = 2, colour = "grey60") +
  ggplot2::geom_point(alpha = 0.8) +
  ggplot2::scale_colour_viridis_c(option = "plasma") +
  ggplot2::theme_classic() +
  ggplot2::labs(
    title = "Topology and time consistency for cluster pairs",
    x = "topology weight",
    y = "temporal flow score",
    colour = "time consistency",
    size = "observed edges"
  )
```

## Build the abstract TATA graph

```{r eval = run_tata_demo}
tata_manual <- build_tata_graph(
  topology_table = topology_table,
  time_table = time_table,
  clusters = sce$tata_cluster_step,
  embedding = SingleCellExperiment::reducedDim(sce, "PCA"),
  timepoint = sce$timepoint,
  alpha = 1,
  beta = 1,
  direction_threshold = 0.10,
  prune_threshold = 0.001,
  time_weight_mode = "directional_confidence"
)

head(
  tata_manual$edge_table[
    order(-tata_manual$edge_table$tata_weight),
    c(
      "cluster_a",
      "cluster_b",
      "topology_weight",
      "flow_score",
      "time_weight",
      "time_consistency",
      "tata_weight",
      "direction_label",
      "kept"
    )
  ],
  10
)
```

The cluster graph is easiest to interpret when it is overlaid onto the same
embedding used to view the cells:

```{r eval = run_tata_demo}
plot_tata_cluster_graph(
  list(
    cluster_graph = tata_manual$graph,
    sce = sce,
    parameters = list(cluster_col = "tata_cluster_step")
  ),
  dimred = "UMAP",
  layout_mode = "embedding",
  show_cells = TRUE,
  colour_by = "tata_cluster_step",
  node_colour_by = "cluster",
  main = "Manual TATA graph aligned to UMAP"
)
```

## Run the full wrapper

For most analyses it is simpler to call `run_tata()` directly and let the
package carry out graph construction, clustering, edge scoring, and
pseudotime.

```{r eval = run_tata_demo}
tata <- run_tata(
  sce = sce,
  dimred = "PCA",
  time_col = "timepoint",
  cluster_col = "tata_cluster",
  k = 30,
  cluster_method = "louvain",
  alpha = 1,
  beta = 1,
  direction_threshold = 0.10,
  prune_threshold = 0.001,
  time_weight_mode = "directional_confidence",
  do_pseudotime = TRUE,
  seed = 101
)

sce_tata <- tata$sce
sce_tata$tata_cluster <- factor(sce_tata$tata_cluster)
```

Inspect the wrapper output on the embedding:

```{r eval = run_tata_demo}
PlotReduction(sce_tata, reduction = "UMAP", colour_by = "tata_cluster")
PlotReduction(sce_tata, reduction = "UMAP", colour_by = "tata_pseudotime_scaled")
```

```{r eval = run_tata_demo}
plot_tata_cluster_graph(
  tata,
  dimred = "UMAP",
  layout_mode = "embedding",
  show_cells = TRUE,
  colour_by = "timepoint",
  node_colour_by = "cluster",
  main = "TATA abstract graph aligned to UMAP"
)
```

`plot_tata_trajectory_embedding()` provides a more classic trajectory overlay.
Use `mode = "full"` to draw all inferred paths or `mode = "combined"` to show
one long summary arrow per major branch.

```{r eval = run_tata_demo}
plot_tata_trajectory_embedding(
  tata,
  dimred = "UMAP",
  colour_by = "timepoint",
  mode = "full",
  smooth_paths = TRUE
)
```

```{r eval = run_tata_demo}
plot_tata_trajectory_embedding(
  tata,
  dimred = "UMAP",
  colour_by = "timepoint",
  mode = "combined",
  max_paths = 3,
  smooth_paths = TRUE
)
```

## Notes

- `build_knn_graph()`, `compute_topology_weights()`, and
  `compute_time_weights()` are available separately for manual inspection and
  method development.
- `plot_tata_cluster_graph()` places the abstract graph either in a graph-only
  layout or directly on top of a chosen embedding.
- `plot_tata_trajectory_embedding()` is useful when you want a simpler
  trajectory-style overlay rather than the full abstract graph.
