---
title: "TATA Step-by-Step Multi-Drug Demo"
format: html
execute:
  echo: true
  warning: false
  message: false
vignette: >
  %\VignetteIndexEntry{TATA step-by-step multi-drug demo}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

## Load the package functions

This demo uses `devtools::load_all()` so the current source tree is loaded
directly. The `reset = TRUE` argument is helpful during development because it
avoids using a stale version of a function from a previous session.

```{r}
library(devtools)
devtools::load_all("..", quiet = TRUE, reset = TRUE)

library(ggplot2)
library(igraph)
library(Matrix)
library(scater)
library(standR)
```

## Simulate a center-rooted multi-drug trajectory

We start with a simulated dataset where the earliest cells sit near the middle
of the embedding. This makes the geometry intentionally awkward for
topology-only trajectory methods:

- `drug_A` extends in one direction and emerges early, around 24 hours
- `drug_C` extends along a third arm and emerges around 48 hours
- `drug_B` extends in the opposite direction to `drug_A` and emerges late,
  around 96 hours

The idea is to create a case where time should help orient the abstract graph,
because the root is not sitting at an obvious visual tip.

```{r}
set.seed(101)

sce <- simulate_tata_multidrug_sce(
  n_cells = 5000,
  n_features = 100,
  n_pcs = 20,
  seed = 101
)

sce$timepoint_factor <- factor(
  sce$timepoint,
  levels = sort(unique(sce$timepoint))
)
```

```{r}
with(as.data.frame(SummarizedExperiment::colData(sce)), table(condition, timepoint))
```

## Visualise the simulated embedding

These plots show the same UMAP coloured in different ways so we can inspect the
simulation before running any trajectory inference.

```{r}
standR::plotDR(sce, dimred = "UMAP", colour = timepoint_factor) +
  ggplot2::ggtitle("Simulated embedding coloured by experimental time")
```

```{r}
standR::plotDR(sce, dimred = "UMAP", colour = condition) +
  ggplot2::ggtitle("Simulated embedding coloured by drug condition")
```

```{r}
standR::plotDR(sce, dimred = "UMAP", colour = simulated_state) +
  ggplot2::ggtitle("Simulated embedding coloured by ground-truth state")
```

```{r}
standR::plotDR(sce, dimred = "UMAP", colour = branch_activation) +
  ggplot2::scale_colour_viridis_c(option = "magma") +
  ggplot2::ggtitle("Simulated embedding coloured by branch activation")
```

## Step 1: Build the cell-cell kNN graph

TATA starts from a low-dimensional representation of cells. Here we use the
stored PCA coordinates and build an undirected kNN graph.

```{r}
knn <- build_knn_graph(
  x = sce,
  dimred = "PCA",
  k = 30
)

c(
  n_cells = igraph::vcount(knn$graph),
  n_edges = igraph::ecount(knn$graph),
  k = knn$k
)
```

To make the cell graph easier to inspect, we draw a random subset of graph
edges on top of the UMAP.

```{r}
umap_df <- as.data.frame(SingleCellExperiment::reducedDim(sce, "UMAP"))
colnames(umap_df)[1:2] <- c("UMAP1", "UMAP2")
umap_df$timepoint <- sce$timepoint

edge_df <- Matrix::summary(Matrix::triu(knn$adjacency, k = 1))
set.seed(101)
edge_df <- edge_df[sample(seq_len(nrow(edge_df)), min(5000, nrow(edge_df))), , drop = FALSE]

edge_plot_df <- data.frame(
  x = umap_df$UMAP1[edge_df$i],
  y = umap_df$UMAP2[edge_df$i],
  xend = umap_df$UMAP1[edge_df$j],
  yend = umap_df$UMAP2[edge_df$j]
)

ggplot() +
  geom_segment(
    data = edge_plot_df,
    aes(x = x, y = y, xend = xend, yend = yend),
    colour = "grey82",
    alpha = 0.18,
    linewidth = 0.2
  ) +
  geom_point(
    data = umap_df,
    aes(x = UMAP1, y = UMAP2, colour = timepoint),
    size = 0.5,
    alpha = 0.8
  ) +
  scale_colour_viridis_c(option = "plasma") +
  theme_classic() +
  labs(
    title = "A sampled view of the cell-cell kNN graph on UMAP",
    x = "UMAP_1",
    y = "UMAP_2",
    colour = "timepoint"
  )
```

## Step 2: Cluster the cell graph into coarse states

The cell graph is next clustered into partitions or states. These states are
the nodes of the final abstract TATA graph.

```{r}
clusters <- cluster_graph_states(
  cell_graph = knn$graph,
  method = "louvain",
  seed = 101
)

sce$tata_cluster_step <- clusters
sce$tata_cluster_step <- factor(sce$tata_cluster_step)

table(sce$tata_cluster_step)
```

```{r}
standR::plotDR(sce, dimred = "UMAP", colour = tata_cluster_step) +
  ggplot2::ggtitle("UMAP coloured by graph-based TATA states")
```

## Step 3: Compute topology-based connectivity between clusters

For every cluster pair, we count inter-cluster cell-cell edges and compare that
count to a degree-aware expectation. This gives the topology term used by TATA.

```{r}
topology_table <- compute_topology_weights(
  adjacency = knn$adjacency,
  clusters = sce$tata_cluster_step
)

head(
  topology_table[
    order(-topology_table$topology_weight),
    c("cluster_a", "cluster_b", "observed_edges", "expected_edges", "topology_weight")
  ],
  12
)
```

```{r}
topology_top <- topology_table[order(-topology_table$topology_weight), ][1:15, ]
topology_top$pair <- paste(topology_top$cluster_a, topology_top$cluster_b, sep = " -- ")

ggplot(topology_top, aes(x = reorder(pair, topology_weight), y = topology_weight)) +
  geom_col(fill = "#2c7fb8") +
  coord_flip() +
  theme_classic() +
  labs(
    title = "Top cluster pairs by topology weight",
    x = "cluster pair",
    y = "topology weight"
  )
```

## Step 4: Compute time consistency for cluster-cluster connections

Now we use the real experimental time labels. For each cluster pair, TATA
computes:

- `flow_score`: the mean sign of time differences across inter-cluster edges
- `time_weight`: the ordered-pair score
- `time_consistency`: a direction-strength score used by the multi-branch-aware
  weighting mode

```{r}
time_table <- compute_time_weights(
  adjacency = knn$adjacency,
  clusters = sce$tata_cluster_step,
  timepoint = sce$timepoint
)

head(
  time_table[
    order(-time_table$time_consistency),
    c("cluster_a", "cluster_b", "n_time_edges", "flow_score", "time_weight", "time_consistency")
  ],
  12
)
```

```{r}
edge_metric_table <- merge(
  topology_table,
  time_table,
  by = c("cluster_a", "cluster_b"),
  all.x = TRUE,
  sort = FALSE
)

ggplot(
  edge_metric_table,
  aes(
    x = topology_weight,
    y = flow_score,
    size = observed_edges,
    colour = time_consistency
  )
) +
  geom_hline(yintercept = c(-0.10, 0.10), linetype = 2, colour = "grey60") +
  geom_point(alpha = 0.8) +
  scale_colour_viridis_c(option = "plasma") +
  theme_classic() +
  labs(
    title = "Topology and time consistency for cluster pairs",
    x = "topology weight",
    y = "temporal flow score",
    colour = "time consistency",
    size = "observed edges"
  )
```

## Step 5: Build the abstract TATA graph

Here we combine topology and time into the final TATA edge weight. We use
`time_weight_mode = "directional_confidence"` because it is more robust in
multi-branch settings where the biological direction may not match the
alphabetical ordering of the cluster labels.

```{r}
tata_graph_manual <- build_tata_graph(
  topology_table = topology_table,
  time_table = time_table,
  clusters = sce$tata_cluster_step,
  embedding = SingleCellExperiment::reducedDim(sce, "PCA"),
  timepoint = sce$timepoint,
  alpha = 1,
  beta = 1,
  direction_threshold = 0.10,
  prune_threshold = 0.001,
  time_weight_mode = "directional_confidence"
)

head(
  tata_graph_manual$edge_table[
    order(-tata_graph_manual$edge_table$tata_weight),
    c(
      "cluster_a",
      "cluster_b",
      "topology_weight",
      "flow_score",
      "time_weight",
      "time_consistency",
      "tata_weight",
      "direction_label",
      "kept"
    )
  ],
  12
)
```

```{r}
plot_tata_cluster_graph(
  list(
    cluster_graph = tata_graph_manual$graph,
    sce = sce,
    parameters = list(cluster_col = "tata_cluster_step")
  ),
  dimred = "UMAP",
  layout_mode = "embedding",
  show_cells = TRUE,
  colour_by = "tata_cluster_step",
  node_colour_by = "cluster",
  main = "Manual TATA graph aligned to UMAP"
)
```

## Step 6: Compute pseudotime on the directed TATA graph

The root cluster is chosen from the earliest clusters, with a centrality-aware
tie-break. Cluster-level pseudotime is then propagated to cells using
within-cluster time ranks.

```{r}
sce_manual <- sce
sce_manual$tata_cluster_step <- sce$tata_cluster_step

pseudotime_manual <- compute_tata_pseudotime(
  cluster_graph = tata_graph_manual$graph,
  sce = sce_manual,
  cluster_col = "tata_cluster_step",
  time_col = "timepoint"
)

sce_manual$tata_pseudotime_step <- pseudotime_manual$cell_pseudotime_scaled

pseudotime_manual$cluster_pseudotime
```

```{r}
standR::plotDR(sce_manual, dimred = "UMAP", colour = tata_pseudotime_step) +
  ggplot2::scale_colour_viridis_c(option = "plasma") +
  ggplot2::ggtitle("UMAP coloured by manually computed TATA pseudotime")
```

## Step 7: Run the convenience wrapper

The `run_tata()` wrapper performs the same major steps in one call and returns
all of the intermediate and final objects in a package-friendly list.

```{r}
tata <- run_tata(
  sce = sce,
  dimred = "PCA",
  time_col = "timepoint",
  cluster_col = "tata_cluster",
  k = 30,
  cluster_method = "louvain",
  alpha = 1,
  beta = 1,
  time_weight_mode = "directional_confidence",
  direction_threshold = 0.10,
  prune_threshold = 0.001,
  do_pseudotime = TRUE,
  seed = 101
)

sce_tata <- tata$sce
sce_tata$tata_cluster <- factor(sce_tata$tata_cluster)
```

```{r}
names(tata)
```

## Step 8: Visualise the wrapper output

The first graph plot is the closest analogue to a PAGA-style compare view:
cluster nodes are placed at their UMAP centroids so the abstract graph lives in
the same visual space as the embedding.

```{r}
plot_tata_cluster_graph(
  tata,
  dimred = "UMAP",
  layout_mode = "embedding",
  show_cells = TRUE,
  colour_by = "tata_cluster",
  node_colour_by = "cluster",
  main = "TATA abstract graph aligned to UMAP"
)
```

The first plot uses the detailed path view. This is useful for debugging
branch-level structure because it shows every retained root-to-terminal path.

```{r}
plot_tata_trajectory_embedding(
  tata,
  dimred = "UMAP",
  colour_by = "timepoint",
  mode = "full",
  smooth_paths = TRUE
) +
  ggplot2::ggtitle("Time-aware TATA trajectories on UMAP: full mode")
```

The second plot uses the new combined mode. It compresses paths with a shared
trunk and shows only one long arrow per major branch.

```{r}
plot_tata_trajectory_embedding(
  tata,
  dimred = "UMAP",
  colour_by = "timepoint",
  mode = "combined",
  max_paths = 3,
  smooth_paths = TRUE
) +
  ggplot2::ggtitle("Time-aware TATA trajectories on UMAP: combined mode")
```

```{r}
standR::plotDR(sce_tata, dimred = "UMAP", colour = tata_cluster) +
  ggplot2::ggtitle("UMAP coloured by TATA-inferred cluster")
```

```{r}
standR::plotDR(sce_tata, dimred = "UMAP", colour = tata_pseudotime_scaled) +
  ggplot2::scale_colour_viridis_c(option = "plasma") +
  ggplot2::ggtitle("UMAP coloured by TATA scaled pseudotime")
```

## Optional comparison: topology-only baseline

Finally, we remove the time contribution by setting `beta = 0`. This gives a
useful visual comparison against the full time-aware result.

```{r}
tata_topology_only <- run_tata(
  sce = sce,
  dimred = "PCA",
  time_col = "timepoint",
  cluster_col = "tata_cluster_topology_only",
  k = 30,
  cluster_method = "louvain",
  alpha = 1,
  beta = 0,
  time_weight_mode = "directional_confidence",
  direction_threshold = 0.10,
  prune_threshold = 0.001,
  do_pseudotime = TRUE,
  seed = 101
)
```

```{r}
plot_tata_trajectory_embedding(
  tata_topology_only,
  dimred = "UMAP",
  colour_by = "timepoint",
  mode = "full",
  max_paths = 3,
  smooth_paths = TRUE
) +
  ggplot2::ggtitle("Topology-only baseline on UMAP: combined mode")
```


# benchmark with other tools

This section compares the time-aware TATA workflow against two widely used
alternatives on the same simulated dataset:

- `slingshot`, a branching trajectory method based on cluster topology and
  principal curves
- `scShaper`, a fast ensemble method designed for linear pseudotime inference

The benchmark is intentionally not summarized by a single "winner" score,
because this simulation was designed to stress a specific weakness of
topology-only methods: the biological origin sits near the middle of the
embedding, and one treatment branch emerges later than the others. We therefore
look at several complementary metrics:

- `graph_based_trustworthiness`: an embedding-based score that asks whether
  cells that become neighbours in the inferred trajectory space were also close
  in the known latent trajectory
- `neighborhood_preservation_score`: an embedding-based score that measures how
  much the local `k`-nearest-neighbour structure is preserved between the known
  latent trajectory and the inferred trajectory space
- `wasserstein_distance_for_embedding`: a higher-is-better score derived from a
  Wasserstein-style comparison between the pairwise-distance distributions of
  the known latent trajectory and the inferred trajectory space
- `local_time_concordance`: the fraction of local cell-cell graph edges whose
  inferred pseudotime direction agrees with the true experimental time
  direction
- `local_temporal_direction_score`: a higher-is-better transformation of the
  local inversion rate, so values near `1` indicate that local pseudotime
  directions mostly agree with experimental time
- `adjacent_timepoint_auc`: for each pair of adjacent timepoints, the
  probability that later cells have larger inferred pseudotime than earlier
  cells
- `branch_onset_order_score`: a higher-is-better rescaling of the Kendall
  correlation between the known branch-emergence order
  (`drug_A` < `drug_C` < `drug_B`) and the inferred onset order
- `branch_onset_accuracy`: a higher-is-better transformation of the normalized
  branch-onset error

Because `scShaper` is not distributed through Bioconductor, the benchmark code
below is optional and will be skipped automatically if either external package
is unavailable. The vignette still renders without the benchmark, but when the
packages are installed this section provides a useful side-by-side comparison.

```{r}
have_benchmark_pkgs <- requireNamespace("slingshot", quietly = TRUE) &&
  requireNamespace("scShaper", quietly = TRUE)

if (have_benchmark_pkgs) {
  library(slingshot)
  library(scShaper)
  library(SingleCellExperiment)
} else {
  message(
    "Benchmark section skipped because `slingshot` or `scShaper` is not installed."
  )
}
```

## Benchmark helpers

The next chunk defines a few small helper functions used only in this vignette.
The more generally useful temporal benchmark metrics are now available directly
from the package; the local helpers below are mainly for constructing the
paper-inspired latent spaces and computing the embedding-based benchmark
metrics.

```{r}
if (have_benchmark_pkgs) {
  scale_to_unit <- function(x) {
    rng <- range(x, na.rm = TRUE)
    if (!all(is.finite(rng)) || diff(rng) < .Machine$double.eps) {
      return(rep(0, length(x)))
    }
    (x - rng[1]) / diff(rng)
  }

  normalize_rows <- function(mat) {
    rs <- rowSums(mat, na.rm = TRUE)
    rs[!is.finite(rs) | rs <= 0] <- 1
    sweep(mat, 1, rs, "/")
  }

  stratified_sample_idx <- function(sce, n_per_group = 70, seed = 101) {
    set.seed(seed)
    group <- interaction(sce$condition, sce$timepoint, drop = TRUE)
    split_idx <- split(seq_len(ncol(sce)), group)
    unlist(lapply(split_idx, function(idx) {
      sample(idx, min(length(idx), n_per_group))
    }), use.names = FALSE)
  }

  make_truth_space <- function(sce) {
    pt <- scale_to_unit(sce$progress)
    branch_levels <- c("drug_A", "drug_B", "drug_C")
    branch_mat <- sapply(branch_levels, function(branch_name) {
      pt * ifelse(sce$simulated_branch == branch_name, sce$branch_activation, 0)
    })
    colnames(branch_mat) <- paste0("truth_", branch_levels)
    cbind(progress = pt, branch_mat)
  }

  make_tata_space <- function(tata_result, sce) {
    cluster_col <- tata_result$parameters$cluster_col
    graph <- tata_result$cluster_graph
    cluster_ids <- as.character(SummarizedExperiment::colData(sce)[[cluster_col]])
    cell_pt <- orient_pseudotime_to_time(
      SummarizedExperiment::colData(sce)[["tata_pseudotime_scaled"]],
      sce$timepoint
    )

    terminals <- igraph::V(graph)$name[igraph::degree(graph, mode = "out") == 0]
    if (length(terminals) == 0L) {
      terminals <- tail(
        tata_result$cluster_pseudotime$cluster[
          order(tata_result$cluster_pseudotime$pseudotime)
        ],
        1
      )
    }

    terminal_weights <- matrix(
      0,
      nrow = igraph::vcount(graph),
      ncol = length(terminals),
      dimnames = list(igraph::V(graph)$name, paste0("tata_", terminals))
    )

    for (cluster_name in rownames(terminal_weights)) {
      dist_out <- as.numeric(igraph::distances(
        graph,
        v = cluster_name,
        to = terminals,
        mode = "out"
      ))
      reachable <- is.finite(dist_out)

      if (any(reachable)) {
        terminal_weights[cluster_name, reachable] <- 1
      } else {
        dist_undirected <- as.numeric(igraph::distances(
          igraph::as_undirected(graph, mode = "collapse"),
          v = cluster_name,
          to = terminals,
          mode = "all"
        ))
        closest <- which.min(dist_undirected)
        terminal_weights[cluster_name, closest] <- 1
      }
    }

    terminal_weights <- normalize_rows(terminal_weights)
    cell_weights <- terminal_weights[cluster_ids, , drop = FALSE]
    cell_weights[!is.finite(cell_weights)] <- 0
    branch_space <- sweep(cell_weights, 1, cell_pt, "*")
    cbind(pseudotime = cell_pt, branch_space)
  }

  make_slingshot_space <- function(sce_sling, truth_reference) {
    pt <- slingshot::slingPseudotime(sce_sling)
    wt <- slingshot::slingCurveWeights(sce_sling)

    if (is.null(dim(pt))) {
      pt <- matrix(pt, ncol = 1)
    }
    if (is.null(dim(wt))) {
      wt <- matrix(wt, ncol = 1)
    }

    pt_scaled <- pt
    for (j in seq_len(ncol(pt_scaled))) {
      pt_scaled[, j] <- orient_pseudotime_to_time(pt_scaled[, j], truth_reference)
    }

    wt[!is.finite(wt)] <- 0
    pt_scaled[!is.finite(pt_scaled)] <- 0
    wt <- normalize_rows(wt)

    aggregate_pt <- rowSums(pt_scaled * wt)
    aggregate_pt <- orient_pseudotime_to_time(aggregate_pt, truth_reference)

    branch_space <- pt_scaled * wt
    colnames(branch_space) <- paste0("sling_lineage_", seq_len(ncol(branch_space)))
    cbind(pseudotime = aggregate_pt, branch_space)
  }

  make_scshaper_space <- function(sce_scshaper, truth_reference) {
    sc_meta <- S4Vectors::metadata(sce_scshaper)[["scshaper"]]
    pt <- sc_meta[["continuous.pseudotime"]]
    if (any(!is.finite(pt))) {
      pt <- sc_meta[["average.discrete.pseudotime"]]
    }
    pt <- orient_pseudotime_to_time(pt, truth_reference)
    cbind(pseudotime = pt)
  }

  graph_based_trustworthiness <- function(truth_space, inferred_space, idx, k = 15L) {
    truth_mat <- as.matrix(truth_space[idx, , drop = FALSE])
    inferred_mat <- as.matrix(inferred_space[idx, , drop = FALSE])
    n <- nrow(truth_mat)

    if (n < 4L) {
      return(NA_real_)
    }

    k <- max(1L, min(as.integer(k), n - 2L))

    truth_dist <- as.matrix(stats::dist(truth_mat))
    inferred_dist <- as.matrix(stats::dist(inferred_mat))
    diag(truth_dist) <- Inf
    diag(inferred_dist) <- Inf

    truth_knn <- t(apply(truth_dist, 1, order))[ , seq_len(k), drop = FALSE]
    inferred_knn <- t(apply(inferred_dist, 1, order))[ , seq_len(k), drop = FALSE]
    truth_rank <- t(apply(truth_dist, 1, rank, ties.method = "average"))

    penalty <- 0
    for (i in seq_len(n)) {
      false_neighbors <- setdiff(inferred_knn[i, ], truth_knn[i, ])
      if (length(false_neighbors) > 0L) {
        penalty <- penalty + sum(truth_rank[i, false_neighbors] - k)
      }
    }

    denom <- n * k * (2 * n - 3 * k - 1)
    if (denom <= 0) {
      return(NA_real_)
    }

    1 - (2 * penalty) / denom
  }

  neighborhood_preservation_score <- function(truth_space, inferred_space, idx, k = 15L) {
    truth_mat <- as.matrix(truth_space[idx, , drop = FALSE])
    inferred_mat <- as.matrix(inferred_space[idx, , drop = FALSE])
    n <- nrow(truth_mat)

    if (n < 3L) {
      return(NA_real_)
    }

    k <- max(1L, min(as.integer(k), n - 1L))

    truth_dist <- as.matrix(stats::dist(truth_mat))
    inferred_dist <- as.matrix(stats::dist(inferred_mat))
    diag(truth_dist) <- Inf
    diag(inferred_dist) <- Inf

    truth_knn <- t(apply(truth_dist, 1, order))[ , seq_len(k), drop = FALSE]
    inferred_knn <- t(apply(inferred_dist, 1, order))[ , seq_len(k), drop = FALSE]

    mean(vapply(seq_len(n), function(i) {
      length(intersect(truth_knn[i, ], inferred_knn[i, ])) / k
    }, numeric(1)))
  }

  wasserstein_distance_for_embedding <- function(
      truth_space,
      inferred_space,
      idx,
      n_quantiles = 200L) {
    truth_dist <- scale_to_unit(as.numeric(stats::dist(truth_space[idx, , drop = FALSE])))
    inferred_dist <- scale_to_unit(as.numeric(stats::dist(inferred_space[idx, , drop = FALSE])))

    if (length(truth_dist) < 3L || length(inferred_dist) < 3L) {
      return(NA_real_)
    }

    probs <- seq(0, 1, length.out = max(20L, as.integer(n_quantiles)))
    truth_q <- stats::quantile(truth_dist, probs = probs, type = 8, na.rm = TRUE)
    inferred_q <- stats::quantile(inferred_dist, probs = probs, type = 8, na.rm = TRUE)
    w1 <- mean(abs(truth_q - inferred_q), na.rm = TRUE)

    1 / (1 + w1)
  }
}
```

## Run Slingshot and scShaper on the same dataset

For a fair comparison, we benchmark all three methods on the same simulated
cells. We keep the TATA wrapper settings the same as above, run Slingshot on the
same graph-derived clusters, and run scShaper in PCA mode. The Slingshot run
uses `dist.method = "simple"` because the default covariance-based cluster
distance is numerically singular on this deliberately center-rooted geometry.

```{r}
if (have_benchmark_pkgs) {
  tata_benchmark_time <- system.time({
    tata_benchmark <- run_tata(
      sce = sce,
      dimred = "PCA",
      time_col = "timepoint",
      cluster_col = "tata_cluster_benchmark",
      k = 30,
      cluster_method = "louvain",
      alpha = 1,
      beta = 1,
      time_weight_mode = "directional_confidence",
      direction_threshold = 0.10,
      prune_threshold = 0.001,
      do_pseudotime = TRUE,
      seed = 101
    )
  })

  slingshot_time <- system.time({
    sce_sling <- slingshot::slingshot(
      sce,
      clusterLabels = tata_benchmark$sce$tata_cluster_benchmark,
      reducedDim = "PCA",
      dist.method = "simple",
      omega = TRUE,
      approx_points = 100
    )
  })

  scshaper_time <- system.time({
    sce_scshaper <- scShaper::RunscShaper(
      sce,
      dim.red.method = "pca",
      num.pcs = 20,
      k.range = seq(4, 24, 4),
      pca.rank = 2,
      span = 0.15
    )
  })
}
```

## Extract comparable pseudotime representations

TATA and Slingshot both infer branching structure, whereas scShaper returns a
single continuous pseudotime. To compare them on the same footing, we extract:

- one global pseudotime per method for colouring and timing analyses
- one branch-aware latent space per method for the embedding-based metrics

```{r}
if (have_benchmark_pkgs) {
  truth_space <- make_truth_space(sce)
  tata_benchmark_space <- make_tata_space(tata_benchmark, tata_benchmark$sce)
  slingshot_space <- make_slingshot_space(sce_sling, sce$timepoint)
  scshaper_space <- make_scshaper_space(sce_scshaper, sce$timepoint)

  benchmark_eval_idx <- stratified_sample_idx(sce, n_per_group = 70, seed = 101)

  benchmark_spaces <- list(
    TATA = tata_benchmark_space,
    Slingshot = slingshot_space,
    scShaper = scshaper_space
  )

  benchmark_pseudotime <- list(
    TATA = tata_benchmark_space[, "pseudotime"],
    Slingshot = slingshot_space[, "pseudotime"],
    scShaper = scshaper_space[, "pseudotime"]
  )
}
```

## Visualise inferred pseudotime from each method

These faceted UMAP plots show how each method distributes cells along its
inferred pseudotime axis. TATA should look more consistent with the known
experimental ordering, while scShaper compresses the branching structure into a
single linear path.

```{r}
if (have_benchmark_pkgs) {
  umap_df <- as.data.frame(SingleCellExperiment::reducedDim(sce, "UMAP"))
  colnames(umap_df)[1:2] <- c("UMAP_1", "UMAP_2")

  benchmark_plot_df <- do.call(rbind, lapply(names(benchmark_pseudotime), function(method_name) {
    data.frame(
      umap_df,
      method = method_name,
      pseudotime = benchmark_pseudotime[[method_name]],
      stringsAsFactors = FALSE
    )
  }))

  ggplot(benchmark_plot_df, aes(x = UMAP_1, y = UMAP_2, colour = pseudotime)) +
    geom_point(size = 0.45, alpha = 0.75) +
    scale_colour_viridis_c(option = "plasma") +
    facet_wrap(~ method, nrow = 1) +
    theme_classic() +
    labs(
      title = "Benchmark methods coloured by inferred pseudotime",
      x = "UMAP_1",
      y = "UMAP_2",
      colour = "pseudotime"
    )
}
```

The next view compares inferred pseudotime directly against the known simulated
progress variable. This makes it easier to see where a method respects the
global progression but collapses or reorders particular branches.

```{r}
if (have_benchmark_pkgs) {
  progress_plot_df <- do.call(rbind, lapply(names(benchmark_pseudotime), function(method_name) {
    data.frame(
      true_progress = sce$progress,
      inferred_pseudotime = benchmark_pseudotime[[method_name]],
      condition = sce$condition,
      method = method_name,
      stringsAsFactors = FALSE
    )
  }))

  ggplot(progress_plot_df, aes(x = true_progress, y = inferred_pseudotime, colour = condition)) +
    geom_point(size = 0.45, alpha = 0.60) +
    facet_wrap(~ method, nrow = 1) +
    theme_classic() +
    labs(
      title = "Known latent progress versus inferred pseudotime",
      x = "true simulated progress",
      y = "inferred pseudotime"
    )
}
```

## Benchmark metric catalog

Before computing the scores, the table below summarizes the benchmark panel in
the same broad style used by scTRAM: embedding-based, pseudotime-based, and
adjacency-based metrics. The main comparison across TATA, Slingshot, and
scShaper uses the first eight rows; the final two graph diagnostics are shown
separately because they are specific to TATA's abstract graph output.

```{r}
metric_catalog <- data.frame(
  category = c(
    "Embedding-based",
    "Embedding-based",
    "Embedding-based",
    "Pseudotime-based",
    "Pseudotime-based",
    "Pseudotime-based",
    "Pseudotime-based",
    "Pseudotime-based",
    "Adjacency-based",
    "Adjacency-based"
  ),
  metric = c(
    "Graph-Based Trustworthiness",
    "Neighborhood Preservation Score",
    "Wasserstein Distance for Embedding",
    "local_time_concordance",
    "local_temporal_direction_score",
    "adjacent_timepoint_auc",
    "branch_onset_order_score",
    "branch_onset_accuracy",
    "directed_reachability_concordance",
    "causal_edge_score"
  ),
  what_it_measures = c(
    "Whether neighbours recovered in the inferred trajectory space were also close in the known latent trajectory",
    "How well local k-nearest-neighbour structure is preserved between the known latent trajectory and the inferred trajectory space",
    "A higher-is-better transformation of the Wasserstein discrepancy between normalized pairwise-distance distributions in the truth and inferred spaces",
    "The fraction of local cell-graph edges whose inferred direction agrees with experimental time",
    "A higher-is-better summary of how rarely local pseudotime steps point backward in experimental time",
    "How often later timepoints receive larger pseudotime than adjacent earlier timepoints",
    "How well the inferred order of branch emergence matches the known order drug_A < drug_C < drug_B",
    "How close the inferred branch-onset positions are to the known onset timing after normalization",
    "Whether earlier clusters can reach later clusters in the directed abstract graph",
    "The fraction of abstract-graph edge weight that does not point backward in time"
  ),
  higher_is_better = TRUE,
  stringsAsFactors = FALSE
)

metric_catalog
```

## Compute scTRAM-style embedding and time-aware benchmark metrics

The embedding-based metrics are designed to resemble the scTRAM evaluation
style, while the temporal columns are more directly targeted at the question
TATA is trying to solve: does the inferred ordering respect the known
experimental time progression? For readability, all scores shown in the main
table below are formatted so that higher values are better.

```{r}
if (have_benchmark_pkgs) {
  benchmark_table <- do.call(rbind, lapply(names(benchmark_spaces), function(method_name) {
    local_result <- local_temporal_order_metrics(
      benchmark_pseudotime[[method_name]],
      tata_benchmark$adjacency_matrix,
      sce$timepoint
    )
    auc_result <- adjacent_timepoint_auc(
      benchmark_pseudotime[[method_name]],
      sce$timepoint,
      group = sce$condition,
      average = "weighted",
      min_cells = 20
    )
    onset_result <- branch_onset_metrics(
      benchmark_pseudotime[[method_name]],
      branch = sce$condition,
      branch_activation = sce$branch_activation,
      true_onset = c(drug_A = 24, drug_C = 48, drug_B = 96),
      activation_threshold = 0.25
    )
    data.frame(
      method = method_name,
      graph_based_trustworthiness = graph_based_trustworthiness(
        truth_space,
        benchmark_spaces[[method_name]],
        benchmark_eval_idx
      ),
      neighborhood_preservation_score = neighborhood_preservation_score(
        truth_space,
        benchmark_spaces[[method_name]],
        benchmark_eval_idx
      ),
      wasserstein_distance_for_embedding = wasserstein_distance_for_embedding(
        truth_space,
        benchmark_spaces[[method_name]],
        benchmark_eval_idx
      ),
      local_time_concordance = local_result$local_time_concordance,
      local_temporal_direction_score = local_result$local_temporal_direction_score,
      adjacent_timepoint_auc = auc_result$summary$adjacent_timepoint_auc,
      branch_onset_order_score = onset_result$summary$branch_onset_order_score,
      branch_onset_accuracy = onset_result$summary$branch_onset_accuracy,
      runtime_seconds = c(
        TATA = tata_benchmark_time[["elapsed"]],
        Slingshot = slingshot_time[["elapsed"]],
        scShaper = scshaper_time[["elapsed"]]
      )[method_name],
      stringsAsFactors = FALSE
    )
  }))

  benchmark_table
}
```

```{r}
if (have_benchmark_pkgs) {
  metric_plot_df <- rbind(
    data.frame(method = benchmark_table$method, metric = "Graph-Based Trustworthiness", value = benchmark_table$graph_based_trustworthiness),
    data.frame(method = benchmark_table$method, metric = "Neighborhood Preservation Score", value = benchmark_table$neighborhood_preservation_score),
    data.frame(method = benchmark_table$method, metric = "Wasserstein Distance for Embedding", value = benchmark_table$wasserstein_distance_for_embedding),
    data.frame(method = benchmark_table$method, metric = "local_time_concordance", value = benchmark_table$local_time_concordance),
    data.frame(method = benchmark_table$method, metric = "local_temporal_direction_score", value = benchmark_table$local_temporal_direction_score),
    data.frame(method = benchmark_table$method, metric = "adjacent_timepoint_auc", value = benchmark_table$adjacent_timepoint_auc),
    data.frame(method = benchmark_table$method, metric = "branch_onset_order_score", value = benchmark_table$branch_onset_order_score),
    data.frame(method = benchmark_table$method, metric = "branch_onset_accuracy", value = benchmark_table$branch_onset_accuracy)
  )

  ggplot(metric_plot_df, aes(x = method, y = value, fill = method)) +
    geom_col(width = 0.72) +
    facet_wrap(~ metric, scales = "free_y") +
    theme_classic() +
    theme(legend.position = "none") +
    labs(
      title = "Benchmark metrics across TATA, Slingshot, and scShaper",
      x = NULL,
      y = "score"
    )
}
```

The runtime plot is not the main scientific comparison, but it is useful for
understanding the practical trade-off between trajectory fidelity and speed.

```{r}
if (have_benchmark_pkgs) {
  ggplot(benchmark_table, aes(x = method, y = runtime_seconds, fill = method)) +
    geom_col(width = 0.72) +
    theme_classic() +
    theme(legend.position = "none") +
    labs(
      title = "Elapsed runtime on the simulated dataset",
      x = NULL,
      y = "seconds"
    )
}
```

## Compare branch-emergence timing explicitly

Finally, we compare the median inferred pseudotime of cells that are already
meaningfully committed to each drug branch. This is a deliberately hard test for
all three methods because the latest branch (`drug_B`) emerges from the shared
center only at 96 hours.

```{r}
if (have_benchmark_pkgs) {
  emergence_plot_df <- do.call(rbind, lapply(names(benchmark_pseudotime), function(method_name) {
    onset_result <- branch_onset_metrics(
      benchmark_pseudotime[[method_name]],
      branch = sce$condition,
      branch_activation = sce$branch_activation,
      true_onset = c(drug_A = 24, drug_C = 48, drug_B = 96),
      activation_threshold = 0.25
    )
    out <- onset_result$onset_table
    colnames(out)[colnames(out) == "branch"] <- "condition"
    colnames(out)[colnames(out) == "inferred_onset"] <- "median_pseudotime"
    out$method <- method_name
    out
  }))

  emergence_plot_df$condition <- factor(
    emergence_plot_df$condition,
    levels = c("drug_A", "drug_C", "drug_B")
  )

  ggplot(
    emergence_plot_df,
    aes(x = condition, y = median_pseudotime, colour = true_onset, group = 1)
  ) +
    geom_line(linewidth = 0.7, colour = "grey50") +
    geom_point(size = 3) +
    scale_colour_viridis_c(option = "plasma") +
    facet_wrap(~ method, nrow = 1) +
    theme_classic() +
    labs(
      title = "Median inferred pseudotime of committed cells in each drug branch",
      x = "drug branch",
      y = "median inferred pseudotime",
      colour = "true onset (h)"
    )
}
```

## TATA-specific graph temporal diagnostics

The previous table compares all methods on pseudotime-derived metrics. TATA also
produces a directed abstract graph, so we can additionally ask whether that
graph itself follows the known direction of time. The two diagnostics below are
useful for checking whether time-aware graph construction improves the abstract
trajectory beyond topology alone.

```{r}
if (have_benchmark_pkgs) {
  tata_graph_time <- directed_reachability_concordance(
    tata_benchmark$cluster_graph,
    clusters = tata_benchmark$sce$tata_cluster_benchmark,
    timepoint = tata_benchmark$sce$timepoint,
    weight_by = "cells"
  )

  tata_graph_causal <- anticausal_edge_mass(
    tata_benchmark$cluster_graph,
    clusters = tata_benchmark$sce$tata_cluster_benchmark,
    timepoint = tata_benchmark$sce$timepoint
  )

  tata_topology_graph_time <- directed_reachability_concordance(
    tata_topology_only$cluster_graph,
    clusters = tata_topology_only$sce$tata_cluster_topology_only,
    timepoint = tata_topology_only$sce$timepoint,
    weight_by = "cells"
  )

  tata_topology_graph_causal <- anticausal_edge_mass(
    tata_topology_only$cluster_graph,
    clusters = tata_topology_only$sce$tata_cluster_topology_only,
    timepoint = tata_topology_only$sce$timepoint
  )

  tata_graph_table <- data.frame(
    model = c("TATA time-aware", "TATA topology-only"),
    directed_reachability_concordance = c(
      tata_graph_time$summary$directed_reachability_concordance,
      tata_topology_graph_time$summary$directed_reachability_concordance
    ),
    causal_edge_score = c(
      tata_graph_causal$summary$causal_edge_score,
      tata_topology_graph_causal$summary$causal_edge_score
    ),
    stringsAsFactors = FALSE
  )

  tata_graph_table
}
```

Taken together, these results show the trade-off that motivated TATA. On this
center-rooted simulation, Slingshot remains strong on geometry-centric metrics
such as `Graph-Based Trustworthiness`, `Neighborhood Preservation Score`, and
`Wasserstein Distance for Embedding`, while TATA is much more faithful to the
known experimental ordering in metrics such as `local_time_concordance`,
`local_temporal_direction_score`, and `adjacent_timepoint_auc`. The extra
graph-level diagnostics then show whether that temporal signal is also reflected
in the abstract graph itself. In other words, TATA is not trying to replace
every geometry-driven method on every task; it is designed to inject real
experimental time directly into the abstract trajectory graph when temporal
plausibility matters.



# test on real data

```{r}
sce_test <- zellkonverter::readH5AD("../legacy/test_anndata.h5ad")

sce_test$time <- vapply(
  strsplit(as.character(sce_test$sample), "_"),
  function(x) as.integer(x[2]),
  integer(1)
)
```

```{r}
standR::plotDR(sce_test, dimred = "X_umap", color = time, size = .1) +
  scale_colour_viridis_c(option = "plasma")
```

```{r}
sce_test$treatment <- stringr::str_split_i(sce_test$sample,"_",1)
```

```{r}
tata_test <- run_tata(
  sce = sce_test,
  dimred = "X_pca",
  time_col = "time",
  cluster_col = "tata_cluster",
  k = 30,
  cluster_method = "louvain",
  alpha = 1,
  beta = 1,
  time_weight_mode = "directional_confidence",
  direction_threshold = 0.10,
  prune_threshold = 0.001,
  do_pseudotime = TRUE,
  seed = 101
)
```


```{r}
plot_tata_trajectory_embedding(
  tata_test,
  dimred = "X_fdl",
  colour_by = "time",
  mode = "full",
  smooth_paths = TRUE
) +
  ggplot2::ggtitle("Time-aware TATA trajectories on UMAP: full mode")
```

```{r}
plot_tata_cluster_graph(
  tata_test,
  dimred = "X_fdl",
  layout_mode = "embedding",
  show_cells = TRUE,
  colour_by = "tata_cluster",
  node_colour_by = "cluster",
  main = "TATA abstract graph aligned to UMAP"
)
```
