---
title: "SCProkaR TATA Workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{SCProkaR TATA Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 5
)

if (file.exists("../DESCRIPTION")) {
  if (requireNamespace("devtools", quietly = TRUE)) {
    devtools::load_all("..")
  } else {
    library(SCProkaR)
  }
} else {
  library(SCProkaR)
}

run_tata_demo <- TRUE
```

## Overview

This vignette demonstrates Time Aware Trajectory Analysis (TATA), a
PAGA-inspired trajectory abstraction that uses real experimental time labels to
score and direct cluster-level transitions.

The workflow below:

- simulates a center-rooted multi-drug time course
- visualizes the stored PCA and UMAP representations
- builds the cell-level kNN graph and graph-based clusters
- computes topology and time-consistency scores between clusters
- constructs the abstract TATA graph
- runs the one-step `run_tata()` wrapper and overlays the inferred trajectories
- computes terminal-branch probabilities and entropy/plasticity
- extracts branch-specific cells and branch-specific gene trends
- summarizes the results with an overview plotting helper

The simulation is intentionally awkward for topology-only methods because the
earliest cells are near the middle of the embedding rather than at a clear tip.

## Load packages

```{r}
library(SingleCellExperiment)
library(SummarizedExperiment)
```

## Simulate a multi-drug trajectory

```{r eval = run_tata_demo}
set.seed(101)

sce <- simulate_tata_multidrug_sce(
  n_cells = 2500,
  n_features = 100,
  n_pcs = 20,
  seed = 101
)

sce$timepoint_factor <- factor(
  sce$timepoint,
  levels = sort(unique(sce$timepoint))
)
```

Inspect the simulation design:

```{r eval = run_tata_demo}
with(as.data.frame(SummarizedExperiment::colData(sce)), table(condition, timepoint))
```

Visualize the stored embedding:

```{r eval = run_tata_demo}
PlotReduction(sce, reduction = "UMAP", colour_by = "timepoint_factor")
PlotReduction(sce, reduction = "UMAP", colour_by = "condition")
PlotReduction(sce, reduction = "UMAP", colour_by = "simulated_state")
```

## Build the cell-cell graph

TATA starts from a low-dimensional representation. Here we use the stored PCA
coordinates and connect each cell to its nearest neighbours.

```{r eval = run_tata_demo}
knn <- build_knn_graph(
  x = sce,
  dimred = "PCA",
  k = 30
)

c(
  n_cells = igraph::vcount(knn$graph),
  n_edges = igraph::ecount(knn$graph),
  k = knn$k
)
```

## Cluster cells into coarse states

The cell graph is clustered into coarse states that serve as nodes in the
abstract TATA graph.

```{r eval = run_tata_demo}
sce$tata_cluster_step <- factor(
  cluster_graph_states(
    cell_graph = knn$graph,
    method = "louvain",
    seed = 101
  )
)

table(sce$tata_cluster_step)
PlotReduction(sce, reduction = "UMAP", colour_by = "tata_cluster_step")
```

## Compute topology and time scores

TATA combines two ingredients for each cluster pair:

- a topology score based on how strongly the two clusters are connected in the
  cell-cell graph
- a time-consistency score based on whether those inter-cluster edges follow
  the observed experimental time order

```{r eval = run_tata_demo}
topology_table <- compute_topology_weights(
  adjacency = knn$adjacency,
  clusters = sce$tata_cluster_step
)

time_table <- compute_time_weights(
  adjacency = knn$adjacency,
  clusters = sce$tata_cluster_step,
  timepoint = sce$timepoint
)

head(
  topology_table[
    order(-topology_table$topology_weight),
    c("cluster_a", "cluster_b", "observed_edges", "expected_edges", "topology_weight")
  ],
  8
)

head(
  time_table[
    order(-time_table$time_consistency),
    c("cluster_a", "cluster_b", "n_time_edges", "flow_score", "time_weight", "time_consistency")
  ],
  8
)
```

```{r eval = run_tata_demo}
edge_metric_table <- merge(
  topology_table,
  time_table,
  by = c("cluster_a", "cluster_b"),
  all.x = TRUE,
  sort = FALSE
)

ggplot2::ggplot(
  edge_metric_table,
  ggplot2::aes(
    x = topology_weight,
    y = flow_score,
    size = observed_edges,
    colour = time_consistency
  )
) +
  ggplot2::geom_hline(yintercept = c(-0.10, 0.10), linetype = 2, colour = "grey60") +
  ggplot2::geom_point(alpha = 0.8) +
  ggplot2::scale_colour_viridis_c(option = "plasma") +
  ggplot2::theme_classic() +
  ggplot2::labs(
    title = "Topology and time consistency for cluster pairs",
    x = "topology weight",
    y = "temporal flow score",
    colour = "time consistency",
    size = "observed edges"
  )
```

## Build the abstract TATA graph

```{r eval = run_tata_demo}
tata_manual <- build_tata_graph(
  topology_table = topology_table,
  time_table = time_table,
  clusters = sce$tata_cluster_step,
  embedding = SingleCellExperiment::reducedDim(sce, "PCA"),
  timepoint = sce$timepoint,
  alpha = 1,
  beta = 1,
  direction_threshold = 0.10,
  prune_threshold = 0.001,
  time_weight_mode = "directional_confidence"
)

head(
  tata_manual$edge_table[
    order(-tata_manual$edge_table$tata_weight),
    c(
      "cluster_a",
      "cluster_b",
      "topology_weight",
      "flow_score",
      "time_weight",
      "time_consistency",
      "tata_weight",
      "direction_label",
      "kept"
    )
  ],
  10
)
```

The cluster graph is easiest to interpret when it is overlaid onto the same
embedding used to view the cells:

```{r eval = run_tata_demo}
plot_tata_cluster_graph(
  list(
    cluster_graph = tata_manual$graph,
    sce = sce,
    parameters = list(cluster_col = "tata_cluster_step")
  ),
  dimred = "UMAP",
  layout_mode = "embedding",
  show_cells = TRUE,
  colour_by = "tata_cluster_step",
  node_colour_by = "cluster",
  main = "Manual TATA graph aligned to UMAP"
)
```

For a more PAGA-style abstract view, the same graph can also be drawn in its
own layout without the embedding in the background:

```{r eval = run_tata_demo}
plot_tata_cluster_graph(
  list(
    cluster_graph = tata_manual$graph,
    sce = sce,
    parameters = list(cluster_col = "tata_cluster_step")
  ),
  layout_mode = "graph",
  node_colour_by = "median_time",
  main = "Manual TATA abstract graph"
)
```

The same edge table can also be summarized in the topology-versus-time plane.
This makes it easier to see which cluster pairs are kept because they are both
well connected and temporally plausible.

```{r eval = run_tata_demo}
plot_tata_edge_causality(
  list(cluster_edge_table = tata_manual$edge_table)
)
```

## Run the full wrapper

For most analyses it is simpler to call `run_tata()` directly and let the
package carry out graph construction, clustering, edge scoring, and
pseudotime. In this simulated example we also set `expected_branches = 3` to
help the downstream branch-probability layer recover the known three-arm
design.

```{r eval = run_tata_demo}
tata <- run_tata(
  sce = sce,
  dimred = "PCA",
  time_col = "timepoint",
  cluster_col = "tata_cluster",
  k = 30,
  cluster_method = "louvain",
  alpha = 1,
  beta = 1,
  direction_threshold = 0.10,
  prune_threshold = 0.001,
  time_weight_mode = "directional_confidence",
  do_pseudotime = TRUE,
  expected_branches = 3,
  seed = 101
)

sce_tata <- tata$sce
sce_tata$tata_cluster <- factor(sce_tata$tata_cluster)
```

Inspect the wrapper output on the embedding:

```{r eval = run_tata_demo}
PlotReduction(sce_tata, reduction = "UMAP", colour_by = "tata_cluster")
PlotReduction(sce_tata, reduction = "UMAP", colour_by = "tata_pseudotime_scaled")
```

```{r eval = run_tata_demo}
plot_tata_cluster_graph(
  tata,
  dimred = "UMAP",
  layout_mode = "embedding",
  show_cells = TRUE,
  colour_by = "timepoint",
  node_colour_by = "cluster",
  main = "TATA abstract graph aligned to UMAP"
)
```

```{r eval = run_tata_demo}
plot_tata_cluster_graph(
  tata,
  layout_mode = "graph",
  node_colour_by = "median_time",
  main = "TATA abstract graph"
)
```

`plot_tata_trajectory_embedding()` provides a more classic trajectory overlay.
Use `max_paths` to limit the number of displayed root-to-terminal paths when
you want a cleaner summary.

```{r eval = run_tata_demo}
plot_tata_trajectory_embedding(
  tata,
  dimred = "UMAP",
  colour_by = "timepoint",
  smooth_paths = TRUE
)
```

```{r eval = run_tata_demo}
plot_tata_trajectory_embedding(
  tata,
  dimred = "UMAP",
  colour_by = "timepoint",
  max_paths = 3,
  smooth_paths = TRUE
)
```

## What time adds beyond topology

The cleanest comparison is against a topology-only ablation of the same TATA
workflow. This keeps graph construction and abstraction the same, but removes
the real time signal from edge scoring and direction assignment.

```{r eval = run_tata_demo}
tata_topology <- run_tata(
  sce = sce,
  dimred = "PCA",
  use_time = FALSE,
  time_col = NULL,
  cluster_col = "tata_cluster_topology",
  k = 30,
  cluster_method = "louvain",
  alpha = 1,
  beta = 1,
  direction_threshold = 0.10,
  prune_threshold = 0.001,
  do_pseudotime = TRUE,
  expected_branches = 3,
  seed = 101
)
```

```{r eval = run_tata_demo}
plot_tata_cluster_graph(
  tata_topology,
  dimred = "UMAP",
  layout_mode = "embedding",
  show_cells = TRUE,
  colour_by = "timepoint",
  node_colour_by = "cluster",
  main = "Topology-only TATA graph aligned to UMAP"
)
```

```{r eval = run_tata_demo}
plot_tata_time_calibration(
  tata,
  branch_col = "condition",
  summary = "median_iqr"
)
```

```{r eval = run_tata_demo}
plot_tata_time_calibration(
  tata_topology,
  time_col = "timepoint",
  branch_col = "condition",
  summary = "median_iqr"
)
```

```{r eval = run_tata_demo}
plot_tata_edge_causality(tata)
```

## Branch probabilities and plasticity

TATA now adds a Palantir-style downstream layer on top of the abstract graph:

- terminal-branch probabilities for each cluster and cell
- a branch-entropy score that measures how committed or ambiguous each cell is
- a hard branch assignment for convenient downstream selection

Inspect the inferred terminal branches:

```{r eval = run_tata_demo}
tata$terminal_clusters

head(
  tata$cluster_branch_probabilities[
    ,
    c(
      "cluster",
      grep("^tata_prob_", colnames(tata$cluster_branch_probabilities), value = TRUE),
      "tata_branch_entropy",
      "tata_branch_assignment"
    )
  ]
)
```

The entropy score highlights ambiguous cells, while the hard assignment provides
a simple branch label:

```{r eval = run_tata_demo}
PlotReduction(sce_tata, reduction = "UMAP", colour_by = "tata_branch_entropy")
PlotReduction(sce_tata, reduction = "UMAP", colour_by = "tata_branch_assignment")
```

To view the full branch-probability landscape, plot one panel per terminal
branch:

```{r eval = run_tata_demo}
plot_tata_branch_probabilities(
  tata,
  dimred = "UMAP",
  ncol = 2
)
```

## Branch selection and representative cells

The branch-selection helper is useful when you want to isolate the cells most
compatible with one lineage.

```{r eval = run_tata_demo}
branch_one <- tata$terminal_clusters[1]

branch_one_cells <- select_tata_branch_cells(
  tata,
  branch = branch_one,
  selection_mode = "soft",
  probability_threshold = 0.60
)

head(branch_one_cells)
sum(branch_one_cells$selected)
```

```{r eval = run_tata_demo}
plot_tata_branch_selection(
  tata,
  branch = branch_one,
  dimred = "UMAP",
  selection_mode = "soft",
  probability_threshold = 0.60
)
```

It is also helpful to inspect a few representative cells directly:

```{r eval = run_tata_demo}
plot_tata_terminal_probabilities(
  tata,
  n_cells = 4
)
```

On this simulated dataset we can also compare the inferred hard branch labels
against the known branch identities.

```{r eval = run_tata_demo}
plot_tata_branch_confusion(
  tata,
  truth_col = "condition",
  predicted_col = "tata_branch_assignment",
  normalize = "truth"
)
```

## Branch-specific gene trends

Once TATA has pseudotime and branch probabilities, we can fit branch-specific
gene-expression trends. For the simulation, we use genes from branch-associated
feature modules stored in `rowData(sce)$feature_module`.

```{r eval = run_tata_demo}
feature_modules <- SummarizedExperiment::rowData(sce_tata)$feature_module

trend_genes <- unique(c(
  rownames(sce_tata)[feature_modules == "branch_A"][1:2],
  rownames(sce_tata)[feature_modules == "branch_B"][1:2],
  rownames(sce_tata)[feature_modules == "branch_C"][1:2],
  rownames(sce_tata)[feature_modules == "condition"][1:2]
))

trend_genes
```

```{r eval = run_tata_demo}
tata_trends <- compute_tata_gene_trends(
  tata,
  genes = trend_genes
)

head(tata_trends$trend_table)
```

```{r eval = run_tata_demo}
plot_tata_gene_trends(
  tata_trends,
  ncol = 3
)
```

The fitted trends can also be clustered by shape and shown as a heatmap:

```{r eval = run_tata_demo}
trend_clusters <- cluster_tata_gene_trends(
  tata_trends,
  n_clusters = 3,
  seed = 101
)

trend_clusters
```

```{r eval = run_tata_demo}
plot_tata_gene_trend_heatmap(
  tata_trends,
  n_clusters = 3
)
```

## Compact TATA result overview

`plot_tata_results()` bundles the most common summary views into a named list of
plots so you can quickly inspect pseudotime, entropy, branch probabilities, and
representative-cell fate profiles.

```{r eval = run_tata_demo}
tata_overview <- plot_tata_results(
  tata,
  dimred = "UMAP",
  n_probability_cells = 4
)

names(tata_overview)
```

```{r eval = run_tata_demo}
tata_overview$pseudotime
tata_overview$entropy
tata_overview$time_calibration
tata_overview$edge_causality
tata_overview$branch_probabilities
tata_overview$representative_cells
```

## Notes

- `build_knn_graph()`, `compute_topology_weights()`, and
  `compute_time_weights()` are available separately for manual inspection and
  method development.
- `plot_tata_cluster_graph()` places the abstract graph either in a graph-only
  layout or directly on top of a chosen embedding.
- `plot_tata_trajectory_embedding()` is useful when you want a simpler
  trajectory-style overlay rather than the full abstract graph.
- `compute_tata_branch_probabilities()` adds terminal-branch probabilities,
  entropy/plasticity, and hard branch assignments to TATA results.
- `compute_tata_gene_trends()` uses TATA pseudotime and branch probabilities to
  fit branch-specific expression trends for downstream interpretation.
- `plot_tata_results()` provides a compact overview of the key TATA outputs.
