---
title: "Time Aware Trajectory Analysis (TATA) with SCProkaR"
author:
- name: Ning Liu
  affiliation: University of Adelaide
- name: Jinxin Zhao
  affiliation: Monash University
date: "`r Sys.Date()`"
package: SCProkaR
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
    toc_depth: 2
abstract: |
  Time Aware Trajectory Analysis (TATA) is the trajectory component of
  SCProkaR. It abstracts a microbial single-cell dataset into a graph of
  cell states and uses the experimental sampling times of a perturbation
  time course, rather than topology alone, to score, orient and prune the
  transitions between those states. This vignette builds a simulated
  three-drug branching time course, walks through the TATA pipeline both
  step by step and through the `run_tata()` wrapper, quantifies what the
  time labels add over a topology-only ablation, and then uses the
  resulting pseudotime and terminal-branch probabilities for branch
  selection and branch-specific gene-trend analysis.
vignette: >
  %\VignetteIndexEntry{2. Time Aware Trajectory Analysis (TATA)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    crop = NULL,
    fig.width = 6,
    fig.height = 4.5,
    dpi = 96,
    fig.retina = 1
)
```

# Introduction

`r BiocStyle::Biocpkg("SCProkaR")` is a `SingleCellExperiment`-centred
workflow for microbial (prokaryotic) single-cell RNA sequencing data. It
standardises bacterial input objects, computes prokaryote-specific quality
control metrics such as ribosomal RNA fraction, integrates batches, scores
the resulting embeddings with scIB-inspired benchmarking metrics, and runs
pseudobulk differential expression with `r BiocStyle::Biocpkg("edgeR")`.
This vignette covers its trajectory component, Time Aware Trajectory
Analysis (TATA).

## The problem TATA addresses

Bacterial single-cell RNA-seq experiments are usually *perturbation time
courses*: a population is challenged with an antibiotic or a stress and
sampled at a handful of known times. The biological question is directional
— which states give rise to which, when does a tolerant subpopulation
diverge, and which genes change along the way.

Standard trajectory inference is poorly suited to this setting for two
reasons that are specific to microbial data:

* **Bacterial transcriptomes are shallow and low-complexity.** With a few
  hundred to a few thousand detected genes and no rich developmental
  programme, the manifold learnt from expression alone is noisy, and the
  geometric "tips" that topology-only methods use as roots and endpoints
  are often not the biologically earliest or latest states.
* **The root is frequently in the middle.** An untreated founder population
  that diverges into several drug-specific responses sits in the centre of
  the embedding, not at a boundary. Purely topological methods, which infer
  direction from graph structure, have no way to know this and will happily
  run the trajectory backwards.

TATA solves this by *not* asking the data where time starts. The
experimental sampling time is a measured covariate, so TATA uses it
directly: it scores each candidate transition between coarse cell states by
combining a topology term (are these two states actually connected in the
cell--cell graph?) with a time term (do the cells on either side of that
connection respect the observed time order?). Edges that are well connected
but temporally implausible are down-weighted or pruned, and surviving edges
are oriented from earlier to later. The result is a directed abstract graph
that is anchored to real experimental time.

## Relationship to Bioconductor single-cell infrastructure

Every object in SCProkaR is a
`r BiocStyle::Biocpkg("SingleCellExperiment")`, so TATA is a drop-in step in
an ordinary Bioconductor workflow rather than a parallel universe:

* Input is a `SingleCellExperiment` whose `reducedDims` slot holds a PCA (or
  any other) representation — you can produce that with
  `r BiocStyle::Biocpkg("scater")` and
  `r BiocStyle::Biocpkg("scran")`, or with `IntegrateBacData()` from this
  package, which wraps `r BiocStyle::Biocpkg("batchelor")` and Harmony.
* Output is written straight back into the same object: cluster labels,
  `tata_pseudotime`, `tata_pseudotime_scaled`, per-cell terminal-branch
  probabilities, branch entropy and branch assignments all land in
  `colData()`, a `"TATA"` embedding lands in `reducedDims()`, and the
  abstract graph and its edge table land in `metadata()`.
* Because of that, downstream Bioconductor tools work unchanged. Branch
  labels can be passed to the pseudobulk `r BiocStyle::Biocpkg("edgeR")`
  helpers in this package, TATA pseudotime can be handed to
  `r BiocStyle::Biocpkg("edgeR")` spline models, and the object can be
  visualised with any `SingleCellExperiment`-aware plotting package.

TATA itself borrows its abstraction idea from PAGA and its terminal-fate
probability layer from Palantir, but replaces their topology-derived
direction with measured experimental time.

# Installation

```{r install, eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("SCProkaR")
```

# Setting up

```{r load-packages, message = FALSE}
library(SCProkaR)
library(SingleCellExperiment)
library(SummarizedExperiment)
```

TATA builds its cell--cell graph with whichever nearest-neighbour backend is
installed. `r BiocStyle::CRANpkg("FNN")` and
`r BiocStyle::Biocpkg("BiocNeighbors")` are both in `Suggests`; if neither is
available the package falls back to an exact `stats::dist()` search, so the
analysis below runs either way — only the speed changes.

```{r knn-backends}
c(
    FNN = requireNamespace("FNN", quietly = TRUE),
    BiocNeighbors = requireNamespace("BiocNeighbors", quietly = TRUE)
)
```

# A simulated multi-drug branching time course

`simulate_tata_multidrug_sce()` generates a `SingleCellExperiment` in which a
shared founder population is sampled at six times (0--120 h) and diverges
into three drug-specific arms with *different onset times*: drug A commits
early (24 h), drug C in the middle (48 h) and drug B late (96 h).

The simulation is deliberately awkward for topology-only inference: the
earliest cells sit near the centre of the embedding rather than at an
obvious tip, so a method that guesses direction from shape alone has no
reason to start there.

```{r simulate-multidrug}
set.seed(101)

sce <- simulate_tata_multidrug_sce(
    n_cells = 3000,
    n_features = 80,
    n_pcs = 15
)

sce$timepoint_factor <- factor(
    sce$timepoint,
    levels = sort(unique(sce$timepoint))
)

sce
```

The design is a balanced condition-by-time grid:

```{r simulation-design}
with(
    as.data.frame(colData(sce)),
    table(condition, timepoint)
)
```

Plotting the stored UMAP by time, by drug and by the ground-truth state
shows the branching structure and, importantly, that the 0 h cells are
interior rather than peripheral.

```{r plot-simulation-time}
PlotReduction(sce, reduction = "UMAP", colour_by = "timepoint_factor")
```

```{r plot-simulation-condition}
PlotReduction(sce, reduction = "UMAP", colour_by = "condition")
```

```{r plot-simulation-branch}
PlotReduction(sce, reduction = "UMAP", colour_by = "simulated_branch")
```

# The TATA pipeline, step by step

`run_tata()` performs the whole analysis in one call, and most users should
start there. The individual steps are exported as well, because inspecting
them is the fastest way to understand *why* a particular graph came out the
way it did — and to swap in your own clustering or edge scoring during
method development.

## Build the cell--cell graph

TATA operates on a low-dimensional representation. Here we use the stored
PCA and connect each cell to its `k` nearest neighbours.

```{r build-knn}
knn <- build_knn_graph(
    x = sce,
    dimred = "PCA",
    k = 30
)

c(
    n_cells = igraph::vcount(knn$graph),
    n_edges = igraph::ecount(knn$graph),
    k = knn$k
)
```

```{r knn-engine}
knn$knn_engine
```

## Cluster cells into coarse states

The cell graph is partitioned into coarse states, which become the *nodes*
of the abstract TATA graph. TATA is not sensitive to the exact number of
clusters — it needs enough of them that a transition can be localised, but
not so many that each cluster is a single time point.

```{r cluster-states}
sce$tata_cluster_step <- cluster_graph_states(
    cell_graph = knn$graph,
    method = "louvain"
)

table(sce$tata_cluster_step)
```

```{r plot-cluster-states}
PlotReduction(sce, reduction = "UMAP", colour_by = "tata_cluster_step")
```

If `r BiocStyle::CRANpkg("cluster")` is installed we can sanity-check the
partition with a silhouette width on a subsample of cells. This is purely
diagnostic; TATA does not require it.

```{r cluster-silhouette}
if (requireNamespace("cluster", quietly = TRUE)) {
    set.seed(101)
    idx <- sample(seq_len(ncol(sce)), 600)
    pcs <- reducedDim(sce, "PCA")[idx, seq_len(5)]
    sil <- cluster::silhouette(
        as.integer(sce$tata_cluster_step[idx]),
        stats::dist(pcs)
    )
    round(mean(sil[, "sil_width"]), 3)
} else {
    message("Install 'cluster' for the optional silhouette diagnostic.")
}
```

## Score cluster pairs on topology and on time

Each candidate edge of the abstract graph gets two independent scores.

`compute_topology_weights()` asks a purely structural question: do these two
clusters share more cell--cell edges than you would expect by chance given
their sizes? The ratio of observed to expected edges is the topology weight.

```{r topology-weights}
topology_table <- compute_topology_weights(
    adjacency = knn$adjacency,
    clusters = sce$tata_cluster_step
)

head(
    topology_table[
        order(-topology_table$topology_weight),
        c(
            "cluster_a", "cluster_b", "observed_edges",
            "expected_edges", "topology_weight"
        )
    ],
    6
)
```

`compute_time_weights()` asks the temporal question: of the cell--cell edges
that cross between these two clusters, do the individual cells respect the
experimental time order? `flow_score` runs from -1 to 1 and gives the net
direction of time across the boundary; `time_consistency` measures how
reproducible that direction is.

```{r time-weights}
time_table <- compute_time_weights(
    adjacency = knn$adjacency,
    clusters = sce$tata_cluster_step,
    timepoint = sce$timepoint
)

head(
    time_table[
        order(-time_table$time_consistency),
        c(
            "cluster_a", "cluster_b", "n_time_edges",
            "flow_score", "time_weight", "time_consistency"
        )
    ],
    6
)
```

Plotting the two scores against each other makes the point of the method
visible. The horizontal band near `flow_score = 0` contains pairs that are
structurally adjacent but temporally uninformative — cells on both sides
come from the same times, so no transition can be claimed. Points far from
that band are the pairs where time supplies real directional evidence.

```{r plot-edge-score-scatter}
edge_metric_table <- merge(
    topology_table,
    time_table,
    by = c("cluster_a", "cluster_b"),
    all.x = TRUE,
    sort = FALSE
)

ggplot2::ggplot(
    edge_metric_table,
    ggplot2::aes(
        x = topology_weight,
        y = flow_score,
        size = observed_edges,
        colour = time_consistency
    )
) +
    ggplot2::geom_hline(
        yintercept = c(-0.10, 0.10),
        linetype = 2,
        colour = "grey60"
    ) +
    ggplot2::geom_point(alpha = 0.8) +
    ggplot2::scale_colour_viridis_c(option = "plasma") +
    ggplot2::theme_classic() +
    ggplot2::labs(
        title = "Topology and time consistency for cluster pairs",
        x = "topology weight",
        y = "temporal flow score",
        colour = "time consistency",
        size = "observed edges"
    )
```

## Build the abstract graph

`build_tata_graph()` combines the two score tables into a single edge weight,
`topology_weight^alpha * time_weight^beta`, prunes edges below
`prune_threshold`, and orients the survivors using `direction_threshold` on
the flow score. Edges whose flow score is inside the threshold are kept but
left undirected, which is the honest answer for a pair that time cannot
separate.

```{r build-graph-manual}
tata_manual <- build_tata_graph(
    topology_table = topology_table,
    time_table = time_table,
    clusters = sce$tata_cluster_step,
    embedding = reducedDim(sce, "PCA"),
    timepoint = sce$timepoint,
    alpha = 1,
    beta = 1,
    direction_threshold = 0.10,
    prune_threshold = 0.001,
    time_weight_mode = "directional_confidence"
)

head(
    tata_manual$edge_table[
        order(-tata_manual$edge_table$tata_weight),
        c(
            "cluster_a", "cluster_b", "topology_weight", "flow_score",
            "time_consistency", "tata_weight", "direction_label", "kept"
        )
    ],
    8
)
```

The graph is easiest to read when drawn on the same embedding used to view
the cells, so that each node sits on the cells it summarises.

```{r plot-manual-graph-embedded}
plot_tata_cluster_graph(
    list(
        cluster_graph = tata_manual$graph,
        sce = sce,
        parameters = list(cluster_col = "tata_cluster_step")
    ),
    dimred = "UMAP",
    layout_mode = "embedding",
    show_cells = TRUE,
    colour_by = "tata_cluster_step",
    node_colour_by = "cluster",
    main = "Manual TATA graph aligned to UMAP"
)
```

The same graph in a PAGA-style abstract layout, with nodes coloured by
cluster median time, shows the temporal gradient directly.

```{r plot-manual-graph-abstract}
plot_tata_cluster_graph(
    list(
        cluster_graph = tata_manual$graph,
        sce = sce,
        parameters = list(cluster_col = "tata_cluster_step")
    ),
    layout_mode = "graph",
    node_colour_by = "median_time",
    main = "Manual TATA abstract graph"
)
```

## Compute pseudotime

`compute_tata_pseudotime()` picks a root (by default the cluster with the
earliest median experimental time), walks the directed graph with inverse
edge weights, and then spreads cells within each cluster by their local time
rank. Because the root is chosen by measured time rather than by graph
geometry, it lands on the founder population even though that population is
in the middle of the embedding.

```{r manual-pseudotime}
sce$tata_cluster_step <- factor(sce$tata_cluster_step)

pt_manual <- compute_tata_pseudotime(
    cluster_graph = tata_manual$graph,
    sce = sce,
    cluster_col = "tata_cluster_step",
    time_col = "timepoint"
)

pt_manual$root_cluster
summary(pt_manual$cell_pseudotime_scaled)
```

The root cluster's median experimental time confirms the choice:

```{r manual-root-check}
tapply(
    sce$timepoint,
    sce$tata_cluster_step,
    median
)[pt_manual$root_cluster]
```

# The `run_tata()` wrapper

For routine analysis, `run_tata()` does everything above in one call: graph
construction, clustering (with an optional conservative temporal refinement
step), edge scoring, abstraction, pseudotime and terminal-branch
probabilities. Setting `expected_branches = 3` tells the branch layer how
many terminal fates to look for, which matches the known three-arm design.

```{r run-tata}
set.seed(101)

tata <- run_tata(
    sce = sce,
    dimred = "PCA",
    time_col = "timepoint",
    cluster_col = "tata_cluster",
    k = 30,
    cluster_method = "louvain",
    alpha = 1,
    beta = 1,
    direction_threshold = 0.10,
    prune_threshold = 0.001,
    time_weight_mode = "directional_confidence",
    do_pseudotime = TRUE,
    expected_branches = 3
)

names(tata)
```

The updated object carries everything TATA inferred:

```{r inspect-coldata}
sce_tata <- tata$sce
sce_tata$tata_cluster <- factor(sce_tata$tata_cluster)

grep("^tata_", colnames(colData(sce_tata)), value = TRUE)
reducedDimNames(sce_tata)
```

```{r plot-tata-clusters}
PlotReduction(sce_tata, reduction = "UMAP", colour_by = "tata_cluster")
```

```{r plot-tata-pseudotime}
PlotReduction(
    sce_tata,
    reduction = "UMAP",
    colour_by = "tata_pseudotime_scaled"
)
```

Pseudotime should increase outward from the interior founder population
along all three arms simultaneously, rather than sweeping across the
embedding from one side.

```{r plot-tata-graph-embedded}
plot_tata_cluster_graph(
    tata,
    dimred = "UMAP",
    layout_mode = "embedding",
    show_cells = TRUE,
    colour_by = "timepoint",
    node_colour_by = "cluster",
    main = "TATA abstract graph aligned to UMAP"
)
```

```{r plot-tata-graph-abstract}
plot_tata_cluster_graph(
    tata,
    layout_mode = "graph",
    node_colour_by = "median_time",
    main = "TATA abstract graph"
)
```

`plot_tata_trajectory_embedding()` gives a more classical trajectory
overlay, tracing root-to-terminal paths through the cluster centroids. Use
`max_paths` to keep the figure readable when the graph has many terminal
states.

```{r plot-trajectory-all-paths}
plot_tata_trajectory_embedding(
    tata,
    dimred = "UMAP",
    colour_by = "timepoint",
    smooth_paths = TRUE
)
```

```{r plot-trajectory-top-paths}
plot_tata_trajectory_embedding(
    tata,
    dimred = "UMAP",
    colour_by = "timepoint",
    max_paths = 3,
    smooth_paths = TRUE
)
```

# What time adds beyond topology

The claim that time labels help is testable, and the cleanest test is an
ablation: run exactly the same pipeline with `use_time = FALSE`, so graph
construction, clustering and abstraction are unchanged but the edge scores
and directions come from topology alone.

```{r run-tata-topology-only}
set.seed(101)

tata_topology <- run_tata(
    sce = sce,
    dimred = "PCA",
    use_time = FALSE,
    time_col = NULL,
    cluster_col = "tata_cluster_topology",
    k = 30,
    cluster_method = "louvain",
    alpha = 1,
    beta = 1,
    direction_threshold = 0.10,
    prune_threshold = 0.001,
    do_pseudotime = TRUE,
    expected_branches = 3
)
```

## Benchmarking helpers

SCProkaR exports five metrics for scoring a trajectory against known
experimental time. All of them are sign-conventional, so before scoring we
orient pseudotime with `orient_pseudotime_to_time()`; otherwise a trajectory
that is perfectly correct but inferred backwards would score near zero for
the wrong reason. This is especially important for the topology-only run,
which has no way to know which end is the start.

```{r orient-pseudotime}
pt_tata <- orient_pseudotime_to_time(
    pseudotime = tata$cell_pseudotime,
    timepoint = tata$sce$timepoint
)

pt_topology <- orient_pseudotime_to_time(
    pseudotime = tata_topology$cell_pseudotime,
    timepoint = tata_topology$sce$timepoint
)
```

**Local temporal order.** `local_temporal_order_metrics()` walks every edge
of the cell--cell graph and asks whether pseudotime and experimental time
move in the same direction. It is a strictly local measure, so it is not
fooled by a trajectory that is globally monotone but locally scrambled.

```{r benchmark-local-order}
local_tata <- local_temporal_order_metrics(
    pseudotime = pt_tata,
    adjacency = tata$adjacency_matrix,
    timepoint = tata$sce$timepoint
)

local_topology <- local_temporal_order_metrics(
    pseudotime = pt_topology,
    adjacency = tata_topology$adjacency_matrix,
    timepoint = tata_topology$sce$timepoint
)

rbind(TATA = local_tata, topology_only = local_topology)
```

**Adjacent-timepoint AUC.** `adjacent_timepoint_auc()` asks a rank question:
given a cell from time *t* and a cell from the next sampled time *t + 1*,
how often does pseudotime put them in the right order? Because it compares
only adjacent times, it is not inflated by the easy 0 h versus 120 h
contrast. Passing `group` computes it within each drug arm first, which
prevents a between-arm offset from masquerading as temporal accuracy.

```{r benchmark-adjacent-auc}
auc_tata <- adjacent_timepoint_auc(
    pseudotime = pt_tata,
    timepoint = tata$sce$timepoint,
    group = tata$sce$condition
)

auc_topology <- adjacent_timepoint_auc(
    pseudotime = pt_topology,
    timepoint = tata_topology$sce$timepoint,
    group = tata_topology$sce$condition
)

rbind(TATA = auc_tata$summary, topology_only = auc_topology$summary)
```

The per-pair table shows which time intervals are actually hard:

```{r benchmark-auc-pairs}
head(auc_tata$pair_table)
```

**Directed reachability.** `directed_reachability_concordance()` moves up to
the abstract graph. For every pair of clusters separated by at least one
sampling interval, it checks whether the earlier cluster can reach the later
one in the directed graph but not vice versa. A score of 0.5 means the graph
is effectively undirected for those pairs — which is exactly what a
topology-only run produces, since it has no basis for orienting anything.

```{r benchmark-reachability}
reach_tata <- directed_reachability_concordance(
    cluster_graph = tata$cluster_graph,
    clusters = tata$sce$tata_cluster,
    timepoint = tata$sce$timepoint,
    min_time_gap = 24
)

reach_tata$summary
head(reach_tata$pair_table)
```

**Anti-causal edge mass.** `anticausal_edge_mass()` is the sharpest
diagnostic for a real dataset because it needs no pseudotime at all. It sums
the weight of directed edges that point backwards in median cluster time,
beyond a tolerance that absorbs sampling jitter. A large anti-causal mass
means the abstract graph is asserting transitions that the experiment
contradicts.

```{r benchmark-anticausal}
anticausal_tata <- anticausal_edge_mass(
    cluster_graph = tata$cluster_graph,
    clusters = tata$sce$tata_cluster,
    timepoint = tata$sce$timepoint,
    tolerance = 24
)

anticausal_tata$summary
```

```{r benchmark-anticausal-edges}
head(
    anticausal_tata$edge_table[
        order(anticausal_tata$edge_table$time_delta),
        c("from", "to", "weight", "time_delta", "temporal_class")
    ]
)
```

**Branch onset ordering.** `branch_onset_metrics()` is the metric closest to
the biological question. Each arm of the simulation commits at a known hour
(A at 24 h, C at 48 h, B at 96 h), so we can ask whether the inferred
pseudotime recovers that *ordering of commitment events*, not just the
overall time axis. This is where topology-only inference usually fails: it
can produce a plausible-looking tree in which the late-diverging arm appears
to branch first.

```{r benchmark-branch-onset}
true_onset <- c(drug_A = 24, drug_B = 96, drug_C = 48)

onset_tata <- branch_onset_metrics(
    pseudotime = pt_tata,
    branch = tata$sce$simulated_branch,
    branch_activation = tata$sce$branch_activation,
    true_onset = true_onset
)

onset_topology <- branch_onset_metrics(
    pseudotime = pt_topology,
    branch = tata_topology$sce$simulated_branch,
    branch_activation = tata_topology$sce$branch_activation,
    true_onset = true_onset
)

rbind(TATA = onset_tata$summary, topology_only = onset_topology$summary)
```

```{r benchmark-onset-table}
onset_tata$onset_table
```

## Calibration and causality plots

`plot_tata_time_calibration()` is the visual counterpart of the AUC metric:
pseudotime should rise monotonically with experimental time, and it should
do so with a similar slope in every drug arm. Systematic differences between
arms indicate that pseudotime is partly encoding treatment identity rather
than progression.

```{r plot-calibration-tata}
plot_tata_time_calibration(
    tata,
    branch_col = "condition",
    summary = "median_iqr"
)
```

```{r plot-calibration-topology}
plot_tata_time_calibration(
    tata_topology,
    time_col = "timepoint",
    branch_col = "condition",
    summary = "median_iqr"
)
```

`plot_tata_edge_causality()` redraws the topology-versus-time plane for the
final graph, marking which edges were kept. Edges in the upper right are
both structurally strong and temporally forward — those are the transitions
you can defend.

```{r plot-edge-causality}
plot_tata_edge_causality(tata)
```

# Branch probabilities and plasticity

On top of the abstract graph, TATA adds a Palantir-style absorbing-Markov
layer. Terminal clusters are identified from the directed graph, and every
cluster and cell receives a probability of ending up in each terminal state.
`compute_tata_branch_probabilities()` computes this directly; `run_tata()`
calls it for you when `do_branch_probs = TRUE`.

```{r branch-probabilities}
branch_result <- compute_tata_branch_probabilities(
    tata_result = tata,
    expected_branches = 3
)

branch_result$terminal_clusters
```

Three quantities are derived from the probability vector of each cell:

* **entropy** — a normalised Shannon entropy over the terminal
  probabilities. High entropy means the cell has not committed; this is the
  operational definition of *plasticity* here.
* **max probability / commitment** — how strongly the cell favours its most
  likely fate.
* **assignment** — a hard label, useful for subsetting, that should be
  treated as a summary of the probabilities rather than a ground truth.

```{r cluster-branch-table}
prob_cols <- grep(
    "^tata_prob_",
    colnames(tata$cluster_branch_probabilities),
    value = TRUE
)

head(
    tata$cluster_branch_probabilities[
        ,
        c("cluster", prob_cols, "tata_branch_entropy",
          "tata_branch_assignment")
    ]
)
```

Entropy is highest in the shared founder population and near the branch
points, and drops as cells commit:

```{r plot-branch-entropy}
PlotReduction(
    sce_tata,
    reduction = "UMAP",
    colour_by = "tata_branch_entropy"
)
```

```{r plot-branch-assignment}
PlotReduction(
    sce_tata,
    reduction = "UMAP",
    colour_by = "tata_branch_assignment"
)
```

One panel per terminal branch gives the full probability landscape:

```{r plot-branch-probabilities}
plot_tata_branch_probabilities(
    tata,
    dimred = "UMAP",
    ncol = 2
)
```

## Selecting the cells of one lineage

`select_tata_branch_cells()` isolates the cells compatible with a single
fate. The `"soft"` mode thresholds the branch probability and therefore
includes uncommitted progenitors that are *en route* to that fate — usually
what you want when looking for early drivers. The `"hard"` mode uses the
winner-take-all assignment and gives a cleaner but smaller set.

```{r select-branch-cells}
branch_one <- tata$terminal_clusters[1]

soft_selection <- select_tata_branch_cells(
    tata,
    branch = branch_one,
    selection_mode = "soft",
    probability_threshold = 0.60
)

hard_selection <- select_tata_branch_cells(
    tata,
    branch = branch_one,
    selection_mode = "hard"
)

c(
    branch = branch_one,
    soft = sum(soft_selection$selected),
    hard = sum(hard_selection$selected)
)
```

```{r plot-branch-selection}
plot_tata_branch_selection(
    tata,
    branch = branch_one,
    dimred = "UMAP",
    selection_mode = "soft",
    probability_threshold = 0.60
)
```

`plot_tata_terminal_probabilities()` shows the fate profile of individual
cells, defaulting to the most ambiguous ones. Flat bars are undecided cells;
a single tall bar is a committed cell.

```{r plot-terminal-probabilities}
plot_tata_terminal_probabilities(
    tata,
    n_cells = 4
)
```

Because this is a simulation, the hard assignments can be compared against
the known arm of origin. Off-diagonal mass is concentrated in the founder
population, whose cells legitimately have no single fate.

```{r plot-branch-confusion}
plot_tata_branch_confusion(
    tata,
    truth_col = "simulated_branch",
    predicted_col = "tata_branch_assignment",
    normalize = "truth"
)
```

# Branch-specific gene trends

With pseudotime and branch probabilities in hand, expression can be modelled
along each lineage separately. `compute_tata_gene_trends()` fits a smooth
trend of expression against pseudotime for each gene and branch, weighting
each cell by its probability of belonging to that branch, so progenitors
contribute to every branch and committed cells contribute to one.

We pick genes from the simulation's branch-specific and late feature modules
recorded in `rowData()`.

```{r select-trend-genes}
feature_modules <- rowData(sce_tata)$feature_module

trend_genes <- unique(c(
    rownames(sce_tata)[feature_modules == "branch_A"][1:2],
    rownames(sce_tata)[feature_modules == "branch_B"][1:2],
    rownames(sce_tata)[feature_modules == "branch_C"][1:2],
    rownames(sce_tata)[feature_modules == "late"][1:2]
))

trend_genes
```

```{r compute-gene-trends}
tata_trends <- compute_tata_gene_trends(
    tata,
    genes = trend_genes,
    grid_length = 60
)

tata_trends$branches
head(tata_trends$trend_table)
```

A branch-specific gene should rise in one panel line and stay flat in the
others; a shared late gene should rise in all of them together. That
contrast is the readout you are after.

```{r plot-gene-trends}
plot_tata_gene_trends(
    tata_trends,
    ncol = 3
)
```

Trends can also be grouped by shape. `cluster_tata_gene_trends()` z-scores
each gene-branch curve and runs k-means on it, which is a quick way to find
programmes that switch on at the same point in pseudotime. It uses
`stats::kmeans()`, so set a seed for reproducibility.

```{r cluster-gene-trends}
set.seed(101)

trend_clusters <- cluster_tata_gene_trends(
    tata_trends,
    n_clusters = 3
)

trend_clusters
```

```{r plot-gene-trend-heatmap}
set.seed(101)

plot_tata_gene_trend_heatmap(
    tata_trends,
    scale_rows = TRUE,
    n_clusters = 3
)
```

# Compact result overview

`plot_tata_results()` bundles the summary views into a named list of
`ggplot2` objects, which is convenient when you want to skim a new dataset
before deciding what to look at closely.

```{r tata-overview}
tata_overview <- plot_tata_results(
    tata,
    dimred = "UMAP",
    n_probability_cells = 4
)

names(tata_overview)
```

```{r plot-overview-pseudotime}
tata_overview$pseudotime
```

```{r plot-overview-entropy}
tata_overview$entropy
```

```{r plot-overview-calibration}
tata_overview$time_calibration
```

```{r plot-overview-branch-probabilities}
tata_overview$branch_probabilities
```

# Practical guidance

* **Choose `k` for the number of cells, not out of habit.** Very small `k`
  fragments the graph and produces clusters that span a single time point,
  which destroys the time signal that TATA depends on.
* **`direction_threshold` controls how confident you require a direction to
  be.** Raising it leaves more edges undirected, which is preferable to
  asserting a transition the sampling design cannot support.
* **Check `anticausal_edge_mass()` on every real dataset.** It requires no
  ground truth and no pseudotime, and a non-trivial anti-causal fraction is
  a reliable sign that the clustering or the neighbourhood size needs
  revisiting.
* **Treat branch assignments as summaries.** The probabilities and the
  entropy carry the information; the hard label discards it. For
  differential expression, prefer soft selection or use the probability as a
  covariate.
* **Time labels must be comparable across conditions.** TATA assumes the
  sampling times mean the same thing in every arm, which is true of a
  designed time course but not of pooled datasets with different schedules.

# Further reading

* Robinson MD, McCarthy DJ, Smyth GK (2010). edgeR: a Bioconductor package
  for differential expression analysis of digital gene expression data.
  *Bioinformatics* 26, 139--140.
  <https://bioconductor.org/packages/edgeR/>
* Haghverdi L, Lun ATL, Morgan MD, Marioni JC (2018). Batch effects in
  single-cell RNA-sequencing data are corrected by matching mutual nearest
  neighbors. *Nature Biotechnology* 36, 421--427. Implemented in
  `r BiocStyle::Biocpkg("batchelor")`.
* Korsunsky I *et al.* (2019). Fast, sensitive and accurate integration of
  single-cell data with Harmony. *Nature Methods* 16, 1289--1296.
* Luecken MD *et al.* (2022). Benchmarking atlas-level data integration in
  single-cell genomics (scIB). *Nature Methods* 19, 41--50. The integration
  metrics in `BenchmarkIntegration()` follow this benchmark.
* Wolf FA *et al.* (2019). PAGA: graph abstraction reconciles clustering
  with trajectory inference. *Genome Biology* 20, 59. TATA's abstract
  cluster graph follows this design.
* Setty M *et al.* (2019). Characterization of cell fate probabilities in
  single-cell data with Palantir. *Nature Biotechnology* 37, 451--460. The
  source of the absorbing-Markov branch-probability layer.
* McInnes L, Healy J, Melville J (2018). UMAP: Uniform Manifold
  Approximation and Projection for dimension reduction.
  <https://arxiv.org/abs/1802.03426>. Available in R via
  `r BiocStyle::CRANpkg("uwot")`.
* Amezquita RA *et al.* (2020). Orchestrating single-cell analysis with
  Bioconductor. *Nature Methods* 17, 137--145.
  <https://bioconductor.org/books/release/OSCA/>

# Session information

```{r session-info}
sessionInfo()
```
